---
title: "No exit: the conservation reliance of species with too little habitat"
output:
    bookdown::word_document2:
        reference_docx: "../HMEI-GDrive/resources/word-templates/ms_general.docx"
        number_sections: false
    bookdown::pdf_document2:
        toc: no
        latex_engine: xelatex
        fig_caption: yes
        keep_tex: yes
link-citations: yes
csl: conservation-biology.csl
biblio-style: apalike
bibliography: [refs.bib, funky-refs.bib, pckgs.bib]
fontsize: 12pt
geometry: margin=2.5cm
mainfont: Calibri
papersize: a4
documentclass: article
header-includes:
- \usepackage{hyperref}
- \usepackage{fancyhdr}
- \pagestyle{fancy}
- \fancyhf{}
- \fancyfoot[CO,CE]{\thepage}
- \usepackage{setspace}
- \doublespacing
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = "../")
knitr::opts_chunk$set(echo = FALSE,message = FALSE,error = FALSE, warning = FALSE,
                      dpi = 300, fig.path = "figs/main/", fig.align = "center", 
                      fig.width = 16.6/2.54, fig.height = 15/2.54)
```

```{r packages}
library(readr)
library(ggplot2)
library(stringr)
library(dplyr)
library(ggtext)
library(png)
library(grid)
library(RColorBrewer)
library(Hmisc)

# Define function to capitalise first letter
capitalise <- function(string) {
  substr(string, 1, 1) <- toupper(substr(string, 1, 1))
  string
}
```

```{r data}
load("HMEI-GDrive/data/sorted/msdata.Rdata")
```

_______________________________________________________________________________________

Sophia Richter^1^&Dagger;, Rebecca A. Senior^2^*&Dagger;, and David S. Wilcove^1,2^

^1^Department of Ecology and Evolutionary Biology, Princeton University, Princeton, NJ, USA.

^2^Princeton School of Public and International Affairs, Princeton University, Princeton, NJ, USA.

*__Corresponding author:__ rebecca.a.senior@gmail.com (R.A. Senior)

&Dagger;These authors contributed equally to this work

__ORCID iDs:__ orcid.org/0000-0002-8208-736X (R.A. Senior)
  
__Authorship:__ R.A.S. and D.S.W. conceived the study. R.A.S. collated the data. S.R. carried out the analyses and wrote the first draft of the manuscript, with substantial contribution from R.A.S. and D.S.W. 

__Funding:__ High Meadows Environmental Institute and High Meadows Foundation.

__Key words:__ conservation; extinction; IUCN Red List; conservation planning; biodiversity; threatened species

__Article type:__ Perspective

__Number of words:__ 4750

__Number of References:__ 63

__Number of Figures:__ 3; __Number of Tables:__ 0 

# Abstract

<!-- The abstract is 274 words right now --> 

The world is currently experiencing a biodiversity crisis driven by habitat loss and conversion. This loss of habitat has led to many species living in highly fragmented and small habitat areas that increase several risk factors for extinction, including stochastic and edge effects. Small habitat area pushes species to become conservation reliant, with no hope for recovering without habitat restoration. Recent advances in satellite data have allowed us to create habitat maps and quantify Area of Habitat (AOH) for 3,869 terrestrial species. Of these species with AOH data, it was found that 2,488 species have an AOH 2000 km^2^. These species are classified as "No Exit Species" (NES) because they meet the threatened category habitat criteria on the IUCN Red List and are unable to exit 'threatened' status without an increase in available habitat. Most NES are, in order, amphibians, birds, and mammals. The absolute difference between each species' AOH and the threshold for downlisting to the next lowest Red List category was calculated, and Extreme No Exit Species (ENES) were identified as the upper quantile of species within each habitat category that are furthest from the threshold. In spatial analysis, it was found that a few key countries house a high number of both NES and ENES, comprising Colombia, Mexico, Madagascar, and Indonesia. We identify the species and regions most in need of habitat protection, where strategic expansion of PAs under "30 x 30" and habitat restoration through the Bonn Challenge could have a major impact. Additionally, conservation NGOs and organizations with limited resources have the ability to redirect their limited resources towards methods that can make the biggest impact. 

__271/300 words__

# Introduction

Many threatened species can maintain self-sustaining population in the wild only with ongoing species-specific management interventions. They are referred to as "conservation reliant species" [@scott_conservation-reliant_2010; @goble_conservation-reliant_2012; @scott_recovery_2005]. With continual management, such as control of invasive species or prescribed fires where early successional habitat is required, conservation reliant species can be sustained and may even rebound [@bocettiUsingConservationManagement2012]. With cessation of management, however, such species quickly decline. We argue that an additional form of conservation reliance is inherent where species recovery from threatened status is fundamentally constrained by available habitat. 

With less habitat comes increased susceptible to stochasticity and edge effects, and thus a heightened risk of extinction and a limit to recovery potential. Most environmental stressors are localized, such that a population with a large habitat area benefits from the averaging effect of environmental effects that are not acting over the entire range. Populations with a small habitat, in contrast, are at a much greater risk because environmental changes are likely to affect the entire population [@lacyConsideringThreatsViability2000]. Demographic stochasticity also disproportionately impacts species with restricted ranges, through heightened risk of inbreeding and loss of genetic diversity, for example [@pflugerHabitatLossCauses2019; @brookContributionInbreedingExtinction2002]. Compounding such stochastic effects are the stresses associated with the relatively large amount of edge that prevails in small habitats, including higher temperatures and easier access for poachers [@ewers_fragmentation_2013;@pfeifer_creation_2017]. Meanwhile, the isolation of small, fragmented habitat can impede dispersal [@crooks_global_2011;@haddad_habitat_2015], and this is increasingly problematic in a warming world where species are on the move towards the poles and uphill [@mcguire_achieving_2016;@senior_global_2019]. 

Regardless of the other conservation interventions that we might implement to maintain populations of threatened species, unless remaining habitat is robustly protected and total habitat area is increased, habitat-limited species will always face a significant risk of extinction. Area of available habitat is one of five criteria for categorising species' risk of extinction on the International Union for Conservation of Nature (IUCN) Red List of Threatened Species. The Red List is the global standard for assessing the state of biodiversity, with categories of extinction risk ranging from Least Concern and Near Threatened, to the 'threatened' categories of: 'Vulnerable' 'Endangered,' and 'Critically Endangered' [@mace_quantification_2008; @iucn_iucn_2020]. In the year of the UN Biodiversity Conference in Kunming, China [@ConventionBiodiversityProposes], it is imperative that we identify those species whose recovery is fundamentally limited, to set realistic expectations regarding conservation success. To pragmatically allocate limited conservation resources, we must also be able to identify where habitat protection and restoration will have the most impact.

Owing to the way in which species are assessed for the Red List, a simple but pertinent observation emerges: a threatened species cannot be downlisted to a lower threat category on the Red List while the total area of its available habitat qualifies it for a higher threat category [@iucn_iucn_2020]. Under this 'geographic range' criterion, species are classed as Vulnerable, Endangered, or Critically Endangered, if their habitat area falls below specific thresholds [@iucn_guidelines_2019]. Previously, estimates of habitat area were largely confined to 'extent of occurrence', defined as the area of the minimum polygon that can be drawn to encompass all known, projected or inferred occurrences of a species [@iucn_guidelines_2019]. These range polygons are available for many species, but frequently suffer from 'errors of commission' when areas exist within the range that are not ecologically suitable for the species, often because the elevation or habitat type is not appropriate. Recent advances in data availability and resolution [@jung_global_2020], coupled with extensive data on species' habitat preferences [@iucn_iucn_2020; @birdlife_2020], now permit the refinement of species' extent of occurrence to the areas within a species' range that have an appropriate habitat type and elevation [@brooks_measuring_2019]: 'Area of Habitat' (AOH). With AOH maps we have unprecedented capacity to quantify how much habitat threatened species have, and the extent to which habitat area limits species' recovery potential.

Here, we explore the conservation reliance of habitat-limited species, based on habitat area alone, by comparing threatened species' AOH area to the thresholds used by the Red List. The habitat area of every species will fall somewhere between the threshold for 'uplisting' the species to a higher threat category, and the threshold for 'downlisting' the species to a lower threat category. In the first case, the proximity of species' habitat area to the uplisting threshold identifies species on the brink, whose remaining habitat must be prioritised for protection if we are to safeguard them from further decline. Conversely, the proximity of species' habitat area to the downlisting threshold identifies species that could be 'easy wins', where a relatively small increase in habitat could permit that species to improve and recover. We use the term 'No Exit Species' to describe all species with habitat area below the threshold for categorization as 'Vulnerable' (2000 km^2^), because these species can never exit threatened status without concerted efforts to protected and restore habitat. 

<!-- Biodiversity is the foundation of ecosystem health, underpinning an array of ecosystem services [@titeux_ipbes_2019] and conferring resilience to anthropogenic change through taxonomic, functional and  genetic diversity [@oliver_biodiversity_2015]. However, the world is currently experiencing alarming rates of biodiversity loss [@wwf_living_2020; @rosenberg_decline_2019]. Across all anthropic changes on Earth, habitat conversion is the leading cause of species extinctions  [@maxwell_biodiversity:_2016;@goncalves-souzaHabitatLossExtinction2020]. Protecting habitat must be at the heart of post-2020 conservation policy [@cbd_update_2020;@visconti_protected_2019;@geldmann_essential_2021]. -->

# Methods

## Area of Habitat

We created AOH maps for X species, following @brooks_measuring_2019. In brief, AOH maps were produced for all threatened, terrestrial species in taxonomic groups that have range maps and have been comprehensively assessed by the IUCN Red List to at least family level (all amphibians, birds, and mammals, plus select families in the classes Magnoliopsida and Reptilia). Range maps were downloaded from the websites of the Red List [@iucn_iucn_2020] and BirdLife International [@birdlife_2020], and refined to AOH by retaining only grid cells that fall within the range polygons, and meet the species' habitat and elevation preferences as defined by the Red List [@iucn_iucn_2020]. All spatial analyses were conducted in Google Earth Engine [@gorelick_google_2017].

Following Red List assessment protocol [@iucn_guidelines_2019], we excluded habitats that are defined by the Red List as 'marginal' in their suitability, or 'passage' in their seasonality of use. For migratory species we considered only the smaller of either the breeding or non-breeding range, in keeping with the Red List protocol [@iucn_guidelines_2019]. We do, however, include habitat types listed with 'unknown' suitability or seasonality, because this likely affects many threatened species that are too rare or poorly known for more refined classifications to be possible.

To identify cells with suitable habitat, we used the 1 km-resolution IUCN habitat types map for the year 2015 [@jung_global_2020]. Level 1 habitat categories refer to the broad categories of Forest, Savanna, Shrubland, Grassland, Wetlands (inland), Rocky Areas, Deserts and Artificial habitats, each of which contains finer level 2 categories, e.g., 'Forest - Boreal.' Where a species' habitat preference is given only to level 1 by IUCN, we conservatively assumed that all nested level 2 categories are suitable. To identify cells with suitable elevation, we used the EarthEnv digital elevation model [@robinson_earthenv-dem90:_2014]. All maps were resampled to 2 km, following @iucn_guidelines_2019, and projected to a Mollweide projection to ensure accuracy of area calculations.

To quantify the extent to which AOH is a limiting factor in decreasing a species' risk of extinction, we compared AOH to the fixed thresholds under criterion B used by the Red List to assign species to the Red List categories. The Red List uses five criteria for this classification, and species are assigned to the highest risk category for which they qualify under any of the criteria [@iucn_guidelines_2019]. 

For the three 'threatened' Red List categories of Vulnerable, Endangered and Critically Endangered, the AOH thresholds are, respectively: < 2000 km2; < 500 km2; and < 10 km2. We categorized species as likely to be conservation reliant if their AOH falls below any of the three thresholds, as all three imply a significant risk of extinction based solely on habitat amount; the higher the Red List category, the greater that risk. For our analysis, we further required that species be at least 1 km2 from that threshold, based on the assumption that 1 km2 is within the expected error of the spatial data. Species that are less than 1 km2 from the threshold were considered to have met the threshold. We note that when the Red List uses these thresholds for classification, it additionally requires that two out of three subcriteria are met. [describe briefly the subcriteria] However, several additional metrics would be needed to assess these subcriteria (habitat fragmentation, and temporal trends in population size or habitat quantity or quality) [@santini_applying_2019]. Such classification is beyond the scope of this study, as we are interested not in automating Red List classification, but rather in the extent to which habitat quantity limits our ability to improve species' prospects. Here, when a species is classed as 'No Exit,' we are referring to situations where no matter what other actions we might take to address threats and improve outcomes for it, its restricted AOH suggests that it will remain threatened to a significant degree unless the habitat-associated extinction risk can be mitigated.

We measured the extent to which a species recovery is limited by its habitat by calculating the difference between its AOH and the nearest threshold for downlisting. This difference is the amount of habitat that a species would need to gain to have any chance of downlisting to a lower category of extinction risk, regardless of what other conservation actions we might take. We summarized both the distribution of these differences from the downlisting threshold, as well as the upper quantile of species within each habitat category that are furthest from the threshold - the most extreme examples of 'No Exit' species. We determined the countries and realms in which each species occurs by overlaying species' extant range maps with maps of global administrative areas [@gadm_2018] and zoogeographic realms [@holt_update_2013].  

## Statistical analyses

All statistical analyses were conducted in R version 4.0.3 [@r_core_team_2020]. The degree to which species are constrained from future downlisting by current AOH was captured by the relative distance between a given species' AOH and the closest downlisting threshold. This was calculated as the absolute difference between a species' AOH and the threshold for downlisting to the next lowest Red List category, divided by the maximum possible difference (the difference between thresholds, for example the maximum possible difference for an Endangered species is 2000 – 500 = 1500 km2). Thus, proximity to threshold is a standardized measure that accounts for the fact that thresholds between categories do not increase linearly. As an example, an Endangered species with 1250 km2 AOH has a larger absolute difference to the downlisting threshold for Vulnerable (2000 - 1250 = 750 km2) than does a Critically Endangered species with 5 km2 AOH that requires only 5 km2 more habitat to be downlisted to Endangered (10 - 5 = 5 km2); however, both have a relative proximity to their nearest thresholds of 0.5 (difference / maximum possible: 750 / 1500 = 0.5 and 5 / 10 = 0.5). We used a generalized linear model (GLM) with a binomial distribution, using the function `glm` in the `lme4` R  package [@bates_fitting_2015], to test whether proximity to threshold varied by taxonomic class, biogeography and absolute AOH. Model inference was based on Likelihood Ratio Tests, dropping each variable in turn and comparing the reduced model to the full model [@zuur_mixed_2009].

# Results

Of the 3,869 species with AOH data, 2,488 species have AOH below the 2000 km2 threshold to be considered as No Exit Species (Fig. 1). Of the 2,488 No Exit Species, most were amphibians (1389), followed by birds (527), mammals (461), plants in the class Magnoliopsida (59), and, finally, reptiles (52). By Red List threshold, 335 species fall below the 10 km2 threshold for Critically Endangered, 720 between 10 km2 and the 500 km2 threshold for Endangered, and 1433 between 500 km2 and the 2000 km2 AOH threshold for Vulnerable. We also identified Extreme No Exit Species, as the upper quartile of each threshold group (top 25%) that had AOH values furthest from the downlisting threshold of 2000 km2. Of the 623 species in the upper quartile, 358 were amphibians, followed by birds (124), mammals (113), assessed reptiles (15), and assessed plants in the class Magnoliopsida (13).

```{r fig-1, fig.cap = "The total number of threatened species classified as 'No Exit Species' (NES), within each taxonomic class and AOH range."}
nes <- nes_df %>% 
  mutate(Category = str_replace(Category, "Critically Endangered", 
                                paste("AOH < 10 km","\U00B2",sep= ""))) %>% 
  mutate(Category = str_replace(Category, "Endangered", 
                                paste("10 < AOH < 500 km","\U00B2",sep= ""))) %>% 
  mutate(Category = str_replace(Category, "Vulnerable", 
                                paste("500 < AOH < 2000 km","\U00B2",sep= "")))

# relevel Categories
nes$Category <- 
  factor(nes$Category, 
         levels = c(paste("AOH < 10 km","\U00B2",sep= ""), 
                    paste("10 < AOH < 500 km","\U00B2",sep= ""), 
                    paste("500 < AOH < 2000 km","\U00B2",sep= "")))
nes$className <- 
  factor(nes$className, 
         levels = unique(nes$className),
         labels = capitalize(unique(nes$className)))

# Code to add pics to labels
classes <- levels(nes$className)
labels <- paste(
  "<img src=", 
  "'",
  file.path("HMEI-GDrive/resources/phylopic", paste(tolower(classes), 'png', sep = ".")),
  "' ",
  "width='10' />", 
  classes,
  sep = ""
)
names(labels) <- classes

max_val <- 834

ggplot(nes, aes(x=className, fill=className)) + 
  geom_bar(alpha = 0.7) + 
  facet_wrap(~Category, labeller = label_wrap_gen(width = 25, multi_line = TRUE)) +
  theme_bw() + theme(panel.grid = element_blank()) +
  labs(fill="Class") +
  xlab(" ") + 
  ylab("Count") + 
  theme(legend.position = "none") + 
  scale_fill_brewer(palette="Set1") +
  scale_y_continuous(breaks = seq(0, max_val, by = 100), limits=c(0, max_val)) +
  theme(axis.text.x = element_markdown(color = "black",
                                       angle = 45, 
                                       hjust = 1, vjust = 1,
                                       margin = unit(c(-5, 0, 0, 0), "pt"))) +
  scale_x_discrete(labels = labels)
```

Across class and Red List thresholds, there is a right skew towards a large difference between species' AOH and the next downlisting threshold, particularly for those species with AOH between 10 km^2 and 500 km^2 (Fig. \@ref(fig:fig-2)). This suggests that, within each range of AOH values (bounded by the different the Red List thresholds), the majority of species have AOH that is as far from the next threshold for downlisting as possible. In other words, most species are closer to the threshold for uplisting than for downlisting. We found that proximity to a downlisting threshold did not differ between taxonomic classes (p-value = 0.988, 4 DF) and biogeographic realms (p-value = 0.999, 6 DF), but increased with increasing absolute AOH (p-value < 2.2e-16, 1 DF). Thus, species are relatively farther from the threshold for downlisting when absolute habitat quantity is lower. In other words, the less total AOH a species has, the lower the probability that it is close to the next downlisting threshold.

```{r fig-2, fig.cap = "Distribution of the difference between species' absolute AOH and the nearest threshold for downlisting. Curves are derived from a smoothed kernel density estimate of species-level values, and shaded by taxonomic class."}

ggplot(nes, aes(x=Difference, fill=className)) +
  geom_density(alpha = 0.3, colour = "transparent") +
  facet_wrap(~ Category, scales = "free", 
             labeller = label_wrap_gen(width = 15, multi_line = TRUE)) + 
  theme_bw() + theme(panel.grid = element_blank()) +
  labs(fill="Class") + 
  ylab("Density") +
  xlab(paste("Difference between AOH and threshold (km","\U00B2",")",sep= "")) + 
  theme(axis.text.x = element_text(angle = 45)) + 
  scale_fill_brewer(palette="Set1") 
```

The three realms with the most No Exit Species are Neotropical (489 species), Oriental (439 species), and Panamanian (431 species). These were also the three realms with the most Extreme No Exit Species, although there are more Extreme No Exit Species in Panamanian than Oriental. By mapping total No Exit Species and Extreme No Exit Species by country (Fig. 3), we find that a few countries harbor a disproportionate share of the world's most habitat-limited species. Globally, the three countries with the most No Exit Species are: Colombia, Mexico, and Madagascar. With the addition of Indonesia, these countries also contain the highest number of Extreme No Exit Species - the species with AOH that is furthest from the threshold for downlisting. No Exit Species amphibians are largely found in Mexico and Colombia, while No Exit Species birds are in the northern half of South America and Indonesia, and mammals in Madagascar and Indonesia.

(ref:cap-3) The total number of threatened species within each country classified as (a) 'No Exit Species' (NES) and (b) 'Extreme No Exit Species' (ENES), which are those species in the upper quartile of differences between AOH and the nearest threshold for downlisting. Countries are shaded from low count (yellow) to high (red).

```{r fig-3, fig.cap = "(ref:cap-3)"}
map <- 
  readPNG("HMEI-GDrive/figs/Final Maps/combi_country_grid.png") %>% 
  rasterGrob(interpolate = TRUE)
ggplot() +
  theme_void() +
  scale_x_continuous(expand = c(0,0)) +
  scale_y_continuous(expand = c(0,0)) +
  annotation_custom(map,
                    xmin = -Inf, xmax = Inf,
                    ymin = -Inf, ymax = Inf)
```

# Discussion

Recent advances in quantifying species' habitat area, facilitated by the availability of global habitat maps and data on species' habitat preferences, allowed us to quantify for the first time the number of threatened amphibian, bird and mammal species whose recovery on the Red List is largely limited by habitat area. We identified 2,488 species with AOH below the thresholds used by the Red List to categorize species as threatened (i.e. Vulnerable, Endangered, Critically Endangered), on the basis of their total habitat area (criterion B) [@@iucn_guidelines_2019]. We refer to these as “No Exit Species” because, according to the Red List thresholds, these species have no capacity to exit 'threatened' status without an increase in the amount of habitat available to them. Because we propose an expansion of conservation reliant species to encompass species with too little habitat to be self-sufficient, these identified No Exit Species are also conservation reliant species.

These 2,488 No Exit Species account for over half of the total 3,869 species for which we had AOH data. Most are amphibians, followed by birds and mammals. Many No Exit Species are recognized as Evolutionarily Distinct and Globally Endangered (EDGE) species, where 40 amphibian, 27 bird, 30 mammal, and 4 reptile No Exit Species are in the top 100 EDGE species for their taxa, including species like Archey's frog (_Leiopelma archeyi_), Barrio's frog (_Insuetophrynus acarpicus_), and kagu (_Rhynochetos jubatus_) [@zoologicalsocietyoflondonEDGE]. These species are highly threatened and have higher genetic distinctiveness within their taxon, where the loss of these species would have a disproportionate impact on phylogenetic diversity. Additionally, there are a number of No Exit Species that are charismatic and serve as useful umbrella species, including the Iberian Lynx (_Lynx pardinus_), the Tapanuli Orangutan (_Pongo tapanuliensis_), and the Indian Rhinoceros (_Rhinoceros unicornis_).

For most No Exit Species, the absolute amount of habitat remaining for them puts them at high risk of crossing the threshold into a higher threat category on the Red List (uplisting), if they were to lose additional habitat (Fig. \@ref(fig:fig-2)). Thus, most No Exit Species are in a precarious state: they are closer to the threshold for uplisting than for downlisting. Given the conversion of natural ecosystems predicted for the remainder of this century [@jenkinsProspectsBiodiversity2003], the future for No Exit Species seems increasingly tenuous. The adoption of the “30 x 30” goal to protect 30% of the world's land and coastal seas by 2030 to protect 30% of the world's land and coastal seas by 2030 [@30by30_2021] by countries such as the UK and USA  [@bidenExecutiveOrder140082021; @50CountriesAnnounce] could prevent many of them from being upgraded to a higher threat category, but, ultimately, it will require more than habitat preservation to render them safe from extinction and not heavily dependent upon active management -- they will require habitat restoration.  The countries identified as having a disproportionate share of No Exit Species, including Colombia, Peru, Mexico, and Madagascar, represent priority areas to consider when assessing where and for what species habitat protection is required. At a regional and national scale, priority sites for protection have been identified in the form of Key Biodiversity Areas, designated according to the populations of threatened species that they support [@iucn_global_2016]. Given these criteria for designation, it is likely that KBAs encompass a high number of No Exit Species. It would be prudent to quantify the capacity for KBAs to protect No Exit Species, to support and prioritize which KBAs receive strict protection. In addition, we recognize the important role of 'Other Effective Area-based Conservation Measures' (OECMs) [@unitAichiBiodiversityTargets2020], such as private and community-managed reserves, which can act to expand coverage substantially, while supporting local livelihoods [@oldekopGlobalAssessmentSocial2016].

The total amount of habitat needed to downlist every species per category is 2,115 km2 for species with < 10 km2 habitat, 500,917 km2 for species with between 10 and 500 km2 habitat, and 6,62,712 km2 for species with between 500 and 2,000 km2 habitat. Undoubtedly there will be overlap between habitat requirements of different species, such that restoring habitat in one area would benefit multiple species. However, given the global distribution of No Exit Species (Fig. \@ref(fig:fig-3)), there remains a lot of ground to cover (restore, actually) to ensure that habitat area is no longer a barrier to downlisting. Initially, we suggest that the focus be on restoring habitat for the highest risk species - those with total habitat area of < 10 km2 - as well as the 'easy wins,' which are the Critically Endangered and Endangered species with only a relatively small amount of additional habitat required to enable downlisting. Countries with a large number of No Exit Species requiring only a small increase in habitat area to qualify for downlisting could achieve relatively large gains in habitat for many species simultaneously where their habitat overlaps, with relatively little conservation investment. These countries include Colombia, with over 60 species close to the threshold for downlisting, as well as Mexico, Madagascar, and Indonesia (40-50 species), and China, Peru, India, and Tanzania (~ 30 species). Notably, these are many of the same countries that also have the most Extreme No Exit Species (Fig. \@ref(fig:fig-3)) - that is, the species that are most in need of habitat protection to prevent uplisting. Thus, habitat restoration could be coupled with habitat protection in a few key countries, with disproportionate benefit for the conservation of global terrestrial biodiversity. However, it is important to note that different habitats have different restoration needs and there is variability in difficulty of restoring habitats [@haleEvaluatingWhereHow2019]. For example, creating old-growth forests for spotted owls may take hundreds of years, whereas the habitat of an early successional species might be restorable in a few years. Additionally, habitat restoration hinges on how badly the habitat has been damaged, where more damaged and industrialized habitats would be more difficult to restore. As ecosystem restoration is a relatively young science, there is also a need for more research on how to most effectively restore habitats. 

Habitat restoration as a conservation strategy is gaining traction around the world and is supported by various initiatives such as the Bonn Challenge and the UN Decade on Ecosystem Restoration. The Initiative 20x20 aimed to protect and restore land in Latin America and the Caribbean in support of the Bonn Challenge and the New York Declaration on Forests. Colombia pledged to restore 1 million ha of degraded land by 2020, Mexico nearly 8.5 million ha of degraded land by 2020, and Peru 3.2 million hectares of degraded land [@Initiative20x20]. Post-2020, there is opportunity for these countries to continue habitat restoration with a concentrated effort on areas where No Exit Species would benefit most. Of the other countries identified as priority regions for No Exit Species, Madagascar has pledged 4,000,000 hectares to restoration efforts, Tanzania 5,200,000 hectares, and India 21,000,000 hectares to the Bonn Challenge by 2030 [@BonnChallenge2016]. China and Indonesia have yet to pledge any habitat to the Bonn Challenge. However, the Chinese government is at the forefront of restoration efforts, including the Natural Forest Conservation Program (NFCP), the Conversion of Cropland to Forest Program (CCFP) as part of the NFCP with considerable success in reforesting 28.2 million hectares by 2014 [@zhangChinaConversionCropland2017]. 

<!-- Not sure about this bit yet -->
<!-- Some companies conduct environmentally damaging business but attempt to offset the damage by funding conservation elsewhere. It is also advantageous to identify areas where companies can do the least harm in their business and areas where their offsetting conservation efforts is the most beneficial. Agendas to restore habitat in these countries are important for many NES identified, and further policies to support restoration in these areas will aid in the efforts to cushion NES in these areas from the risk of extinction.  -->

As this study identifies No Exit Species using the AOH available to individual species, we note several limitations to our approach. The first is that we consider only habitat area and not other habitat characteristics that could further restrict the true amount of usable habitat, such as habitat quality and fragmentation [@santini_applying_2019]. Similarly, while population size and trends are influenced by habitat availability, many species' populations are also depressed by additional threats that we do not consider here, including invasive species and overexploitation, for example [@sodhiCausesConsequencesSpecies2009a]. Thus, the list of No Exit Species that we present here represents the species for which habitat availability is likely to be a limiting factor in their recovery, but hardly the only limiting factor. Restoration of habitat would remove this barrier, and thus permit - but not guarantee - recovery of the many species that are threatened by habitat loss. Finally, our identification of No Exit Species was determined by comparing species' habitat area to the thresholds used by the Red List to classify species into categories of extinction risk, but some have argued these thresholds are inevitably not equally meaningful across taxa [@cardosoInvertebrates2012; @martinAreIUCNStandard2009]. Nonetheless, although imperfect, the Red List represents a heavily reviewed, widely accepted, and standardized protocol for measuring progress towards biodiversity targets [@keithScientificFoundationsIUCN2013].

Habitat loss and degradation will continue into the foreseeable future [@jenkinsProspectsBiodiversity2003], presenting ever greater challenges to the species that already have too little habitat. Here, as we enter the post-2020 era of biodiversity conservation, we identify the species and regions most in need of habitat protection. Our analysis of which species are closest and furthest from the threshold for downlisting identifies, respectively, cases where relatively small gains in habitat area carry disproportionate benefit, versus cases for which major restoration efforts or assisted colonization will be needed to facilitate downlisting. In the latter case, identifying the species for which habitat availability is a major barrier to long-term recovery is vital for conservation organizations to effectively manage their limited resources. 

## Data availability 

The datasets used to support the findings of this study are available for download by request from their respective providers. Species assessments can be accessed from the IUCN Red List of Threatened Species website at https://www.iucnredlist.org, and range maps can be requested at https://www.iucnredlist.org/resources/spatial-data-download. Protected area maps can be requested from the World Database on Protected Areas, found at https://www.unep-wcmc.org/resources-and-data/analysis/main/wdpa. 

## Code availability

Custom Python code to calculate species Area of Habitat in Google Earth Engine will be published at a later date, and is available on request from R.A.S.

## Acknowledgements

Funding for this research was provided by the High Meadows Environmental Institute and High Meadows Foundation.

## Author contributions

R.A.S. and D.S.W. conceived the study. R.A.S. collated the data. S.R. carried out the analyses and wrote the first draft of the manuscript, with substantial contribution from R.A.S. and D.S.W. 

## Competing interests

The authors declare no competing interests.

# References
