---
title: "No exit: the conservation reliance of species with too little habitat"
output:
    bookdown::word_document2:
        reference_docx: "../HMEI-GDrive/resources/word-templates/ms_general.docx"
        number_sections: false
    bookdown::pdf_document2:
        toc: no
        latex_engine: xelatex
        fig_caption: yes
        keep_tex: yes
link-citations: yes
csl: conservation-biology.csl
biblio-style: apalike
bibliography: [refs.bib, funky-refs.bib, pckgs.bib]
fontsize: 12pt
geometry: margin=2.5cm
mainfont: Calibri
papersize: a4
documentclass: article
header-includes:
- \usepackage{hyperref}
- \usepackage{fancyhdr}
- \pagestyle{fancy}
- \fancyhf{}
- \fancyfoot[CO,CE]{\thepage}
- \usepackage{setspace}
- \doublespacing
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = "../")
knitr::opts_chunk$set(echo = FALSE,message = FALSE,error = FALSE, warning = FALSE,
                      dpi = 300, fig.path = "figs/main/", fig.align = "center", 
                      fig.width = 16.6/2.54, fig.height = 15/2.54)
```

```{r packages}
library(readr)
library(ggplot2)
library(stringr)
library(dplyr)
library(ggtext)
library(png)
library(grid)
library(RColorBrewer)
library(Hmisc)

# Define function to capitalise first letter
capitalise <- function(string) {
  substr(string, 1, 1) <- toupper(substr(string, 1, 1))
  string
}
```

```{r data}
load("HMEI-GDrive/data/sorted/msdata.Rdata")
```

_______________________________________________________________________________________

Sophia Richter^1^&Dagger;, Rebecca A. Senior^2^*&Dagger;, and David S. Wilcove^1,2^

^1^Department of Ecology and Evolutionary Biology, Princeton University, Princeton, NJ, USA.

^2^Princeton School of Public and International Affairs, Princeton University, Princeton, NJ, USA.

*__Corresponding author:__ rebecca.a.senior@gmail.com (R.A. Senior)

&Dagger;These authors contributed equally to this work

__ORCID iDs:__ orcid.org/0000-0002-8208-736X (R.A. Senior)
  
__Authorship:__ R.A.S. and D.S.W. conceived the study. R.A.S. collated the data. S.R. carried out the analyses and wrote the first draft of the manuscript, with substantial contribution from R.A.S. and D.S.W. 

__Funding:__ High Meadows Environmental Institute and High Meadows Foundation.

__Key words:__ conservation; extinction; IUCN Red List; conservation planning; biodiversity; threatened species

__Article type:__ Perspective

__Number of words:__ 4750

__Number of References:__ 63

__Number of Figures:__ 3; __Number of Tables:__ 0 

# Abstract

<!-- The abstract is 274 words right now --> 

The world is currently experiencing a biodiversity crisis driven by habitat loss and conversion. This loss of habitat has led to many species living in highly fragmented and small habitat areas that increase several risk factors for extinction, including stochastic and edge effects. Small habitat area pushes species to become conservation reliant, with no hope for recovering without habitat restoration. Recent advances in satellite data have allowed us to create habitat maps and quantify Area of Habitat (AOH) for 3,869 terrestrial species. Of these species with AOH data, it was found that 2,488 species have an AOH 2000 km^2^. These species are classified as "No Exit Species" (NES) because they meet the threatened category habitat criteria on the IUCN Red List and are unable to exit ‘threatened’ status without an increase in available habitat. Most NES are, in order, amphibians, birds, and mammals. The absolute difference between each species’ AOH and the threshold for downlisting to the next lowest Red List category was calculated, and Extreme No Exit Species (ENES) were identified as the upper quantile of species within each habitat category that are furthest from the threshold. In spatial analysis, it was found that a few key countries house a high number of both NES and ENES, comprising Colombia, Mexico, Madagascar, and Indonesia. We identify the species and regions most in need of habitat protection, where strategic expansion of PAs under "30 x 30" and habitat restoration through the Bonn Challenge could have a major impact. Additionally, conservation NGOs and organizations with limited resources have the ability to redirect their limited resources towards methods that can make the biggest impact. 

__271/300 words__

# Introduction

Biodiversity is the foundation of ecosystem health, underpinning an array of ecosystem services [@titeux_ipbes_2019] and conferring resilience to anthropogenic change through taxonomic, functional and  genetic diversity [@oliver_biodiversity_2015]. However, the world is currently experiencing alarming rates of biodiversity loss [@wwf_living_2020; @rosenberg_decline_2019]. Across all anthropic changes on Earth, habitat conversion is the leading cause of species extinctions  [@maxwell_biodiversity:_2016;@goncalves-souzaHabitatLossExtinction2020]. The underlying driver of habitat conversion is agricultural expansion [@wwf_living_2020; @williams_proactive_2020; @gibbs_tropical_2010], which is expected to continue to pose a major threat to biodiversity. As demand for food increases, driven by population growth and eating habits, the agricultural sector is expected to grow and develop in tandem. Improvements in efficiency, such as increasing yields per area, decreasing waste, and shifting away from meat-based diets in some cultures [@tilman_global_2011; @tilman_future_2017; @williams_proactive_2020], could help to abate habitat loss for some threatened species, but many will face further conversion of habitat - particularly in the tropics [@gibbs_tropical_2010]. Protecting habitat must be at the heart of post-2020 conservation policy [@cbd_update_2020;@visconti_protected_2019;@geldmann_essential_2021].

Habitat area and fragmentation are directly related to extinction risk. Indeed, the International Union for the Conservation of Nature (IUCN) include habitat area as one of the five listing criteria for the Red List, according to which, assessed species are placed into categories of extinction risk ranging from Least Concern and Near Threatened, to the 'threatened' categories of: 'Vulnerable', 'Endangered', and 'Critically Endangered' [@mace_quantification_2008; @iucn_iucn_2020]. Ultimately, populations confined to smaller areas are more susceptible to stochasticity and edge effects. Most environmental stressors are localized, such that a population with a large area of habitat benefits from the averaging effect of environmental effects that are not over the entire range, while populations with a small habitat are at a much higher risk of extinction because environmental changes are likely to affect the entire population [@lacyConsideringThreatsViability2000]. Species with small habitat area are also more likely to be affected by demographic stochasticity, including a heightened risk of inbreeding and loss of genetic diversity [@pflugerHabitatLossCauses2019; @brookContributionInbreedingExtinction2002]. Compounding stochastic effects are the stresses associated with the relatively large amount of edge in small habitats, such as higher temperatures and easier access for poachers [@ewers_fragmentation_2013;@pfeifer_creation_2017]. The isolation of small, fragmented habitat can also make it difficult for species to disperse elsewhere [@crooks_global_2011;@haddad_habitat_2015], which is increasingly problematic in a warming world where species are on the move towards the poles and uphill [@mcguire_achieving_2016;@senior_global_2019]. 

Many species in the wild can only maintain a self-sustaining population with ongoing species-specific management interventions. They are referred to as "Conservation Reliant Species" [@scott_conservation-reliant_2010; @goble_conservation-reliant_2012; @scott_recovery_2005]. The extreme vulnerability to stochastic and edge effects for species with too little habitat inevitably renders them more dependent on conservation interventions to maintain viable populations. Thus, we can think of insufficient habitat area as imposing a form of conservation reliance. Regardless of the other conservation interventions that we might implement to maintain the status quo - such as invasive species control or regulation of trade - unless remaining habitat is robustly protected and total habitat area eventually increased, we cannot hope to recover these habitat-limited species. Yet, recover them we must, if we are to meet Goal 15 of the Sustainable Development Goals (SDGs): to "...halt and reverse land degradation and halt biodiversity loss" [@TransformingOurWorld2015]. 

As conservation policy is updated in the post-2020 era, underpinned by huge international efforts to document the state of nature [@ipbes_global_2019], now is the time to identify where habitat recovery begets species recovery. Habitat protection falls directly in line with the proposed "30 x 30" goal to protect 30% of the world’s land and coastal seas by 2030 [@30by30_2021], which has recently been adopted by countries such as the UK and USA  [@bidenExecutiveOrder140082021; @50CountriesAnnounce], and is anticipated to be adopted by the UN Biodiversity Conference held in Kunming, China in 2022  [@ConventionBiodiversityProposes]. Simultaneously, the undoing of previous loss and degradation sits within the remit of large-scale restoration initiatives, such as the Bonn Challenge, which aims to restore 350 million hectares of degraded and deforested landscapes by 350 million hectares by 2030 [@BonnChallenge2016], and the UN Decade on Ecosystem Restoration[@un_decade_2021]. 

Not only do we see, empirically, that habitat availability is a major barrier to species' recovery  [@huxelHabitatLossFragmentation1999; @kerrHabitatLossLimits2004], but in practice, a species cannot be downlisted to a lower threat category on the Red List while its 'geographic range' still qualifies it for the higher category [@iucn_iucn_2020]. Under criterion B, species are classed as Vulnerable, Endangered, or Critically Endangered, if the total 'Area of Occupancy' (hereafter: AOO; calculated as the sum of occupied cells within a 2 x 2 km grid) [@iucn_guidelines_2019] falls below the thresholds of 2000, 500 and 10 km^2^, respectively. AOO represents a subset of the more commonly used metric 'Extent of Occurrence' (hereafter: 'EOO'), defined as the area of the minimum polygon that can be drawn to encompass all known, projected or inferred occurrences of a species [@iucn_guidelines_2019]. While EOO is less accurate because it may contain several sites that are ecologically unsuitable for a species, AOO is difficult to measure. However, recent advances in satellite data and resulting habitat maps [@jung_global_2020], coupled with extensive data on species' habitat preferences [@iucn_iucn_2020;@birdlife_2020], now make it possible to refine species' EOO to the areas within a species' range that have an appropriate habitat type and elevation [@brooks_measuring_2019]. The refined map is referred to as 'Area of Habitat' (hereafter: AOH), and as AOH is functionally equivalent to AOO, the same AOO thresholds can be used to estimate species' minimum Red List category [@santini_applying_2019]. Thus, we have unprecedented capacity to quantify how much habitat threatened species have, and the extent to which habitat area limits species' recovery potential. 

By comparing species' AOH to the Red List thresholds, we identify two related and important facets of conservation reliance for habitat-limited species, based on habitat area alone. The first is how close a species is to being uplisted on the Red List to a higher threat category. The second is how far a species is from being downlisted to a lower threat category. We refer to all species with AOH falling below any of the Red List threatened thresholds as 'No Exit Species' (hereafter: NES), because they can never exit their threatened status without concerted efforts to protect and restore habitat. In the case of NES close to the uplisting threshold, it is essential to protect what little habitat remains, which can be achieved through strategic expansion of the Protected Area (PA) network under 30 x 30 [@30by30_2021]. For all NES, there is a need to reverse historic habitat loss, which can only be achieved through controversial translocation of species to new habitats outside of their current range  [@lawlerReframingDebateAssisted2011; @luntUsingAssistedColonisation2013], or through restoration efforts like the Bonn Challenge [@BonnChallenge2016]. It is also imperative that we identify NES in order to pragmatically allocate limited conservation resources. Conservation is not cheap [@weiseFinancialCostsLarge2014;@mccarthy_financial_2012]. Conservation stakeholders must consider whether other conservation interventions are worthwhile as a means to maintain the status quo when gains in habitat area cannot be achieved, or whether such efforts are a drain on limited resources for species with no hope of recovery. Restoration efforts, meanwhile, should ideally be targeted towards the species that would benefit most. 

# Methods

## Area of Habitat

To identify species whose recovery is limited by habitat area (NES), we calculated Area of Habitat (AOH) following @brooks_measuring_2019. In brief, AOH maps were produced for all threatened, terrestrial species in taxonomic groups that have range maps and have been comprehensively assessed by the IUCN Red List to at least family level (all amphibians, birds and mammals, plus select families in the classes Magnoliopsida and Reptilia). Range maps were downloaded from the websites of the Red List [@iucn_iucn_2020] and BirdLife International [@birdlife_2020], and refined to AOH by retaining only grid cells that fall within the range polygons, and meet the species' habitat and elevation preferences as defined by the Red List [@iucn_iucn_2020]. All spatial analyses were conducted in Google Earth Engine[@gorelick_google_2017].

Following Red List assessment protocol [@iucn_guidelines_2019], we exclude habitats that are defined by the Red List as 'marginal' in their suitability, or 'passage' in their seasonality of use. As we are interested in the minimum habitat area needed to sustain a healthy global population, for migratory species we consider only the smaller of either the breeding or non-breeding range, in keeping with the Red List protocol [@iucn_guidelines_2019]. We do, however, retain habitat types listed with 'unknown' suitability or seasonality, because this likely affects many threatened species that are too rare or poorly known for such classification to be possible.

To identify cells with suitable habitat, we used the 1 km-resolution IUCN habitat types map for the year 2015 [@jung_global_2020]. Level 1 habitat categories refer to the broad categories of Forest, Savanna, Shrubland, Grassland, Wetlands (inland), Rocky Areas, Deserts and Artificial habitats, each of which contains finer level 2 categories, e.g., 'Forest - Boreal'. Where a species' habitat preference is given only to level 1, we conservatively assume that all nested level 2 categories are also suitable. To identify cells with suitable elevation, we use the EarthEnv digital elevation model [@robinson_earthenv-dem90:_2014]. All maps were resampled to 2 km, following @iucn_guidelines_2019, projected to a Mollweide projection to ensure accuracy of area calculations. 

To quantify the extent to which AOH is a limiting factor in decreasing a species' risk of extinction, we compare AOH to the fixed thresholds used by the Red List to assign species to the Red List categories. The Red List use five criteria for this classification, and species are assigned to the highest risk category for which they qualify under any of the criteria [@iucn_guidelines_2019]. Here, we are specifically referring to criterion B, under which species are classified based on their geographic range. The area of a species' range can be measured based on Extent of Occurrence (EOO) or Area of Occupancy (AOO). AOH is analagous to AOO and can be used in the same way, using the same thresholds [@brooks_measuring_2019]. 

For the three 'threatened' Red List categories of Vulnerable, Endangered and Critically Endangered, the AOH thresholds are, respectively: < 2000 km^; < 500  km^2^; and, < 10  km^2^. We categorise species as 'No Exit' if their AOH falls below any of the three thresholds. We further require that species are at least 1  km^2^ from that threshold, based on the assumption that 1  km^2^ is a trivial difference that can be easily gained or lost, and is within the expected error of the spatial data. Species that are less than 1  km^2^ from the threshold are considered to have met the threshold. We note that when the Red List use these thresholds for classification, they additionally require that two out of three subcriteria are met. However, several additional metrics would be needed to assess these subcriteria (habitat fragmentation, and temporal trends in population size or habitat quantity or quality) [@santini_applying_2019]. Such classification is beyond the scope of this study, as we are interested not in automating Red List classification, but rather in the extent to which habitat quantity limits our ability to improve species' prospects. Here, when a species is classed as 'No Exit', we are referring to situations where no matter what other actions we might take to address threats and improve outcomes for threatened species, their restricted AOH means that NES will always be limited in the extent to which extinction risk can be mitigated.

In downstream analyses, we first summarize the taxonomy and biogeography of NES. Next, we explore the magnitude of species' reliance on habitat by calculating the difference between species' AOH and the threshold for downlisting. This difference is the amount of habitat that a species would need to gain to have any chance of downlisting to a lower category of extinction risk on the Red List, regardless of what other conservation actions we might take. We summarize both the distribution of these differences from the downlisting threshold, as well as the upper quantile of species within each habitat category that are furthest from the threshold - the 'Extreme No Exit' species (ENES). ENES are the species with the most additional habitat required to safeguard them from extinction, and thus the species most likely to require major and concerted effort to recover. We determined the countries and realms in which each species occurs by overlaying species' extant range maps with maps of global administrative areas [@gadm_2018] and zoogeographic realms [@holt_update_2013].  

## Statistical analyses

All statistical analyses were conducted in R version 4.0.3 [@r_core_team_2020]. The degree to which species are constrained from future downlisting by current AOH was captured by the relative distance of AOH from the downlisting threshold. This 'proximity to threshold' measure was calculated as the absolute difference between a species' AOH and the threshold for downlisting to the next lowest Red List category, divided by the maximum possible difference. Thus, proximity to threshold is a standardised measure that accounts for the fact that thresholds between categories do not increase linearly. As an example, a species with 1250 km^2^ AOH has a larger absolute difference to the downlisting threshold (2000 - 1250 = 750 km^2^) than a species with 5 km^2^ AOH (10 - 5 = 5  km^2^), but both have a relative proximity to threshold of 0.5. We used a generalized linear model (GLM) with a binomial distribution, using the function `glm` in the `lme4` R  package [@bates_fitting_2015], to test whether proximity to threshold varied by taxonomic class, biogeography and absolute AOH. Model inference was based on Likelihood Ratio Tests, dropping each variable in turn and comparing the reduced model to the full model [@zuur_mixed_2009].

# Results

Of the 3,869 species with AOH data, 2,488 species have habitat areas below the 2000  km^2^ threshold to be considered as NES (Fig. \@ref(fig:fig-1)). Of the 2,488 NES, most species were amphibians (1389), followed by birds (527), mammals (461), assessed plants in the class Magnoliopsida (59), and, finally, assessed reptiles (52). By Red List threshold, 335 species fall below the 10  km^2^ threshold for Critically Endangered, 720 between the 10  km^2^ and 500  km^2^ thresholds (Endangered), and 1433 have between 500  km^2^ and 2000  km^2^ AOH (Vulnerable). We also identified ENES, as the upper quartile of each threshold group (top 25%) that had AOH values furthest from the downlisting threshold. Of the 623 species in the upper quartile, 358 were amphibians, followed by birds (124), mammals (113), assessed reptiles (15), and assessed plants in the class Magnoliopsida (13).

```{r fig-1, fig.cap = "The total number of threatened species classified as 'No Exit Species' (NES), within each taxonomic class and AOH range."}
nes <- nes_df %>% 
  mutate(Category = str_replace(Category, "Critically Endangered", 
                                paste("AOH < 10 km","\U00B2",sep= ""))) %>% 
  mutate(Category = str_replace(Category, "Endangered", 
                                paste("10 < AOH < 500 km","\U00B2",sep= ""))) %>% 
  mutate(Category = str_replace(Category, "Vulnerable", 
                                paste("500 < AOH < 2000 km","\U00B2",sep= "")))

# relevel Categories
nes$Category <- 
  factor(nes$Category, 
         levels = c(paste("AOH < 10 km","\U00B2",sep= ""), 
                    paste("10 < AOH < 500 km","\U00B2",sep= ""), 
                    paste("500 < AOH < 2000 km","\U00B2",sep= "")))
nes$className <- 
  factor(nes$className, 
         levels = unique(nes$className),
         labels = capitalize(unique(nes$className)))

# Code to add pics to labels
classes <- levels(nes$className)
labels <- paste(
  "<img src=", 
  "'",
  file.path("HMEI-GDrive/resources/phylopic", paste(tolower(classes), 'png', sep = ".")),
  "' ",
  "width='10' />", 
  classes,
  sep = ""
)
names(labels) <- classes

max_val <- 834

ggplot(nes, aes(x=className, fill=className)) + 
  geom_bar() + 
  facet_wrap(~Category, labeller = label_wrap_gen(width = 25, multi_line = TRUE)) +
  theme_bw() + theme(panel.grid = element_blank()) +
  labs(fill="Class") +
  xlab(" ") + 
  ylab("Count") + 
  theme(legend.position = "none") + 
  scale_fill_brewer(palette="Set1") +
  scale_y_continuous(breaks = seq(0, max_val, by = 100), limits=c(0, max_val)) +
  theme(axis.text.x = element_markdown(color = "black",
                                       angle = 45, 
                                       hjust = 1, vjust = 1,
                                       margin = unit(c(-5, 0, 0, 0), "pt"))) +
  scale_x_discrete(labels = labels)
```

<!-- Added stats for p-value and degrees of freedom when each variable was dropped for GLM. Residual deviance for all three was 713.16 -->

Across class and Red List thresholds, there is a right skew towards a large difference between species' AOH and the downlisting threshold, particularly for those species with AOH between 10 km^2 and 500 km^2 (Fig. \@ref(fig:fig-2)). This suggests that, within each range of AOH values (bounded by the different the Red List thresholds), the majority of species have AOH that is as far from the threshold for downlisting as possible. In other words, most species are closer to the threshold for uplisting than for downlisting. We found that proximity to downlisting threshold did not differ between taxonomic classes (p-value = 0.988, 4 DF) and biogeographic realms (p-value = 0.999, 6 DF), but increased with increasing absolute AOH (p-value < 2.2e-16, 1 DF). Thus, species are relatively further from the threshold for downlisting when absolute habitat quantity is lower.

```{r fig-2, fig.cap = "Distribution of the difference between species' absolute AOH and the nearest threshold for downlisting. Curves are derived from a smoothed kernel density estimate of species-level values, and shaded by taxonomic class."}

ggplot(nes, aes(x=Difference, fill=className)) +
  geom_density(alpha = 0.3, colour = "transparent") +
  facet_wrap(~ Category, scales = "free", 
             labeller = label_wrap_gen(width = 15, multi_line = TRUE)) + 
  theme_bw() + theme(panel.grid = element_blank()) +
  labs(fill="Class") + 
  ylab("Density") +
  xlab(paste("Difference between AOH and threshold (km","\U00B2",")",sep= "")) + 
  theme(axis.text.x = element_text(angle = 45)) + 
  scale_fill_brewer(palette="Set1") 
```

The three realms with the most NES are Neotropical (489 species), Oriental (439 species), and Panamanian (431 species). These were also the three realms with the most ENES, although there are more ENES in Panamanian than Oriental. Mapping total NES and ENES by country (Fig. 3), we see that certain countries - such as Colombia, Peru, Mexico and Madagascar - harbor a striking proportion of the world's most habitat-limited species.

(ref:cap-3) The total number of threatened species within each country classified as (a) 'No Exit Species' (NES) and (b) 'Extreme No Exit Species' (ENES), which are those species in the upper quartile of differences between AOH and the nearest threshold for downlisting. Countries are shaded from low count (yellow) to high (red).

```{r fig-3, fig.cap = "(ref:cap-3)"}
map <- 
  readPNG("HMEI-GDrive/figs/Final Maps/combi_country_grid.png") %>% 
  rasterGrob(interpolate = TRUE)
ggplot() +
  theme_void() +
  scale_x_continuous(expand = c(0,0)) +
  scale_y_continuous(expand = c(0,0)) +
  annotation_custom(map,
                    xmin = -Inf, xmax = Inf,
                    ymin = -Inf, ymax = Inf)
```

# Discussion

Recent advances in quantifying species' habitat area, facilitated by the availability of global habitat maps and data on species' habitat preferences, allowed us to quantify for the first time the number of threatened amphibian, bird and mammal species whose recovery on the Red List is limited by habitat area. In total, we identified 2,488 species with AOH below the thresholds used by the Red List to categorize species as threatened (i.e. Vulnerable, Endangered, Critically Endangered), on the basis of their total habitat area (criterion B) [@@iucn_guidelines_2019]. We refer to these as "No Exit Species" (NES) because, according to the Red List thresholds, these species have no capacity to exit 'threatened' status without an increase in the amount of their available habitat. Even for those species that have stable population trends or occupy protected areas, a single event can greatly affect the population of NES and lead to uplisting. Without habitat gains, NES will continue to be limited in their ability to recover. The 2,488 species that are NES account for over half of the total 3,869 species for which we had AOH data. From the list of NES, most were amphibians, followed by birds and mammals. Many NES are Evolutionarily Distinct and Globally Endangered (EDGE) species, including archey’s frog (_Leiopelma archeyi_), the kroombit tinker frog (_Taudactylus pleione_), the eungella day frog (_Taudactylus eungellensis_), and barrio’s frog (_Insuetophrynus acarpicus_), and the kagu (_Rhynochetos jubatus_) and kakapo (_Strigops habroptila_), each of which are top 10 amphibians and birds [@zoologicalsocietyoflondonEDGE]. Among the NES are also several WWF priority mammal species such as the iberian lynx (_Lynx pardinus_), the tapanuli orangutan (_Pongo tapanuliensis_), and the indian rhinoceros (_Rhinoceros unicornis_) [@PriorityEndangeredSpecies2020]. 

For most NES, the absolute amount of habitat available puts them at high risk of crossing the threshold into a higher threat category on the Red List (uplisting), if they were to lose additional habitat (Fig. \@ref(fig:fig-2)). NES represent species on the brink: most are closer to the threshold for uplisting than for downlisting. Given the continued conversion of habitat that are predicted for this Century [@jenkinsProspectsBiodiversity2003], and the number of species threatened by habitat loss [@hoekstraConfrontingBiomeCrisis2005], the future of global biodiversity depends upon protection of what little habitat remains for NES. The "30 x 30" goal can aid in this international effort for habitat protection [@30by30_2021], with strategic expansion of the current PA network into habitat with a high number of NES. Globally, the three countries with the most NES are: Colombia, Mexico, and Madagascar. With the addition of Indonesia, these countries also contain the highest number of ENES - the species with AOH that is furthest from the threshold for downlisting. NES amphibians are largely found in Mexico and Colombia, while NES birds are in the northern half of South America and Indonesia, and mammals in Madagascar and Indonesia. These countries represent priority areas to consider when assessing where and for what species habitat protection is required. At a regional and national scale, priority sites for protection have been identified in the form of Key Biodiversity Areas, designated according to the populations of threatened species that they support [@iucn_global_2016]. Given these criteria for designation, it is likely that KBAs encompass a high number of NES. It would be prudent to quantify the capacity for KBAs to protect NES, to support and prioritize which KBAs receive strict protection. In addition, we recognise the important role of 'Other Effective Area-based Conservation Measures' (OECMs) [@unitAichiBiodiversityTargets2020], such as private and community-managed reserves, which can act to expand coverage substantially, while supporting local livelihoods [@oldekopGlobalAssessmentSocial2016].

Across all species, the average area of additional habitat needed to enable downlisting is 468 km^2. This number represents the average difference between habitat area of each NES and the nearest threshold for downlisting. Summing across all species, the total amount of additional habitat area needed to enable downlisting of all NES is 1,165,744 km^2^, comprising 2,115 km^2^ for species with < 10 km^2^ habitat, 500,917 km^2^ for species with between 10 and 500 km^2^ habitat, and the remaining 6,62,710 km^2^ for species with between 500 and 2,000 km^2^ habitat. Undoubtedly there will be overlap between habitat requirements of different species, such that restoring habitat in one area would benefit multiple species. However, given the global distribution of NES (Fig. \@ref(fig:fig-3)), there remains a lot of ground to cover to ensure that habitat area is no longer a barrier to downlisting. Initially, we suggest that the focus be on restoring habitat for the highest risk species - those with total habitat area of < 10 km^2^ - as well as the 'easy wins', which are the species with only a relatively small amount of additional habitat required to enable downlisting. Countries with a large number of NES that require only a small increase in habitat area to qualify for downlisting could achieve relatively large gains in habitat for many species simultaneously, with relatively little conservation investment. These countries include Colombia, with over 60 species close to the threshold for downlisting, as well as Mexico, Madagascar, and Indonesia (40-50 species), and China, Peru, India, and Tanzania (~ 30 species). Notably, these are many of the same countries that also have the most ENES (Fig. \@ref(fig:fig-3)) - that is, the species that are most in need of habitat protection to prevent uplisting. Thus, habitat restoration could be coupled with habitat protection in a few key countries, with disproportionate benefit for the conservation of global terrestrial biodiversity.

Habitat restoration as a conservation strategy is gaining traction around the world, and is supported by various initiatives like the Bonn Challenge and the UN Decade on Ecosystem Restoration. The Initiative 20x20 aimed to protect and restore land in Latin America and the Caribbean in support of the Bonn Challenge and the New York Declaration on Forests. Colombia pledged to restore 1 million ha of degraded land by 2020, Mexico nearly 8.5 million ha of degraded land by 2020, and Peru 3.2 million hectares of degraded land [@Initiative20x20]. Post-2020, there is opportunity for these countries to continue habitat restoration with a concentrated effort on areas where NES would benefit most. Of the other countries identified as priority regions for NES, Madagascar has pledged 4,000,000 hectares to restoration efforts, Tanzania 5,200,000 hectares, and India 21,000,000 hectares to the Bonn Challenge by 2030 [@BonnChallenge2016]. China and Indonesia have yet to pledge any habitat to the Bonn Challenge. Indonesia is currently implementing the Peatland Restoration Agency (BRG), which was extended to 2024 and aims to restore over 2.6 million hectares of forest and peatlands destroyed by fires in 2015 [@UNOPSRestoringIndonesianPeatlands2020]. However, the general effects of deforestation and a lack of habitat restoration outside of areas affected by the fires in 2015 create a serious need for further policies and organizations to restore land in Indonesia. Although China has not pledged to the Bonn Challenge, the government is at the forefront of restoration efforts, including the the Natural Forest Conservation Program (NFCP), the Conversion of Cropland to Forest Program (CCFP) as part of the NFCP with considerable success in reforesting 28.2 million hectares by 2014 [@zhangChinaConversionCropland2017]. 

<!-- Not sure about this bit yet -->
<!-- Some companies conduct environmentally damaging business but attempt to offset the damage by funding conservation elsewhere. It is also advantageous to identify areas where companies can do the least harm in their business and areas where their offsetting conservation efforts is the most beneficial. Agendas to restore habitat in these countries are important for many NES identified, and further policies to support restoration in these areas will aid in the efforts to cushion NES in these areas from the risk of extinction.  -->

As this study classifies NES using AOH, we note several limitations to our definition of NES. The first is that we consider only habitat area and not other habitat characteristics that could further restrict the true amount of usable habitat, such as habitat quality and fragmentation [@santini_applying_2019]. Similarly, while population size and trends are influenced by habitat availability, species' populations could also be suppressed by additional threats besides habitat loss that we do not consider here, including invasive species and overexploitation, for example [@sodhiCausesConsequencesSpecies2009a]. Thus, the list of NES that we present here represents the species for which habitat availability is likely to be a limiting factor in their recovery. Restoration of habitat would remove this barrier, and thus permit - but not guarantee - recovery of the many species that are threatened by habitat loss. Finally, our identification of NES was determined by comparing species' habitat area to the thresholds used by the Red List to classify species into categories of extinction risk, and these thresholds are inevitably not equally meaningful across taxa. However, as the global standard for quantifying extinction risk, although imperfect, the Red List represents a simple and standard protocol for measuring progress towards biodiversity targets [@keithScientificFoundationsIUCN2013].

Lack of habitat considerably increases species' risk of extinction. Species with small habitats are more vulnerable to environmental stochasticity [@lacyConsideringThreatsViability2000], demographic stochasticity [@pflugerHabitatLossCauses2019; @brookContributionInbreedingExtinction2002], and edge effects [@ewers_fragmentation_2013;@pfeifer_creation_2017]. Accordingly, the size of a species' geographic range is one of the five listing criteria used by the Red List [@iucn_guidelines_2019], and species with too little habitat will always be limited in the extent to which they can recover on the Red List. Habitat loss and degradation will continue into the foreseeable future [@jenkinsProspectsBiodiversity2003], presenting ever greater challenges to the species that already have too little habitat. Here, as we enter the post-2020 era of biodiversity conservation, we identify the species and regions most in need of habitat protection, where strategic expansion of PAs under "30 x 30" could have a major impact. Simultaneously, our analysis of which species are closest and furthest from the threshold for downlisting identifies, respectively, where relatively small gains in habitat area carry disproportionate benefit, versus the species for which major restoration efforts or assisted colonization will be needed to facilitate downlisting. In the latter case, identifying the species for which habitat availability is a major barrier to long-term recovery is vital for conservation organizations to effectively manage their limited resources.

## Data availability 

The datasets used to support the findings of this study are available for download by request from their respective providers. Species assessments can be accessed from the IUCN Red List of Threatened Species website at https://www.iucnredlist.org, and range maps can be requested at https://www.iucnredlist.org/resources/spatial-data-download. Protected area maps can be requested from the World Database on Protected Areas, found at https://www.unep-wcmc.org/resources-and-data/analysis/main/wdpa. 

## Code availability

Custom Python code to calculate species Area of Habitat in Google Earth Engine will be published at a later date, and is available on request from R.A.S.

## Acknowledgements

Funding for this research was provided by the High Meadows Environmental Institute and High Meadows Foundation.

## Author contributions

R.A.S. and D.S.W. conceived the study. R.A.S. collated the data. S.R. carried out the analyses and wrote the first draft of the manuscript, with substantial contribution from R.A.S. and D.S.W. 

## Competing interests

The authors declare no competing interests.

# References
