---
title: "No exit: the conservation reliance of species with too little habitat"
output:
    bookdown::word_document2:
        reference_docx: ms_general.docx
        number_sections: false
    bookdown::pdf_document2:
        toc: no
        latex_engine: xelatex
        fig_caption: yes
        keep_tex: yes
link-citations: yes
csl: conservation-letters.csl
biblio-style: apalike
bibliography: [refs.bib, funky-refs.bib, pckgs.bib]
fontsize: 12pt
geometry: margin=2.5cm
mainfont: Calibri
papersize: a4
documentclass: article
header-includes:
- \usepackage{hyperref}
- \usepackage{fancyhdr}
- \pagestyle{fancy}
- \fancyhf{}
- \fancyfoot[CO,CE]{\thepage}
- \usepackage{setspace}
- \doublespacing
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = "../")
knitr::opts_chunk$set(echo = FALSE,message = FALSE,error = FALSE, warning = FALSE,
                      dpi = 300, fig.path = "figs/main/", fig.align = "center", 
                      fig.width = 16.6/2.54, fig.height = 15/2.54)
```

```{r packages}
library(readr)
library(ggplot2)
library(stringr)
library(dplyr)
library(ggtext)
library(png)
library(grid)
library(RColorBrewer)
library(Hmisc)

# Define function to capitalise first letter
capitalise <- function(string) {
  substr(string, 1, 1) <- toupper(substr(string, 1, 1))
  string
}
```

```{r data}
load("HMEI-GDrive/data/sorted/msdata.Rdata")
```

_______________________________________________________________________________________

Sophia Richter^1^ , Rebecca A. Senior^2^*, and David S. Wilcove^1,2^

^1^Department of Ecology and Evolutionary Biology, Princeton University, Princeton, NJ, USA.

^2^Princeton School of Public and International Affairs, Princeton University, Princeton, NJ, USA.

*__Corresponding author:__ rebecca.a.senior@gmail.com (R.A. Senior)

__ORCID iDs:__ orcid.org/0000-0002-8208-736X (R.A. Senior)
  
__Authorship:__ R.A.S. and D.S.W. conceived the study. R.A.S. collated the data. S.R. carried out the analyses and wrote the first draft of the manuscript, with substantial contribution from R.A.S. and D.S.W. 

__Funding:__ High Meadows Environmental Institute and High Meadows Foundation.

__Key words:__ conservation; extinction; IUCN Red List; conservation planning; biodiversity; threatened species

__Article type:__ Perspective

__Number of words:__ 

__Number of References:__ 

__Number of Figures:__ ; __Number of Tables:__ 0 

Conservation Letters formatting guidelines:

Perspectives provide a forum for authors to discuss recent developments or ideas with broad and significant policy relevance to conservation. They should be forward looking and written with a view to informing non-specialist readers (thus should be presented using straightforward prose and avoiding jargon). Perspectives may be up to 3000 words in length and should contain no more than 5 figures and/or tables, and fewer than 30 references.

# Abstract


__/200 words__

# Introduction

Biodiversity is the foundation of ecosystem health, underpinning an array of ecosystem services [@titeux_ipbes_2019] and conferring resilience to anthropogenic change through taxonomic, functional and  genetic diversity [@oliver_biodiversity_2015]. However, the world is currently experiencing alarming rates of biodiversity loss [@wwf_living_2020; @rosenberg_decline_2019]. Across all anthropic changes on Earth, habitat conversion is the leading cause of species extinctions  [@maxwell_biodiversity:_2016;@goncalves-souzaHabitatLossExtinction2020]. The underlying driver of habitat conversion is agricultural expansion [@wwf_living_2020; @williams_proactive_2020; @gibbs_tropical_2010], which is expected to continue to pose a major threat to biodiversity. As demand for food increases, driven by population growth and eating habits, the agricultural sector is expected to grow and develop in tandem. Improvements in efficiency, such as increasing yields per area, decreasing waste, and shifting away from meat-based diets in some cultures [@tilman_global_2011; @tilman_future_2017; @williams_proactive_2020], could help to abate habitat loss for some threatened species, but many will face further conversion of habitat - particularly in the tropics [@gibbs_tropical_2010]. Protecting habitat must be at the heart of post-2020 conservation policy [@cbd_update_2020;@visconti_protected_2019;@geldmann_essential_2021].

Habitat area and fragmentation are directly related to extinction risk. Indeed, the International Union for the Conservation of Nature (IUCN) include habitat area as one of the five listing criteria for the Red List, according to which, assessed species are placed into categories of extinction risk ranging from Least Concern and Near Threatened, to the 'threatened' categories of: 'Vulnerable', 'Endangered', and 'Critically Endangered' [@mace_quantification_2008; @iucn_iucn_2020]. Ultimately, populations confined to smaller areas are more susceptible to stochasticity and edge effects. Most environmental stressors are localized, such that a population with a large area of habitat benefits from the averaging effect of environmental effects that are not over the entire range, while populations with a small habitat are at a much higher risk of extinction because environmental changes are likely to affect the entire population [@lacyConsideringThreatsViability2000]. Species with small habitat area are also more likely to be affected by demographic stochasticity, including a heightened risk of inbreeding and loss of genetic diversity [@CITE] [@pflugerHabitatLossCauses2019; @brookContributionInbreedingExtinction2002]. Compounding stochastic effects are the stresses associated with the relatively large amount of edge in small habitats, such as higher temperatures and easier access for poachers [@ewers_fragmentation_2013;@pfeifer_creation_2017]. The isolation of small, fragmented habitat can also make it difficult for species to disperse elsewhere [@crooks_global_2011;@haddad_habitat_2015], which is increasingly problematic in a warming world where species are on the move towards the poles and uphill [@mcguire_achieving_2016;@senior_global_2019]. 

Many species in the wild can only maintain a self-sustaining population with ongoing species-specific management interventions. They are referred to as "Conservation Reliant Species" [@scott_conservation-reliant_2010; @goble_conservation-reliant_2012; @scott_recovery_2005]. The extreme vulnerability to stochastic and edge effects for species with too little habitat inevitably renders them more dependent on conservation interventions to maintain viable populations. Thus, we can think of insufficient habitat area as imposing a form of conservation reliance. Regardless of the other conservation interventions that we might implement to maintain the status quo - such as invasive species control or regulation of trade - unless remaining habitat is robustly protected and total habitat area eventually increased, we cannot hope to recover these habitat-limited species. Yet, recover them we must, if we are to meet Goal 15 of the Sustainable Development Goals (SDGs): to "...halt and reverse land degradation and halt biodiversity loss" [@TransformingOurWorld2015]. 

As conservation policy is updated in the post-2020 era, underpinned by huge international efforts to document the state of nature [@ipbes_global_2019], now is the time to identify where habitat recovery begets species recovery. Habitat protection falls directly in line with the proposed "30 x 30" goal to protect 30% of the world’s land and coastal seas by 2030 [@30by30_2021], which has recently been adopted by countries such as the UK and USA [@CITE] [@bidenExecutiveOrder140082021; @50CountriesAnnounce], and is anticipated to be adopted by the UN Biodiversity Conference held in Kunming, China in October 2021 [@CITE] [@ConventionBiodiversityProposes]. Simultaneously, the undoing of previous loss and degradation sits within the remit of large-scale restoration initiatives, such as the Bonn Challenge, which aims to restore 350 million hectares of degraded and deforested landscapes by 350 million hectares by 2030 [@BonnChallenge2016]. 

Not only do we see, empirically, that habitat availability is a major barrier to species' recovery [@CITE] [@huxelHabitatLossFragmentation1999; @kerrHabitatLossLimits2004], but in practice, a species cannot be downlisted to a lower threat category on the Red List while its 'geographic range' still qualifies it for the higher category [@iucn_iucn_2020]. Under criterion B, species are classed as Vulnerable, Endangered, or Critically Endangered, if the total 'Area of Occupancy' (hereafter: AOO; calculated as the sum of occupied cells within a 2 x 2 km grid) [@iucn_guidelines_2019] falls below the thresholds of 2000, 500 and 10km^2^, respectively. AOO represents a subset of the more commonly used metric 'Extent of Occurrence' (hereafter: 'EOO'), defined as the area of the minimum polygon that can be drawn to encompass all known, projected or inferred occurrences of a species [@iucn_guidelines_2019]. While EOO is less accurate because it may contain several sites that are ecologically unsuitable for a species, AOO is difficult to measure. However, recent advances in satellite data and resulting habitat maps [@jung_global_2020], coupled with extensive data on species' habitat preferences [@iucn_iucn_2020;@birdlife_2020], now make it possible to refine species' EOO to the areas within a species' range that have an appropriate habitat type and elevation [@brooks_measuring_2019]. The refined map is referred to as 'Area of Habitat' (hereafter: AOH), and as AOH is functionally equivalent to AOO, the same AOO thresholds can be used to estimate species' minimum Red List category [@santini_applying_2019]. Thus, we have unprecedented capacity to quantify how much habitat threatened species have, and the extent to which habitat area limits species' recovery potential. 

By comparing species' AOH to the Red List thresholds, we identify two related and important facets of conservation reliance for habitat-limited species, based on habitat area alone. The first is how close a species is to being uplisted on the Red List to a higher threat category. The second is how far a species is from being downlisted to a lower threat category. We refer to all species with AOH falling below any of the Red List threatened thresholds as 'No Exit Species' (hereafter: NES), because they can never exit their threatened status without concerted efforts to protect and restore habitat. In the case of NES close to the uplisting threshold, it is essential to protect what little habitat remains, which can be achieved through strategic expansion of the Protected Area (PA) network under 30 x 30 [@30by30_2021]. For all NES, there is a need to reverse historic habitat loss, which can only be achieved through controversial translocation of species to new habitats outside of their current range [@CITE] [@lawlerReframingDebateAssisted2011; @luntUsingAssistedColonisation2013], or through restoration efforts like the Bonn Challenge [@BonnChallenge2016]. It is also imperative that we identify NES in order to pragmatically allocate limited conservation resources. Conservation is not cheap [@weiseFinancialCostsLarge2014;@mccarthy_financial_2012]. Conservation stakeholders must consider whether other conservation interventions are worthwhile as a means to maintain the status quo when gains in habitat area cannot be achieved, or whether such efforts are a drain on limited resources for species with no hope of recovery. Restoration efforts, meanwhile, should ideally be targeted towards the species that would benefit most. 

# Methods

## Area of Habitat

To identify species whose recovery is limited by habitat area (NES), we calculated Area of Habitat (AOH) following @brooks_measuring_2019. In brief, AOH maps were produced for all threatened, terrestrial species in taxonomic groups that have range maps and have been comprehensively assessed by the IUCN Red List to at least family level (all amphibians, birds and mammals, plus select families in the classes Magnoliopsida and Reptilia). Range maps were downloaded from the websites of the Red List [@iucn_iucn_2020] and BirdLife International [@birdlife_2020], and refined to AOH by retaining only grid cells that fall within the range polygons, and meet the species' habitat and elevation preferences as defined by the Red List [@iucn_iucn_2020]. All spatial analyses were conducted in Google Earth Engine[@gorelick_google_2017].

Following Red List assessment protocol [@iucn_guidelines_2019], we exclude habitats that are defined by the Red List as 'marginal' in their suitability, or 'passage' in their seasonality of use. As we are interested in the minimum habitat area needed to sustain a healthy global population, for migratory species we consider only the smaller of either the breeding or non-breeding range, in keeping with the Red List protocol [@iucn_guidelines_2019]. We do, however, retain habitat types listed with 'unknown' suitability or seasonality, because this likely affects many threatened species that are too rare or poorly known for such classification to be possible.

To identify cells with suitable habitat, we used the 1 km-resolution IUCN habitat types map for the year 2015 [@jung_global_2020]. Level 1 habitat categories refer to the broad categories of Forest, Savanna, Shrubland, Grassland, Wetlands (inland), Rocky Areas, Deserts and Artificial habitats, each of which contains finer level 2 categories, e.g., 'Forest - Boreal'. Where a species' habitat preference is given only to level 1, we conservatively assume that all nested level 2 categories are also suitable. To identify cells with suitable elevation, we use the EarthEnv digital elevation model [@robinson_earthenv-dem90:_2014]. All maps were resampled to 2 km, following @iucn_guidelines_2019, projected to a Mollweide projection to ensure accuracy of area calculations. 

To quantify the extent to which AOH is a limiting factor in decreasing a species' risk of extinction, we compare AOH to the fixed thresholds used by the Red List to assign species to the Red List categories. The Red List use five criteria for this classification, and species are assigned to the highest risk category for which they qualify under any of the criteria [@iucn_guidelines_2019]. Here, we are specifically referring to criterion B, under which species are classified based on their geographic range. The area of a species' range can be measured based on Extent of Occurrence (EOO) or Area of Occupancy (AOO). AOH is analagous to AOO and can be used in the same way, using the same thresholds [@brooks_measuring_2019]. 

For the three 'threatened' Red List categories of Vulnerable, Endangered and Critically Endangered, the AOH thresholds are, respectively: < 2000 km^; < 500 km^2; and, < 10 km^2. We categorise species as 'No Exit' if their AOH falls below any of the three thresholds. We further require that species are at least 1 km^2 from that threshold, based on the assumption that 1 km^2 is a trivial difference that can be easily gained or lost, and is within the expected error of the spatial data. Species that are less than 1 km^2 from the threshold are considered to have met the threshold. We note that when the Red List use these thresholds for classification, they additionally require that two out of three subcriteria are met. However, several additional metrics would be needed to assess these subcriteria (habitat fragmentation, and temporal trends in population size or habitat quantity or quality) [@santini_applying_2019]. Such classification is beyond the scope of this study, as we are interested not in automating Red List classification, but rather in the extent to which habitat quantity limits our ability to improve species' prospects. Here, when a species is classed as 'No Exit', we are referring to situations where no matter what other actions we might take to address threats and improve outcomes for threatened species, their restricted AOH means that NES will always be limited in the extent to which extinction risk can be mitigated.

In downstream analyses, we first summarize the taxonomy and biogeography of NES. Next, we explore the magnitude of species' reliance on habitat by calculating the difference between species' AOH and the threshold for downlisting. This difference is the amount of habitat that a species would need to gain to have any chance of downlisting to a lower category of extinction risk on the Red List, regardless of what other conservation actions we might take. We summarize both the distribution of these differences from the downlisting threshold, as well as the upper quantile of species within each habitat category that are furthest from the threshold - the 'Extreme No Exit' species (ENES). ENES are the species with the most additional habitat required to safeguard them from extinction, and thus the species most likely to require major and concerted effort to recover. We determined the countries and realms in which each species occurs by overlaying species' extant range maps with maps of global administrative areas [@gadm_2018] and zoogeographic realms [@holt_update_2013].  

## Statistical analyses

All statistical analyses were conducted in R version 4.0.3 [@r_core_team_2020]. The degree to which species are constrained from future downlisting by current AOH was captured by the relative distance of AOH from the downlisting threshold. This 'proximity to threshold' measure was calculated as the absolute difference between a species' AOH and the threshold for downlisting to the next lowest Red List category, divided by the maximum possible difference. Thus, proximity to threshold is a standardised measure that accounts for the fact that thresholds between categories do not increase linearly. As an example, a species with 1250 km^2^ AOH has a larger absolute difference to the downlisting threshold (2000 - 1250 = 750 km^2^) than a species with 5 km^2^ AOH (10 - 5 = 5 km^2), but both have a relative proximity to threshold of 0.5. We used a generalized linear model (GLM) with a binomial distribution, using the function `glm` in the `lme4` R  package [@bates_fitting_2015], to test whether proximity to threshold varied by taxonomic class, biogeography and absolute AOH. Model inference was based on Likelihood Ratio Tests, dropping each variable in turn and comparing the reduced model to the full model [@zuur_mixed_2009].

# Results

Of the 3,869 species with AOH data, 2,488 species have habitat areas below the 2000 km^2 threshold to be considered as NES (Figure \@ref(fig:class-and-threshold-graph)). Of the 2,488 NES, most species were amphibians (1389), followed by birds (527), mammals (461), assessed plants in the class Magnoliopsida (59), and, finally, assessed reptiles (52). By Red List threshold, 335 species fall below the 10 km^2 threshold for Critically Endangered, 720 between the 10 km^2 and 500 km^2 thresholds (Endangered), and 1433 have between 500 km^2 and 2000 km^2 AOH (Vulnerable). We also identified ENES, as the upper quartile of each threshold group (top 25%) that had AOH values furthest from the downlisting threshold. Of the 623 species in the upper quartile, 358 were amphibians, followed by birds (124), mammals (113), assessed reptiles (15), and assessed plants in the class Magnoliopsida (13).

```{r class-and-threshold-graph}
nes <- nes_df %>% 
  mutate(Category = str_replace(Category, "Critically Endangered", "AOH < 10 km^2")) %>% 
  mutate(Category = str_replace(Category, "Endangered", "10 < AOH < 500 km^2")) %>% 
  mutate(Category = str_replace(Category, "Vulnerable", "500 < AOH < 2000 km^2"))

# relevel Categories
nes$Category <- 
  factor(nes$Category, 
         levels = c("AOH < 10 km^2", "10 < AOH < 500 km^2", "500 < AOH < 2000 km^2"))
nes$className <- 
  factor(nes$className, 
         levels = unique(nes$className),
         labels = capitalize(unique(nes$className)))

# Code to add pics to labels
classes <- levels(nes$className)
labels <- paste(
  "<img src=", 
  "'",
  file.path("HMEI-GDrive/resources/phylopic", paste(tolower(classes), 'png', sep = ".")),
  "' ",
  "width='10' />", 
  classes,
  sep = ""
)
names(labels) <- classes

max_val <- 834

ggplot(nes, aes(x=className, fill=className)) + 
  geom_bar() + 
  facet_wrap(~Category, labeller = label_wrap_gen(width = 25, multi_line = TRUE)) +
  theme_bw() + theme(panel.grid = element_blank()) +
  ggtitle("Class and AOH Threshold of NES") + 
  labs(fill="Class") +
  xlab(" ") + 
  ylab("Count") + 
  theme(legend.position = "none") + 
  scale_fill_brewer(palette="Set1") +
  scale_y_continuous(breaks = seq(0, max_val, by = 100), limits=c(0, max_val)) +
  theme(axis.text.x = element_markdown(color = "black",
                                       angle = 45, 
                                       hjust = 1, vjust = 1,
                                       margin = unit(c(-5, 0, 0, 0), "pt"))) +
  scale_x_discrete(labels = labels)
```

<!-- Added stats for p-value and degrees of freedom when each variable was dropped for GLM. Residual deviance for all three was 713.16 -->

Across class and Red List thresholds, there is a right skew towards a large difference between species' AOH and the downlisting threshold, particularly for those species with AOH between 10 km^2 and 500 km^2 (Figure \@ref(fig:difference-histogram)). This suggests that, within each range of AOH values (bounded by the different the Red List thresholds), the majority of species have AOH that is as far from the threshold for downlisting as possible. In other words, most species are closer to the threshold for uplisting than for downlisting. We found that proximity to downlisting threshold did not differ between taxonomic classes (p-value = 0.988, 4 DF) and biogeographic realms (p-value = 0.999, 6 DF), but increased with increasing absolute AOH (p-value < 2.2e-16, 1 DF). Thus, species are relatively further from the threshold for downlisting when absolute habitat quantity is lower.

<!-- Should the following be included if it is already said above? -->

The GLM ran on proximity to threshold varying by taxonomic class, biogeography and absolute AOH found that only absolute AOH had a significant effect on proximity to threshold, while taxonomic class and realm had no significance. 

```{r difference-histogram}
ggplot(nes, aes(x=Difference, fill=className)) + 
  geom_histogram(alpha=0.7) + 
  ggtitle("Distribution of Differences to Next Threshold \n Across Class and Category") +
  facet_wrap(~ Category, scales = "free", 
             labeller = label_wrap_gen(width = 15, multi_line = TRUE)) + 
  theme_bw() + theme(panel.grid = element_blank()) +
  labs(fill="Class") + 
  ylab("Count") + xlab("Difference (km^2)") + 
  theme(axis.text.x = element_text(angle = 45)) + 
  scale_fill_brewer(palette="Set1") 
```

The three realms with the most NES are Neotropical (489 species), Oriental (439 species), and Panamanian (431 species). These were also the three realms with the most ENES, although there are more ENES in Panamanian than Oriental. Mapping total NES and ENES by country (Figure 3), we see that certain countries - such as Colombia, Peru, Mexico and Madagascar - harbour a striking proportion of the world's most habitat-limited species.

```{r map}
map <- 
  readPNG("HMEI-GDrive/figs/Final Maps/combi_country_grid.png") %>% 
  rasterGrob(interpolate = TRUE)
ggplot() +
  theme_void() +
  scale_x_continuous(expand = c(0,0)) +
  scale_y_continuous(expand = c(0,0)) +
  annotation_custom(map,
                    xmin = -Inf, xmax = Inf,
                    ymin = -Inf, ymax = Inf)
```

# Discussion

* P1 = summary of basic summary stats e.g. "Recent advances in quantifying species' habitat area, facilitated by the new-found availability of global habitat maps and data on species' habitat preferences, allowed us to quantify for the first time the number of threatened amphibian, bird and mammal species whose recovery on the Red List is limited by habitat area. We found that X species have AOH below the Red List thresholds. We refer to these as NES because... From the list of NES, most were amphibians, followed by ...."
* P2 = summary of geographic distribution. Where are NES? What characterises these counties (i.e. all in the Neotropics? Tropical?). Any surprises?
* P3 = summary of distance from downlisting threshold. Most species closer to uplisting - need to protect what habitat remains (30 x 30), or situation could get worse fast
* P4 = focus on how much habitat restoration is needed. Is this feasible? Focus on countries with most NES that stand to gain. Link to global and national restoration agendas
* P5 = caveats. Only considered habitat area (not other facets of habitat, like quality or fragmentation, and nothing to do with population size or threats other than habitat loss). Limited by available data (biased to particular taxa and regions with better data, AOH likely wrong in some places). Use RL thresholds becuase RL is a global standard, but some species don't need as much habitat area.
* P6 = conclusion. Bring it back to why habitat area is important (stochasticiy, edge effects etc.) and fact that species cannot be downlisted if limited by habitat, and that many species are threatened by habitat loss and degradation (which will continue). Need to act now to save species on the brink, and while the time is ripe within global conservation policy. Also need this info for triage, which is important when resources are limited

<!-- I have all of the citations for the discussion section ready, I am planning to add them after the first round of edits -->

<!-- Should I talk about the cat change data and why certain species are -2 change in some part of the discussion? --> 

<!-- I am happy to change species/statistic examples anywhere in the discussion and/or make those sections shorter (I made them longer in case you wanted to decide which parts are relevant) --> 

Recent advances in quantifying species' habitat area, facilitated by the new-found availability of global habitat maps and data on species' habitat preferences, allowed us to quantify for the first time the number of threatened amphibian, bird and mammal species whose recovery on the Red List is limited by habitat area. We found that 2,488 species have AOH below the thresholds used by the Red List to categorize species as threatened (i.e. Vulnerable, Endangered, Critically Endangered), according to their geographic range (criterion B). We refer to these as "No Exit Species" (NES) because they have a habitat that inhibits recovery and increases the risk of extinction. Even for those species that have stable population trends or occupy protected areas, a single event can greatly affect the population and lead to uplisting. Without habitat restoration or an increase in habitat area, NES will continue to be limited by their habitat. The 2,488 species that are NES account for over half of the total 3,869 species that have AOH data. From the list of NES, most were amphibians, followed by birds and mammals. Key amphibians from the list include Archey’s Frog, the Kroombit Tinker Frog, the Eungella Day Frog, and Barrio’s Frog, each of which are top 10 Evolutionarily Distinct and Globally Endangered (EDGE) amphibians [@CITE]. Among the NES include WWF priority species the Iberian Lynx, the Tapanuli orangutan, and the Indian rhinoceros [@CITE]. ENES include the Red Wolf, protected under the US Endangered Species Act and listed as Critically Endangered by the IUCN [@CITE], and the Golden Lion Tamarin, a key species for seed dispersal in the Atlantic coastal regions of southeastern Brazil [@CITE]. 


A spatial analysis revealed the geographic regions and countries with the highest number of NES and ENES, among other metrics such as proportion of species and number of species with the smallest difference to threshold. The countries with the largest total NES are Colombia, Mexico, and Madagascar. This is the same for ENES, with the inclusion of Indonesia as well. The Panamanian and Neotropical Realms contain the most NES amphibians, Neotropical and Oriental the most birds, and the Oriental and Afrotropical realms contain the most mammals. NES amphibians are largely concentrated in the countries from Mexico and Colombia, birds in the northern half of South America and Indonesia, and mammals in Madagascar and Indonesia. These specific geographic regions and countries have a high number of NES species and will need to be a main focus when assessing where and for what species will require habitat restoration. 

The difference between absolute AOH and the threshold for downlisting for the NES shows a clear skewed distribution towards having a large difference to the next threshold, particularly for species with AOH between 10 and 500 km^2. This indicates that most NES are close to the habitat threshold for uplisting, and that if the species loses that amount of habitat it will either become uplisted to the new threat level or is at risk of uplisting to that level in the event of something affecting the population. To prevent this from occurring, it is important to protect what remaining habitat is left for these species. The "30 x 30" goal will aid in this international effort for habitat protection [@30by30_2021]. Ideally, part of the 30% of land and coastal seas restored by this project will be areas occupied by NES and ENES. 

<!-- I decided to create a map of the 25% species with the smallest difference to next threshold to assess the absolute numbers - it is also added to the google drive --> 

The sum of all the differences to the habitat threshold for a lower category is 1,165,744 km^2 of habitat, while the average difference is 468 km^2. For species with AOH < 10 km^2 the total differences sum to 2,115 while the average is 6.31 km^2. Species between 10 and 500 km^2 sum to 500,917 km^2 and the remaining NES have a sum of 6,62,710 km^2 differences. Although some of these habitat areas are likely overlapping and restoring one area affects multiple species, there is still a lot of ‘ground to cover’ so to speak in order for each of these NES to have enough habitat restored to qualify for a downlisted habitat threshold category. It is certainly not feasible to restore all 1.17 million km^2 of habitat, but it may be feasible to focus on restoring the smaller goal of the habitats of species with less than 10 km^2 or the habitats of ENES (totalling 422,503 km^2 of habitat differences). Countries and regions with a large number of NES that require only a small increase in habitat area to qualify for the downlisted category stand the most to gain in habitat restoration. These countries include Colombia with over 60 species having the smallest 25% differences, Mexico, Madagascar, and Indonesia with 40-50 species, and China, Peru, India, and Tanzania each with around 30 species. Notably, these are many of the same countries that also have the most ENES, species that would also benefit from habitat restoration in addition to habitat protection. 

The Initiative 20x20 aims to protect and restore land in Latin America and the Caribbean in support of the Bonn Challenge and the New York Declaration on Forests. Colombia has pledged to restore 1 million ha of degraded land by 2020, Mexico plans to restore nearly 8.5 million ha of degraded land by 2020, and Peru to restore 3.2 million hectares of degraded land [@CITE]. Each of these pledges are to both the Initiative 20x20 and the Bonn Challenge. In addition, Madagascar has pledged 4,000,000 hectares, Tanzania 5,200,000 hectares, and India 21,000,000 hectares to the Bonn Challenge [@BonnChallenge2016]. China and India have yet to pledge any habitat to the Bonn Challenge. Indonesia is currently implementing the Peatland Restoration Agency (BRG), which was extended to 2024 and aims to restore over 2.6 million hectares of forest and peatlands destroyed by fires in 2015 [@CITE]. However, the general effects of deforestation and a lack of habitat restoration outside of areas affected by the fires in 2015 contribute to a serious need for further policies and organizations to restore land in Indonesia. Although China has not pledged to the Bonn Challenge, the government is at the forefront of restoration efforts, including the the Natural Forest Conservation Program (NFCP), the Conversion of Cropland to Forest Program (CCFP) as part of the NFCP with considerable success in reforesting 28.2 million hectares [@CITE]. Global and local government agendas to restore habitat in these countries are important for many NES identified, and further policies to support restoration in these areas will aid in the efforts to cushion NES in these areas from the risk of extinction. 

Although this study classifies NES using AOH, there are several factors to consider in relation to the parameters used to identify NES. The first is that this study only considers habitat area and not other facets of habitat like quality or fragmentation, or other important factors that affect whether a species is at risk of extinction like population size and trends, invasive species, illegal trade, and climate change. Of the many species that live on this planet, only 3,869 species had available AOH data. The species with available data are biased to particular taxa and regions, and AOH may be likely wrong in some areas where data was not easily available. Additionally, the habitat area of each species was compared to the AOH thresholds on the IUCN Red List because it is a global standard. However, some species don’t need as much habitat area as the Red List thresholds set. These species were therefore classified as NES due to absolute AOH, but it should be noted that they may not be in danger of extinction due to lack of habitat. 

Lack of habitat area and fragmentation considerably increases the risk of extinction, due to increasing species vulnerability to threats that are specific and exacerbated by small habitat area. Species with small habitats are more vulnerable to environmental stochasticity which can affect the entire population if the habitat area is small [@lacyConsideringThreatsViability2000], demographic stochasticity such as inbreeding and genetic diversity loss [@pflugerHabitatLossCauses2019; @brookContributionInbreedingExtinction2002], and edge effects like temperature and poaching [@ewers_fragmentation_2013;@pfeifer_creation_2017]. Small habitat also prevents a species from becoming downlisted by limiting the species to geographic range criteria on the IUCN Red List. As habitat loss and degradation continues into the foreseeable future, NES and other species will face big challenges and an increased risk of extinction due to small habitats. In order to save species currently on the brink of extinction and those with extremely small habitats, we need to act now while global conservation policy is continuing to push habitat restoration and protection. It is important to identify the species with the least amount of AOH to assess which geographic regions stand the most to gain from habitat restoration and protection, and which species would benefit from habitat conservation versus other techniques such as translocation or invasive species management. With the limited resources available to conservation groups and NGOs, this list of NES will help organizations effectively manage their resources and efforts depending on a species’ needs and habitat area. 


## Data availability 

The datasets used to support the findings of this study are available for download by request from their respective providers. Species assessments can be accessed from the IUCN Red List of Threatened Species website at https://www.iucnredlist.org, and range maps can be requested at https://www.iucnredlist.org/resources/spatial-data-download. Protected area maps can be requested from the World Database on Protected Areas, found at https://www.unep-wcmc.org/resources-and-data/analysis/main/wdpa. 

## Code availability

Custom Python code to calculate species Area of Habitat in Google Earth Engine will be published at a later date, and is available on request from R.A.S.

## Acknowledgements

Funding for this research was provided by the High Meadows Environmental Institute and High Meadows Foundation.

## Author contributions

R.A.S. and D.S.W. conceived the study. R.A.S. collated the data. S.R. carried out the analyses and wrote the first draft of the manuscript, with substantial contribution from R.A.S. and D.S.W. 

## Competing interests

The authors declare no competing interests.

# References
