---
title: "No exit: the conservation reliance of species with too little habitat"
subtitle: 'Supplementary Information'
output:
    bookdown::pdf_document2:
        toc: no
        latex_engine: xelatex
        fig_caption: yes
        keep_tex: yes
        includes:
            in_header: header.tex 
    bookdown::word_document2:
        reference_docx: ms_general.docx
    bookdown::html_document2:
        toc: no
fontsize: 12pt
geometry: margin=2.5cm
papersize: a4
documentclass: article
link-citations: yes
csl: elsevier-harvard.csl
biblio-style: apalike
bibliography: [refs.bib, funky-refs.bib, pckgs.bib]
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = "../")
knitr::opts_chunk$set(echo = FALSE,message = FALSE,error = FALSE, warning = FALSE,
                      dpi = 300, fig.path = "figs/som/", fig.align = "center", 
                      fig.width = 16.6/2.54, fig.height = 15/2.54)
```

```{r packages}
library(readr)
library(ggplot2)
library(stringr)
library(dplyr)
library(ggtext)
library(png)
library(grid)
library(RColorBrewer)
library(Hmisc)

# Define function to capitalise first letter
capitalise <- function(string) {
  substr(string, 1, 1) <- toupper(substr(string, 1, 1))
  string
}
```

```{r data}
load("HMEI-GDrive/data/sorted/msdata.Rdata")
```

_______________________________________________________________________________________

Sophia Richter^1^ , Rebecca A. Senior^2^*, and David S. Wilcove^1,2^

^1^Department of Ecology and Evolutionary Biology, Princeton University, Princeton, NJ, USA.

^2^Princeton School of Public and International Affairs, Princeton University, Princeton, NJ, USA.

*__Corresponding author:__ rebecca.a.senior@gmail.com (R.A. Senior)

__ORCID iDs:__ orcid.org/0000-0002-8208-736X (R.A. Senior)

\pagebreak
\raggedright
\setcounter{tocdepth}{2}
\tableofcontents
\beginsupplement

\pagebreak

# Supplementary Figures

(ref:cap-1) The total number of threatened species within each country classified as 'No Exit Species' (NES). Results are summarised for each taxonomic group (rows), , and shaded from low NES count (yellow) to high (red).

```{r fig-s-1, fig.cap = "(ref:cap-1)", fig.width = 16.6/2.54, fig.height = 18/2.54}
map <- 
  readPNG("HMEI-GDrive/figs/NES (new data) figs/nes_map_country.png") %>% 
  rasterGrob(interpolate = TRUE)
ggplot() +
  theme_void() +
  scale_x_continuous(expand = c(0,0)) +
  scale_y_continuous(expand = c(0,0)) +
  annotation_custom(map,
                    xmin = -Inf, xmax = Inf,
                    ymin = -Inf, ymax = Inf)
```

\pagebreak

# Supplementary Tables

\pagebreak

# Supplementary References

