---
title: "No exit: the conservation reliance of species with little remaining habitat"
output:
    bookdown::word_document2:
        reference_docx: "../HMEI-GDrive/resources/word-templates/ms_general.docx"
        number_sections: false
    bookdown::pdf_document2:
        toc: no
        latex_engine: xelatex
        fig_caption: yes
        keep_tex: yes
link-citations: yes
csl: conservation-biology.csl
biblio-style: apalike
bibliography: [refs.bib, funky-refs.bib, pckgs.bib]
fontsize: 12pt
geometry: margin=2.5cm
mainfont: Calibri
papersize: a4
documentclass: article
header-includes:
- \usepackage{hyperref}
- \usepackage{fancyhdr}
- \pagestyle{fancy}
- \fancyhf{}
- \fancyfoot[CO,CE]{\thepage}
- \usepackage{setspace}
- \doublespacing
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = "../")
knitr::opts_chunk$set(echo = FALSE,message = FALSE,error = FALSE, warning = FALSE,
                      dpi = 300, fig.path = "figs/main/", fig.align = "center", 
                      fig.width = 16.6/2.54, fig.height = 15/2.54)
```

```{r packages}
library(readr)
library(kableExtra)
library(ggplot2)
library(stringr)
library(dplyr)
library(ggtext)
library(png)
library(grid)
library(RColorBrewer)
library(Hmisc)

# Define function to capitalise first letter
capitalise <- function(string) {
  substr(string, 1, 1) <- toupper(substr(string, 1, 1))
  string
}
```

```{r data}
load("HMEI-GDrive/data/sorted/msdata.Rdata")
```

```{r functions}
# Function to make significance levels pretty
siglevel <- function(pval){
  ifelse(pval < 0.001, "P < 0.001",
         ifelse(pval < 0.05, 
                paste("P = ", signif(pval, 2), sep = ""),
                "P > 0.05 ")
         )
}
```

_______________________________________________________________________________________

Sophia Richter^1^&Dagger;, Rebecca A. Senior^2,3^*&Dagger;, and David S. Wilcove^1,2^

^1^Department of Ecology and Evolutionary Biology, Princeton University, Princeton, NJ, USA.

^2^Princeton School of Public and International Affairs, Princeton University, Princeton, NJ, USA.

^3^Conservation Ecology Group, Department of Biosciences, Durham University, Durham DH1 3LE, UK

*__Corresponding author:__ rebecca.a.senior@gmail.com (R.A. Senior)

&Dagger;These authors contributed equally to this work

__ORCID iDs:__ orcid.org/0000-0002-8208-736X (R.A. Senior)
  
__Authorship:__ R.A.S. and D.S.W. conceived the study. R.A.S. collated the data. S.R. carried out the analyses and wrote the first draft of the manuscript, with substantial contribution from R.A.S. and D.S.W. 

__Funding:__ High Meadows Environmental Institute and High Meadows Foundation.

__Key words:__ conservation; extinction; IUCN Red List; conservation planning; biodiversity; threatened species

__Running head:__ Limits on recovery by habitat area

__Article type:__ Practice and Policy

__Number of words:__ 4627/6000

__Number of References:__ 44

__Number of Figures:__ 3; __Number of Tables:__ 1

# Abstract

Species that are "conservation reliant" require specific and continued management to maintain self-sustaining populations. The global standard for assessing species’ risk of extinction is the International Union for Conservation of Nature Red List of Threatened Species (Red List). Species can be listed as threatened based on habitat area and, conversely, improvements in threat status are constrained by habitat area. This reflects the real risks to species whose populations are confined to habitat of insufficient area. Recent advances in satellite data have allowed us to create habitat maps and quantify Area of Habitat (AOH) for `r nrow(all_df)` threatened terrestrial species. Of these, we find that `r nrow(nes_df)` species have AOH < 2,000 km^2^ - the threshold for classifying species as threatened, under criterion B2. For these "No Exit Species", no amount of effort will permit their exit from threatened status unless accompanied by gains in habitat area. The average amount of additional habitat required is sobering; most No Exit Species are closer to the threshold for uplisting (increased extinction risk) than the threshold for downlisting (decreased extinction risk), and species at greatest risk of extinction tend to be furthest from the downlisting threshold. The majority of No Exit Species are concentrated in a few key countries, including Colombia, Mexico, Madagascar, and Indonesia. As we establish the framework for post-2020 conservation policy, it is vital that we allocate limited conservation resources to where they can have the biggest impact. Expansion of Protected Areas under "30 x 30" will help to safeguard species on the brink, while firm commitments to habitat restoration - through initiatives like the UN Decade on Ecosystem Restoration - will be pivotal for the recovery of the majority of threatened species. Without such concerted effort, No Exit Species will be conservation reliant for the long haul.

__295/300 words__

##  Article Impact Statement

Recovery of threatened species is fundamentally constrained by habitat area; protection and restoration remain priorities for conservation.

__139/140 characters__

# Introduction

Many threatened species can maintain self-sustaining population in the wild only with ongoing species-specific management interventions. They are referred to as "conservation reliant species" [@scott_conservation-reliant_2010; @goble_conservation-reliant_2012; @scott_recovery_2005]. With continual management, such as control of invasive species or prescribed fires where early successional habitat is required, conservation reliant species can be sustained and may even rebound [@bocettiUsingConservationManagement2012]. With cessation of management, however, such species quickly decline. Here, we argue that an additional form of conservation reliance is inherent in cases where species recovery is fundamentally constrained by the amount of available habitat. We calculate how many of the world's threatened species fall into this category, and therefore require habitat restoration if they are to attain viable, self-sustaining populations. In the absence of restoration, such species will require close monitoring and, in many cases, active management (e.g., translocating individuals to maintain genetic diversity, captive propagation).  Decision makers must account for the cost of these steps in developing conservation plans if they wish to avert the extinction of these species.

Species with limited amounts of habitat face increased susceptibility to edge effects and environmental and demographic stochasticity, and thus a heightened risk of extinction and a limit to recovery potential. Many environmental stressors are localized, such that a population with a large amount of habitat benefits from the averaging effect of environmental effects that are not acting over its entire range. Populations with a small habitat, in contrast, are at much greater risk because comparable stressors are likely to affect the entire population [@lacyConsideringThreatsViability2000], a manifestation of environmental stochasticity. Demographic stochasticity also disproportionately impacts species with restricted ranges, through heightened risk of inbreeding and loss of genetic diversity, for example [@pflugerHabitatLossCauses2019; @brookContributionInbreedingExtinction2002]. Compounding such stochastic effects are the stresses associated with the relatively large ratio of edge to interior habitat that prevails in small habitat patches, including higher temperatures and easier access for poachers [@ewers_fragmentation_2013;@pfeifer_creation_2017]. Meanwhile, the isolation of small, fragmented habitat patches can impede dispersal [@crooks_global_2011;@haddad_habitat_2015]; this will be increasingly problematic in a warming world where populations of species must move to remain in climatically suitable habitat, often towards the poles or uphill [@mcguire_achieving_2016;@senior_global_2019]. 

Regardless of the other conservation interventions (e.g., control of predators, habitat enhancement) that might be implemented to maintain populations of threatened species, unless remaining habitat is robustly protected and total habitat area is increased, habitat-limited species will always face a significant risk of extinction. Accordingly, area of available habitat is one of five criteria for categorising species' risk of extinction on the International Union for Conservation of Nature (IUCN) Red List of Threatened Species. The Red List is the global standard for assessing the state of biodiversity, with categories of extinction risk ranging from Least Concern and Near Threatened to the 'threatened' categories of: 'Vulnerable' 'Endangered,' and 'Critically Endangered' [@mace_quantification_2008; @iucn_iucn_2020]. As nations develop their post-2020 goals for biodiversity protection as part of the Convention on Biological Diversity, it is imperative that they identify those species whose recovery is fundamentally limited by available habitat, in order to set realistic expectations regarding conservation success. Moreover, to pragmatically allocate limited conservation resources, decision makers must be able to identify situations where habitat protection and restoration will have the most impact in terms of saving species.

The Red List sets quantitative criteria for each of its extinction risk categories [@iucn_guidelines_2019]. These criteria include absolute population sizes and trends, but also the quantity and quality of species' habitat. Under the 'geographic range' criterion (criterion B), species are classed as Vulnerable, Endangered, or Critically Endangered, if their habitat area falls below specific thresholds. A simple but pertinent observation thus emerges: a threatened species cannot be moved ("downlisted") to a lower threat category on the Red List if the total area of its available habitat qualifies it for a higher threat category [@iucn_iucn_2020]. In other cases, species will have a large total amount of habitat that exceeds the threshold to be classed as threatened, yet they qualify as threatened because of their small or declining population size. This includes some well-known threatened species with large geographic distributions, such as the leopard (_Panthera pardus_), giant anteater (_Myrmecophaga tridactyla_) and yellow-ridged toucan (_Ramphastos culminatus_).

Previously, estimates of habitat area were largely confined to 'extent of occurrence' (criterion B1), defined as the area of the minimum polygon that can be drawn to encompass all known, projected or inferred occurrences of a species [@iucn_guidelines_2019]. These range polygons are available for many species, but frequently suffer from 'errors of commission' when areas exist within the range that are not ecologically suitable for the species, often because the elevation or habitat type is not appropriate. Recent advances in data availability and resolution [@jung_global_2020], coupled with extensive data on species' habitat preferences [@iucn_iucn_2020; @birdlife_2020], now permit the refinement of species' extent of occurrence to the areas within a species' range that have an appropriate habitat type and elevation [@brooks_measuring_2019]: 'Area of Habitat' (AOH). With AOH maps we have unprecedented capacity to quantify both how much habitat each threatened species has, and the extent to which habitat area limits its recovery potential.

We explore the conservation reliance of habitat-limited species, based on habitat area alone, by comparing threatened species' AOH area to the thresholds used by the Red List (criterion B2). The habitat area of every species will fall somewhere between the threshold for moving the species to a higher threat category ("uplisting"), and the threshold for moving it to a lower threat category ("downlisting"). These thresholds, in turn, can be used to estimate roughly the prospects for near and mid-term conservation success.  Thus, the proximity of a species' habitat area to a threshold for uplisting identifies that species as being on the brink, one whose remaining habitat must be protected to safeguard it from further decline. Conversely, the proximity of a species' habitat area to the threshold for downlisting identifies that species as potentially being an 'achievable win', whereby a relatively small increase in habitat could permit that species to be downlisted. More ominously, because these thresholds identify species with habitat area below the threshold for categorization as 'Vulnerable', they implicitly highlight the species that can never exit threatened status without concerted efforts to restore additional habitat.  As such, they are destined to remain conservation dependent until restoration efforts succeed in increasing their AOH and, presumably, their population sizes. We use the term 'No Exit Species' to describe species in this precarious state.

<!-- Biodiversity is the foundation of ecosystem health, underpinning an array of ecosystem services [@titeux_ipbes_2019] and conferring resilience to anthropogenic change through taxonomic, functional and  genetic diversity [@oliver_biodiversity_2015]. However, the world is currently experiencing alarming rates of biodiversity loss [@wwf_living_2020; @rosenberg_decline_2019]. Across all anthropic changes on Earth, habitat conversion is the leading cause of species extinctions  [@maxwell_biodiversity:_2016;@goncalves-souzaHabitatLossExtinction2020]. Protecting habitat must be at the heart of post-2020 conservation policy [@cbd_update_2020;@visconti_protected_2019;@geldmann_essential_2021]. -->

# Methods

## Area of Habitat

Following @brooks_measuring_2019, we created AOH maps for all threatened (Critically Endangered, Endangered, and Vulnerable) terrestrial species in taxonomic groups that have range data and have been comprehensively assessed by the IUCN Red List to at least family level (5,008 species: all amphibians, birds, and mammals, plus select families in the classes Magnoliopsida and Reptilia; Table \@ref(tab:tab-1)). Range polygons were downloaded from the websites of the Red List [@iucn_iucn_2020] and BirdLife International [@birdlife_2020], and refined to AOH by retaining only grid cells that fall within the range polygons, and meet the species' habitat and elevation preferences as defined by the Red List [@iucn_iucn_2020]. All spatial analyses were conducted in Google Earth Engine [@gorelick_google_2017].

```{r tab-1}
# Function to count species in each assessed group
comprehensive_spp <- function(i){
  clade_level <- paste(i["clade"], "Name", sep = "")
  clade_name <- i["scientific"]
  # Get species in this clade
  spp <- all_df[all_df[,clade_level] == clade_name, "species"]
  # Return
  return(length(spp))
}

tab_1 <- 
  read.csv("HMEI-GDrive/data/assessed_final_2022-01-13.csv", 
           stringsAsFactors = FALSE) %>% 
  filter(class %in% unique(all_df$className)) %>% 
  dplyr::select(-included)
tab_1$count <- unlist(apply(tab_1, MARGIN = 1, comprehensive_spp))
class <- tab_1[tab_1$clade == "class",]
family <- tab_1[tab_1$clade == "family" &
                  tab_1$scientific %in% unique(all_df$familyName),]
order <- tab_1[tab_1$clade == "order" & tab_1$scientific == "crocodylia" ,]
tab_1 <-
  rbind(family, class, order) %>% 
    filter(count > 0) %>% 
  mutate(clade = factor(clade,
                        levels = c("class", "order", "family"))) %>% 
  arrange(clade, class) %>% 
  rename("English name" = common,
         "Scientific name" = scientific,
         "Taxonomic level" = clade,
         "Taxonomic class" = class,
         "Total threatened species" = count)
tab_1 <- as.data.frame(apply(tab_1, MARGIN = 2, capitalise))

# Format table
kable(tab_1, "simple", booktabs = TRUE, align = "l",
      caption = "Taxa comprehensively assessed on the IUCN Red List and included in our analyses.")
```

Following Red List assessment protocol [@iucn_guidelines_2019], we excluded habitats that are defined by the Red List as either 'marginal' in their suitability for a given species, or as 'passage', which reflects seasonality of use (e.g. stopover habitat for migratory birds). For migratory species we considered only the smaller of either the breeding or non-breeding range, in keeping with the Red List protocol [@iucn_guidelines_2019]. We did, however, include habitat types classified has having 'unknown' suitability or seasonality, because this situation likely affects many threatened species that are too rare or poorly known for more refined classifications to be possible.

To identify cells with suitable habitat, we used the 1 km-resolution IUCN habitat types map for the year 2015 [@jung_global_2020]. The Red List categorizes habitats at increasingly fine levels of resolution. Thus, level 1 habitat categories refer to the broad categories of Forest, Savanna, Shrubland, Grassland, Wetlands (inland), Rocky Areas, Deserts and Artificial habitats, each of which contains finer level 2 categories, e.g., 'Forest - Boreal' or 'Grassland - Tundra'. Where a species' habitat preference is given only to level 1 by IUCN, we conservatively assumed that all nested level 2 categories are suitable for that species. To identify cells with suitable elevation, we used the EarthEnv digital elevation model [@robinson_earthenv-dem90:_2014]. All maps were resampled to 2 km, following @iucn_guidelines_2019, and projected to a Mollweide projection to ensure accuracy of area calculations.

To quantify the extent to which AOH is a limiting factor in decreasing a species' risk of extinction, we compared AOH to the fixed thresholds under criterion B2 used by the Red List to assign species to the Red List categories. The Red List uses five criteria for this classification, and species are assigned to the highest risk category for which they qualify under any of the criteria [@iucn_guidelines_2019]. For the three 'threatened' Red List categories of Vulnerable, Endangered and Critically Endangered, the AOH thresholds are, respectively, < 2000 km^2^; < 500 km^2^; and < 10 km^2^. We categorized species as likely to be conservation reliant if their AOH falls below any of the three thresholds, as all three imply a significant risk of extinction based solely on habitat amount; the higher the Red List category, the greater that risk.

Red List categorization under criterion B also requires that two of three additional subcriteria are met. These are: (1) the habitat is severely fragmented _or_ restricted to &le; 10 locations (Vulnerable), &le; 5 locations (Endangered), or 1 location (Critically Endangered); (2) there is a continuing decline in habitat area or quality, or the number of locations or number of mature individuals; and (3) there are extreme fluctuations in habitat area or quality, or in the number of locations or number of mature individuals. To accurately incorporate these subcriteria into our analyses would have required us to quantify habitat fragmentation and temporal trends in population size and habitat quality and quantity for all of the species [@santini_applying_2019]. Unfortunately, quantification of these characteristics cannot yet be reliably automated for all threatened species in our dataset. We therefore chose to follow a precautionary approach, based on habitat area alone. In other words, a species is assumed to be at or approaching the level of risk associated with each Red List category if its habitat area falls below the area thresholds defined under criterion B2. This is acceptable only because we are explicitly not attempting to validate or automate Red List classifications. Rather, we are exploring the extent to which habitat area limits species' capacities to recover. 

We classify species as 'No Exit Species' when their extant habitat area falls below the threshold for categorization as Vulnerable (< 2000 km^2^) under Red List criterion B2. We know _a priori_ that all No Exit Species are at high risk of extinction because we assessed only species that are classified as threatened (Critically Endangered, Endangered, or Vulnerable) on the Red List. No Exit Species are constrained to remain threatened indefinitely because of their limited habitat area. No matter what other actions are taken to address threats and improve outcomes for such species, they are condemned to remain threatened unless the habitat-associated extinction risk can be mitigated through habitat restoration or creation.

To give a hypothetical example, if a species is classified as Endangered and has 300 km^2^ of extant habitat, it cannot be downlisted to Vulnerable until its habitat area is increased above the threshold for Endangered (i.e. > 500 km^2^). In this case, at least 200 km^2^ (500 - 300 km^2^) of additional habitat would be required before the species could qualify for downlisting. To quantify the minimum additional habitat area that is required by each No Exit Species, we calculated the difference between each species' current AOH and the nearest threshold value that must be exceeded to permit the species to be downlisted to the next lowest threat category (henceforth: downlisting threshold). In other words, after identifying the lowest threshold below which each species' AOH falls, we calculated the difference between its current AOH value and that threshold. Where the AOH falls just beneath the threshold, relatively little additional habitat must be restored to permit the species to be downlisted (assuming, of course, that the restored area matched the habitat requirements of the species in question and that the species can colonize the new habitat). However, where AOH falls well below the downlisting threshold, relatively large gains in habitat area are needed to permit downlisting, and, indeed, the of AOH of such species may instead be closer to the threshold beneath which, under criterion B2, it could qualify for uplisting to the next highest threat category (henceforth: uplisting threshold). 

## Statistical analyses

Red List thresholds for classification as 'threatened' under criterion B2 do not increase linearly as extinction risk decreases. For a species to move from Critically Endangered to Endangered, its AOH must increase above 10 km^2^, which at most represents an increase in habitat area of 10 km^2^. Moving from Endangered to Vulnerable can require up to 490 km^2^ of additional habitat (500 - 10 = 490 km^2^), and from Vulnerable to Near Threatened up to 1500 km^2^ of additional habitat (2000 - 500 km^2^). We refer to this quantity as the 'absolute distance to downlisting threshold', with a maximum value of either 10 km ^2^, 490 km ^2^ or 1500 km ^2^, depending on which downlisting threshold is the nearest. The minimum value, in contrast, is 0 km^2^, which represents species with an AOH precisely equal to the value of the nearest downlisting threshold (i.e. a species with AOH of exactly 10 km^2^, 500 km^2^, or 2000 km^2^). Thus, all No Exit Species fall into one of three groups depending on the downlisting threshold that they are nearest to; all have AOH < 2000 km^2^, and an absolute distance to downlisting threshold that ranges between 0-10 km^2^, 0-490 km ^2^ or 0-1500 km^2^. 

The absolute distance to downlisting threshold is informative because it represents the actual amount of additional habitat that is required for each threatened species to be able to be downlisted. However, it is also useful to consider a relative measure of distance to downlisting threshold, because this enables comparison across the different threshold groups. Thus, within each group, we identified species whose AOH falls in the upper quantile of absolute distance to downlisting threshold. These we termed 'Extreme No Exit Species', to describe those species whose AOH is furthest from the nearest downlisting threshold. We then calculated the 'relative distance to downlisting threshold' as the absolute distance divided by the maximum possible distance for each threshold (i.e. 10, 490 and 1500 km^2^). As an example, an Endangered species with 1250 km^2^ AOH has a larger absolute distance to the downlisting threshold for Vulnerable (2000 - 1250 = 750 km^2^) than does a Critically Endangered species with 5 km^2^ AOH, which requires only 5 km^2^ more habitat to be capable of being downlisted to Endangered (10 - 5 = 5 km^2^); however, both species have a relative distance to downlisting threshold of 0.5 (absolute distance / maximum distance: 750 / 1500 = 0.5 and 5 / 10 = 0.5). Thus, relative distance to downlisting threshold is a standardized measure that accounts for the fact that the threshold values do not increase linearly. To explore geographic patterns in the distance to downlisting threshold, we determined the countries and zoogeographic realms in which each species occurs, based on the overlap between species' range polygons and maps of global administrative areas [@gadm_2018] and zoogeographic realms [@holt_update_2013].

To compare the relative distance to downlisting threshold for all No Exit Species, we first plotted the frequency distribution of values using a Gaussian kernel density estimate [@wickham_ggplot2_2016]. A density curve is similar to a histogram, and used to display the distribution of continuous data based on the probability of observing data at each location along the x axis. The frequency distribution for relative distance to downlisting threshold is indicative of the extent to which species' AOH values are clustered around the downlisting threshold vs. the uplisting threshold. If the former, most species require relatively little additional habitat to permit them to be downlisted. If the latter, most species are as far from the downlisting threshold as possible. Indeed, in the latter case, species may instead be closer to the uplisting threshold, where relatively little loss of habitat would potentially cause species to be uplisted. 

Lastly, we used a generalized linear model (GLM) with a binomial distribution, using the function `glm` in the `lme4` R  package [@bates_fitting_2015], to test whether relative distance to downlisting threshold varied by taxonomic class, zoogeographic realm and absolute AOH. To avoid double-counting species in this analysis, we limited our focus to species that occur in a single realm. Model inference was based on Likelihood Ratio Tests, dropping each variable in turn and comparing the reduced model to the full model [@zuur_mixed_2009]. All statistical analyses were conducted in R version 4.1.2 [@r_core_team_2021].

# Results

Of the `r nrow(all_df)` threatened species with AOH data, `r nrow(nes_df)` species(`r signif(100*nrow(nes_df)/nrow(all_df),2)%`) are classed as No Exit Species: species that have a total habitat area of < 2000 km^2^ (Fig. \@ref(fig:fig-1)). The remaining `r nrow(all_df) - nrow(nes_df)` threatened species with AOH > 2000 km^2^ are, therefore, not considered to be limited by available habitat quantity, and instead they are likely to be listed as threatened because of their small or declining population size.

Most No Exit Species are amphibians (`r length(which(nes_df$className == "amphibia"))`; `r signif(length(which(nes_df$className == "amphibia")) / nrow(nes_df) * 100,2)`%), followed by birds (`r length(which(nes_df$className == "aves"))`; `r signif(length(which(nes_df$className == "aves")) / nrow(nes_df) * 100,2)`%), mammals (`r length(which(nes_df$className == "mammalia"))`; `r signif(length(which(nes_df$className == "mammalia")) / nrow(nes_df) * 100,2)`%), plants in assessed families of Magnoliopsida (`r length(which(nes_df$className == "magnoliopsida"))`; `r signif(length(which(nes_df$className == "magnoliopsida")) / nrow(nes_df) * 100,2)`%), and, finally, species in assessed families of reptiles (`r length(which(nes_df$className == "reptilia"))`; `r signif(length(which(nes_df$className == "reptilia")) / nrow(nes_df) * 100,2)`%). Based on the Red List thresholds for listing under criterion B2, `r length(which(nes_df$ten_lowest))` species in our dataset have a total habitat area that falls below the 10 km^2^ threshold for Critically Endangered, `r length(which(nes_df$five_lowest))` species have more than 10 km^2^ of habitat but less than 500 km^2^ (Endangered), and `r length(which(nes_df$two_lowest))` species have between 500-2000 km^2^ of habitat (Vulnerable). 

Extreme No Exit Species were those species whose existing habitat area was furthest (top 25%) from the nearest downlisting threshold. These `r sum(nes_df$diff_top25)` threatened species, whose recovery is most severely limited by habitat area, comprise `r sum(nes_df$diff_top25[nes_df$className=="amphibia"])` amphibians, `r sum(nes_df$diff_top25[nes_df$className=="aves"])` birds, `r sum(nes_df$diff_top25[nes_df$className=="mammalia"])` mammals, `r sum(nes_df$diff_top25[nes_df$className=="reptilia"])` of the assessed reptiles, and `r sum(nes_df$diff_top25[nes_df$className=="magnoliopsida"])` of the assessed plants in the class Magnoliopsida.

```{r fig-1, fig.cap = "The total number of threatened species classified as 'No Exit Species' (NES), within each taxonomic class and AOH range. AOH ranges correspond to the threshold values used by the Red List to classify species under, criterion B2, as Critically Endangered (a), Endangered (b), or Vulnerable (c).",fig.width = 16/2.54, fig.height = 10/2.54}
nes <- nes_df %>% 
  mutate(Category = str_replace(Category, "Critically Endangered", 
                                paste("(a) AOH < 10 km","\U00B2",sep= ""))) %>% 
  mutate(Category = str_replace(Category, "Endangered", 
                                paste("(b) 10 < AOH < 500 km","\U00B2",sep= ""))) %>% 
  mutate(Category = str_replace(Category, "Vulnerable", 
                                paste("(c) 500 < AOH < 2000 km","\U00B2",sep= "")))

# relevel Categories
nes$Category <- 
  factor(nes$Category, 
         levels = c(paste("(a) AOH < 10 km","\U00B2",sep= ""), 
                    paste("(b) 10 < AOH < 500 km","\U00B2",sep= ""), 
                    paste("(c) 500 < AOH < 2000 km","\U00B2",sep= "")))
nes$className <- 
  factor(nes$className, 
         levels = unique(nes$className),
         labels = capitalize(unique(nes$className)))

# Code to add pics to labels
classes <- levels(nes$className)
labels <- paste(
  "<img src=", 
  "'",
  file.path("HMEI-GDrive/resources/phylopic", paste(tolower(classes), 'png', sep = ".")),
  "' ",
  "width='10' />", 
  classes,
  sep = ""
)
names(labels) <- classes

max_val <- 834
ggplot(nes, aes(x=className, fill=className)) + 
    geom_bar(alpha = 0.7) + 
    facet_wrap(~Category, labeller = label_wrap_gen(width = 25, multi_line = TRUE)) +
    theme_bw() + theme(panel.grid = element_blank()) +
    labs(fill="Class") +
    ylab("Count") + 
    scale_fill_brewer(palette="Set1") +
    scale_y_continuous(breaks = seq(0, max_val, by = 100), limits=c(0, max_val)) +
    theme(axis.text.x = element_markdown(color = "black",
                                         angle = 45, 
                                         hjust = 1, vjust = 1,
                                         margin = unit(c(-5, 0, 0, 0), "pt")),
          axis.title.x = element_blank(),
          legend.position = "none") +
    scale_x_discrete(labels = labels)
```

Considering the difference between species' AOH and the nearest downlisting threshold highlights those threatened species that are particularly near or far from being able to be downlisted to the next lowest threat category. In other words, the extent to which a species' potential for future downlisting is contingent on habitat restoration or creation. For No Exit Species, there are three possible ranges of AOH values, bounded by the different Red List thresholds of criterion B2: AOH < 10 km^2^; 10 < AOH &le; 500 km^2^; and 500 < AOH &le; 2000 km^2^. In each range and across different taxonomic classes, the AOH of most species falls closer to the threshold for uplisting (maximum difference) than the threshold for downlisting (minimum difference, i.e. 0 km^2^). This can be seen in the left skewed distribution of differences between AOH and the nearest downlisting threshold (Fig. \@ref(fig:fig-2)). 

We found no indication that species' distance to the nearest downlisting threshold differs between taxonomic classes (Deviance = `r signif(class_results[2,4],3)`, DF = `r class_results[2,3]`, `r siglevel(class_results[2,5])`) or biogeographic realms (Deviance = `r signif(realm_results[2,4],3)`, DF = `r realm_results[2,3]`, `r siglevel(realm_results[2,5])`). Species' with greater absolute AOH are relatively closer to the threshold for downlisting, and thus require relatively less additional habitat area to permit future downlisting (Deviance = `r signif(aoh_results[2,4],3)`, DF = `r aoh_results[2,3]`, `r siglevel(aoh_results[2,5])`), although this value may still be large in absolute terms.

```{r fig-2, fig.cap = "Distribution of the difference between species' absolute AOH and the nearest threshold for downlisting. Curves are shaded by taxonomic class are derived from a Gaussian kernel density estimate. This is an alternative to a histogram that is used to display the distribution of continuous data, based on the probability of observing data at each location along the x axis.",fig.width = 16/2.54, fig.height = 10/2.54}
ggplot(nes_df, aes(x=Difference, fill=className)) +
  geom_density(alpha = 0.3, colour = "transparent") +
  facet_wrap(~ Category, scales = "free", 
             labeller = label_wrap_gen(width = 15, multi_line = TRUE)) + 
  theme_bw() + theme(panel.grid = element_blank()) +
  labs(fill="Class") + 
  ylab("Density") +
  xlab(paste("Difference between AOH and downlisting threshold (km","\U00B2",")",sep= "")) + 
  theme(axis.text.x = element_text(angle = 45)) + 
  scale_fill_brewer(palette="Set1") 
```

By mapping total No Exit Species by country (Fig. 3), we find that a few countries harbor a disproportionate share of all species whose recovery on the Red List is limited by habitat area. Across all taxa, the three countries with the most No Exit Species are Colombia, Mexico, and Madagascar. With the addition of Indonesia, these countries also contain the highest number of Extreme No Exit Species - the species with AOH furthest, in relative terms, from the nearest threshold for downlisting. Comparing among zoogeographic realms, the three realms with the most No Exit Species are Neotropical (489 species), Oriental (439 species), and Panamanian (431 species). These are also the three realms with the most Extreme No Exit Species.

Breaking down our geographic results by taxonomic class, No Exit amphibians are largely found in Mexico, Ecuador, Colombia and Madagascar, while No Exit birds are in tropical South America (Peru, Colombia, Brazil) and Indonesia, and mammals in Madagascar and Indonesia. Not all threatened magnoliopsid and reptile groups were able to be considered in our analyses, owing to missing data. China contains the majority of assessed No Exit plants in the class Magnoliopsida, while No Exit reptiles included in our analyses are largely found in Madagascar.

(ref:cap-3) The total number of threatened species within each country classified as (a) 'No Exit Species' (NES) and (b) 'Extreme No Exit Species' (ENES), which are those species in the upper quartile of differences between AOH and the nearest threshold for downlisting. Countries are shaded from low count (yellow) to high (red).

```{r fig-3, fig.cap = "(ref:cap-3)"}
map <- 
  readPNG("HMEI-GDrive/figs/Final Maps/combi_country_grid.png") %>% 
  rasterGrob(interpolate = TRUE)
ggplot() +
  theme_void() +
  scale_x_continuous(expand = c(0,0)) +
  scale_y_continuous(expand = c(0,0)) +
  annotation_custom(map,
                    xmin = -Inf, xmax = Inf,
                    ymin = -Inf, ymax = Inf)
```

# Discussion

Recent advances in quantifying species' habitat area, facilitated by the availability of global habitat maps and data on species' habitat preferences, allowed us to quantify for the first time the number of threatened, terrestrial species whose recovery from endangerment is fundamentally limited by habitat area. We identified `r nrow(nes_df)` species whose remaining habitat is below the threshold amount used by the Red List to categorize species as threatened (i.e. Vulnerable, Endangered, Critically Endangered) [@iucn_guidelines_2019]. We refer to these as "No Exit Species", because they cannot exit 'threatened' status without an increase in the amount of habitat available to them. Such species either require active management to maintain existing populations, or the removal of restrictions to recovery through habitat restoration. Intrinsically, therefore, No Exit Species fall within the definition of "conservation reliant" [@scott_conservation-reliant_2010; @goble_conservation-reliant_2012; @scott_recovery_2005].

Alarmingly, the `r nrow(nes_df)` No Exit Species identified here account for over 60% of the `r nrow(all_df)` threatened species for which we had AOH data. Most are amphibians, followed by birds and mammals and species in select magnoliopsid and reptile families. Many No Exit Species are recognized as Evolutionarily Distinct and Globally Endangered (EDGE) species [@zoologicalsocietyoflondonEDGE], with 40 amphibian, 27 bird, 30 mammal, and 4 reptile No Exit Species included in the top 100 EDGE species for their taxa. Examples include Archey's frog (_Leiopelma archeyi_), Barrio's frog (_Insuetophrynus acarpicus_), and the kagu (_Rhynochetos jubatus_). These species are both highly threatened and highly distinctive in evolutionary terms, such that the loss of these species would have a disproportionate impact on phylogenetic diversity. Yet other No Exit Species are charismatic flagships and serve as useful umbrella species, including the Iberian Lynx (_Lynx pardinus_), the Tapanuli Orangutan (_Pongo tapanuliensis_), and the Indian Rhinoceros (_Rhinoceros unicornis_).

For most No Exit Species, the absolute amount of their remaining habitat puts them at high risk of crossing the uplisting threshold into a higher threat category on the Red List (Fig. \@ref(fig:fig-2)). Indeed, most No Exit Species are in a precarious state: they are closer to the threshold for uplisting than for downlisting. Given the extent of conversion of natural ecosystems predicted for the remainder of this century [@jenkinsProspectsBiodiversity2003], the future for No Exit Species seems increasingly tenuous. The "30 x 30" goal to protect 30% of the world's land and coastal seas by 2030 [@30by30_2021], adopted by countries such as the UK and USA  [@bidenExecutiveOrder140082021; @50CountriesAnnounce], could prevent some No Exit Species from being uplisted to a higher threat category on the basis of habitat loss. At a regional and national scale, priority sites for protection have been identified in the form of Key Biodiversity Areas, designated according to their importance for the global persistence of biodiversity [@iucn_global_2016]. Given their criteria for designation, it is likely that KBAs encompass a high number of No Exit Species. However, habitat protection per se cannot rescue No Exit Species from endangerment; they will require habitat restoration. 

<!-- Designation requires that one of five criteria are met, including sites that provide habitat for many threatened species and sites that represent the majority of the geographic range or population of particular threatened species. Thus, it is likely that KBAs encompass a high number of No Exit Species. -->

The total amount of habitat needed to permit all of the No Exit Species that we identified to be downlisted is `r prettyNum(round(sum(nes_df$Difference[nes_df$Category == "Critically Endangered"]),0),big.mark = ",")` km^2^ for species with < 10 km^2^ habitat, `r prettyNum(round(sum(nes_df$Difference[nes_df$Category == "Endangered"]),0),big.mark = ",")` km^2^ for species with between 10 and 500 km^2^ habitat, and `r prettyNum(round(sum(nes_df$Difference[nes_df$Category == "Vulnerable"]),0),big.mark = ",")` for species with between 500 and 2,000 km^2^ habitat. Undoubtedly there will be overlap between habitat requirements of different species, such that restoring habitat in one area benefits multiple species. However, given the global distribution of No Exit Species (Fig. \@ref(fig:fig-3)), there remains a lot of ground to cover (restore, actually) to ensure that habitat area is no longer a barrier to species' recovery. Initially, we suggest that the focus be on restoring habitat for the highest risk species - those with total habitat area of < 10 km2 - as well as the 'easy wins', which are the Critically Endangered and Endangered species with only a relatively small amount of additional habitat required to enable downlisting. Countries with a large number of No Exit Species requiring only a small increase in habitat area to qualify for downlisting could achieve relatively large gains in habitat for many species simultaneously, where their habitat overlaps. Such countries include Colombia, with over 60 species close to the threshold for downlisting, as well as Mexico, Madagascar, and Indonesia (40-50 species), and China, Peru, India, and Tanzania (~ 30 species). Notably, these are many of the same countries that also have the highest numbers of Extreme No Exit Species (Fig. \@ref(fig:fig-3)) - that is, the species that are most in need of habitat protection to prevent uplisting. Thus, habitat restoration could be coupled with habitat protection in a few key countries, with disproportionate benefit for the conservation of global terrestrial biodiversity. 

Habitat restoration as a conservation strategy is gaining traction around the world and is supported by various initiatives such as the Bonn Challenge and the UN Decade on Ecosystem Restoration. The Initiative 20x20 aimed to protect and restore land in Latin America and the Caribbean in support of the Bonn Challenge and the New York Declaration on Forests. Colombia pledged to restore 1 million ha of degraded land by 2020, Mexico nearly 8.5 million ha of degraded land by 2020, and Peru 3.2 million hectares of degraded land [@Initiative20x20]. Post-2020, there is opportunity for these countries to continue habitat restoration with a concentrated effort on areas where No Exit Species would benefit most. Of the other countries identified as priority regions for No Exit Species, Madagascar has pledged 4,000,000 hectares to restoration efforts, Tanzania 5,200,000 hectares, and India 21,000,000 hectares to the Bonn Challenge by 2030 [@BonnChallenge2016]. China and Indonesia have yet to pledge any habitat to the Bonn Challenge. However, the Chinese government is at the forefront of restoration efforts, including the Natural Forest Conservation Program (NFCP), the Conversion of Cropland to Forest Program (CCFP) as part of the NFCP with considerable success in reforesting 28.2 million hectares by 2014 [@zhangChinaConversionCropland2017]. 

It is, however, important to note that different habitats have different restoration needs, and there is substantial variation in the difficulty of restoring different habitat types [@haleEvaluatingWhereHow2019]. Creating old-growth forests for spotted owls, for example, may take hundreds of years, whereas the habitat of an early successional species could feasibly be restored in a few years. Additionally, the difficulty of restoring habitat to a former state hinges on the extent of damage, with more damaged and industrialized habitats often more difficult to restore. As ecosystem restoration is a relatively young science, there remains a need for more research on how to most effectively restore different habitat types. 

As this study identifies No Exit Species using the AOH available to individual species, we note several limitations to our approach. The first is that we consider only habitat area and not other habitat characteristics that could further restrict the true amount of usable habitat, such as habitat quality and fragmentation [@santini_applying_2019]. Similarly, while population size and trends are influenced by habitat availability, many species' populations are also depressed by additional threats that we do not consider here, including invasive species and overexploitation, for example [@sodhiCausesConsequencesSpecies2009a]. Thus, the list of No Exit Species that we present here represents the species for which habitat availability is likely to be a limiting factor in their recovery, but hardly the only limiting factor. Restoration of habitat would remove this barrier, and thus permit - but not guarantee - recovery of the many species that are threatened by habitat loss. Finally, our identification of No Exit Species was determined by comparing species' habitat area to the thresholds used by the Red List to classify species into categories of extinction risk, but some have argued these thresholds are inevitably not equally meaningful across taxa [@cardosoInvertebrates2012; @martinAreIUCNStandard2009]. Nonetheless, although imperfect, the Red List represents a heavily reviewed, widely accepted, and standardized protocol for measuring progress towards biodiversity targets [@keithScientificFoundationsIUCN2013].

Habitat loss and degradation will continue into the foreseeable future [@jenkinsProspectsBiodiversity2003], presenting ever greater challenges to the species that already have too little habitat. Here, as we approach the 2022 UN Biodiversity Conference and the post-2020 era of biodiversity conservation, we identify the species and regions most in need of habitat protection. Our analysis of which species are closest and furthest from the threshold for downlisting identifies, respectively, cases where relatively small gains in habitat area carry disproportionate benefit, versus cases for which major restoration efforts will be needed to facilitate downlisting. In the latter case, identifying the species for which habitat availability is a major barrier to long-term recovery is vital for conservation organizations to effectively manage their limited resources. As the major driver of species extinctions and with no obvious end in sight, insatiable habitat conversion must be met with equal and opposite commitment to habitat protection and restoration. 

## Data availability 

The datasets used to support the findings of this study are available for download by request from their respective providers. Species assessments can be accessed from the IUCN Red List of Threatened Species website at https://www.iucnredlist.org, and range maps can be requested at https://www.iucnredlist.org/resources/spatial-data-download. Protected Area maps can be requested from the World Database on Protected Areas, found at https://www.unep-wcmc.org/resources-and-data/analysis/main/wdpa. 

## Code availability

Custom Python code to calculate species Area of Habitat in Google Earth Engine will be published at a later date, and is available on request from R.A.S.

## Acknowledgements

Funding for this research was provided by the High Meadows Environmental Institute and High Meadows Foundation.

## Author contributions

R.A.S. and D.S.W. conceived the study. R.A.S. collated the data. S.R. carried out the analyses and wrote the first draft of the manuscript, with substantial contribution from R.A.S. and D.S.W. 

## Competing interests

The authors declare no competing interests.

# References
