---
title: "No exit: the conservation reliance of species with too little habitat"
output:
    bookdown::word_document2:
        reference_docx: ms_general.docx
        number_sections: false
    bookdown::pdf_document2:
        toc: no
        latex_engine: xelatex
        fig_caption: yes
        keep_tex: yes
link-citations: yes
csl: conservation-letters.csl
biblio-style: apalike
bibliography: [refs.bib, funky-refs.bib, pckgs.bib]
fontsize: 12pt
geometry: margin=2.5cm
mainfont: Calibri
lang: en-GB
papersize: a4
documentclass: article
header-includes:
- \usepackage{hyperref}
- \usepackage{fancyhdr}
- \pagestyle{fancy}
- \fancyhf{}
- \fancyfoot[CO,CE]{\thepage}
- \usepackage{setspace}
- \doublespacing
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = "../")
knitr::opts_chunk$set(echo = FALSE,message = FALSE,error = FALSE, warning = FALSE,
                      dpi = 300, fig.path = "output/main/", fig.align = "center", 
                      fig.width = 16.6/2.54, fig.height = 15/2.54)
```

_______________________________________________________________________________________

Sophia Richter^1^ , Rebecca A. Senior^2^*, and David S. Wilcove^1,2^

^1^Department of Ecology and Evolutionary Biology, Princeton University, Princeton, NJ, USA.

^2^Princeton School of Public and International Affairs, Princeton University, Princeton, NJ, USA.

*__Corresponding author:__ rebecca.a.senior@gmail.com (R.A. Senior)

__ORCID iDs:__ orcid.org/0000-0002-8208-736X (R.A. Senior)
  
__Authorship:__ R.A.S. and D.S.W. conceived the study. R.A.S. collated the data. S.R. carried out the analyses and wrote the first draft of the manuscript, with substantial contribution from R.A.S. and D.S.W. 

__Funding:__ High Meadows Environmental Institute and High Meadows Foundation.

__Key words:__ conservation; extinction; IUCN Red List; conservation planning; biodiversity; threatened species

__Article type:__ Perspective

__Number of words:__ 

__Number of References:__ 

__Number of Figures:__ ; __Number of Tables:__ 0 

Conservation Letters formatting guidelines:

Perspectives provide a forum for authors to discuss recent developments or ideas with broad and significant policy relevance to conservation. They should be forward looking and written with a view to informing non-specialist readers (thus should be presented using straightforward prose and avoiding jargon). Perspectives may be up to 3000 words in length and should contain no more than 5 figures and/or tables, and fewer than 30 references.

# Abstract


__/200 words__

# Introduction

Biodiversity is the foundation of ecosystem health, underpinning an array of ecosystem services [@IPBES] and conferring resilience to anthropogenic change through taxonomic, functional and  genetic diversity [@oliver_biodiversity_2015]. However, the world is currently experiencing alarming rates of biodiversity loss [@wwf_living_2020; @rosenberg_decline_2019]. Across all anthropic changes on Earth, habitat conversion is the leading cause of species extinctions  [@maxwell_biodiversity:_2016;@goncalves-souzaHabitatLossExtinction2020]. The underlying driver of habitat conversion is agricultural expansion [@wwf_living_2020; @williams_proactive_2020; @gibbs_tropical_2010], which is expected to continue to pose a major threat to biodiversity. As demand for food increases, driven by population growth and eating habits, the agricultural sector is expected to grow and develop in tandem. Improvements in efficiency, such as increasing yields per area, decreasing waste, and shifting away from meat-based diets in some cultures [@tilman_global_2011; @tilman_future_2017; @williams_proactive_2020], could help to abate habitat loss for some threatened species, but many will face further conversion of habitat - particularly in the tropics [@gibbs_tropical_2010]. Protecting habitat must be at the heart of post-2020 conservation policy [@cbd_update_2020;@visconti_protected_2019;@geldmann_essential_2021].

<!-- P2 = link between habitat area and extinction risk; this should briefly touch on environmental stochasticity (environmental perturbations more likely to effect entire population), demographic stochasticity (sex ratio imbalance, inbreeding, loss of genetic diversity -> impaired potential for local adaptation), and edge effects (smaller habitat means more edge relative to interior, which causes additional problems)  -->

<!-- Excuse the shameless self promotion below... -->

Habitat area and fragmentation are directly related to extinction risk. Indeed, the International Union for the Conservation of Nature (IUCN) include habitat area as one of the five listing criteria for the Red List, according to which, assessed species are placed into categories of extinction risk ranging from Least Concern and Near Threatened, to the 'threatened' categories of: 'Vulnerable', 'Endangered', and 'Critically Endangered' [@mace_quantification_2008; @iucn_iucn_2020]. Ultimately, populations confined to smaller areas are more susceptible to stochasticity and edge effects. Most environmental stressors are localized, such that a population with a large area of habitat benefits from the averaging effect of environmental effects that are not over the entire range, while populations with a small habitat are at a much higher risk of extinction because environmental changes are likely to affect the entire population [@lacyConsideringThreatsViability2000]. Species with small habitat area are also more likely to be affected by demographic stochasticity, including a sex ratio imbalance, inbreeding, and loss of genetic diversity [@CITE]. Compounding stochastic effects are the stresses associated with the relatively large amount of edge in small habitats, such as higher temperatures and easier access for poachers [@ewers_fragmentation_2013;@pfeifer_creation_2017]. The isolation of small, fragmented habitat can also make it difficult for species to disperse elsewhere [@crooks_global_2011;@haddad_habitat_2015], which is increasingly problematic in a warming world where species are on the move towards the poles and uphill [@mcguire_achieving_2016;@senior_global_2019]. 

Many species in the wild can only maintain a self-sustaining population with ongoing species-specific management interventions. They are referred to as "Conservation Reliant Species" [@scott_conservation-reliant_2010; @goble_conservation-reliant_2012; @scott_recovery_2005]. The extreme vulnerability to stochastic and edge effects for species with too little habitat inevitably renders them more dependent on conservation interventions to maintain viable populations. Thus, we can think of insufficient habitat area as imposing a form of conservation reliance. Regardless of the other conservation interventions that we might implement to maintain the status quo - such as invasive species control or regulation of trade - we cannot hope to recover habitat-limited species unless their remaining habitat is robustly protected, alongside positive gains in habitat area. We know that such recovery is indeed humanity's hope, as we, under Goal 15 of the Sustainable Development Goals (SDGs), seek to "...halt and reverse land degradation and halt biodiversity loss" [@TransformingOurWorld2015]. Habitat protection falls directly in line with the proposed "30 x 30" goal to protect 30% of the world’s land and coastal seas by 2030 [@30by30_2021], which has recently been adopted by countries such as the UK and USA [@CITE], and is anticipated to be brought into law in X [@CITE]. Simultaneously, the undoing of previous loss and degradation sits within the remit of large-scale restoration initiatives, such as the Bonn Challenge, which aims to restore 350 million hectares of degraded and deforested landscapes by 350 million hectares by 2030 [@BonnChallenge2016]. 

Not only do we see, empirically, that habitat availability is a major barrier to species' recovery [@CITE], but in practice, a species cannot be downlisted to a lower threat category on the Red List while its 'geographic range' still qualifies it for the higher category [@iucn_iucn_2020]. Under criterion B, species are classed as Vulnerable, Endangered, or Critically Endangered, if the total 'Area of Occupancy' (hereafter: AOO; calculated as the sum of occupied cells within a 2 x 2 km grid)[@iucn_guidelines_2019] falls below the thresholds of 2000, 500 and 10km^2^, respectively. AOO represents a subset of the more commonly used habitat area metric 'Extent of Occurrence' (hereafter: 'EOO'), defined as the area of the minimum polygon that can be drawn to encompass all known, projected or inferred occurrences of a species [@iucn_guidelines_2019]. While EOO is less accurate because it may contain several sites that are ecologically unsuitable for a species, AOO is difficult to measure. However, recent advances in satellite data and resulting habitat maps [@jung_global_2020], coupled with extensive data on species' habitat preferences [@iucn_iucn_2020;@birdlife_2020], now make it possible to refine species' EOO to the areas within a species' range that have an appropriate habitat type and elevation [@brooks_measuring_2019]. The refined map is referred to as 'Area of Habitat' (hereafter: AOH), and as it is functionally equivalent to AOO, the same AOO thresholds can be used to estimate species' minimum Red List category [@santini_applying_2019]. Thus, we have unprecedented capacity to quantify how much habitat threatened species have, and the extent to which habitat area limits species' recovery potential. 

By comparing species' AOH to the Red List thresholds, we identify two related and important facets of conservation reliance for habitat-limited species, based on habitat area alone. The first is how close a species is to being uplisted on the Red List to a higher threat category. The second is how far a species is from being downlisted to a lower threat category. We refer to all species with AOH falling below any of the Red List thresholds as 'No Exit Species' (hereafter: NES), because they can never exit their threatened status without concerted efforts to protect and restore habitat. Here, global conservation policy has a key role to play. In the case of NES close to the uplisting threshold, it is essential to protect what little habitat remains, which can be achieved through strategic expansion of the Protected Area (PA) network under 30 x 30 [@30by30_2021]. For all NES, there is a clear need to reverse historic habitat loss, which can only be achieved through controversial translocation of species to new habitats outside of their current range [@CITE], or through restoration efforts like the Bonn Challenge [@BonnChallenge2016]. It is also imperative that we identify NES in order to pragmatically allocate limited conservation resources. Put simply, conservation is not cheap [@weiseFinancialCostsLarge2014;@mccarthy_financial_2012]. Restoration efforts should ideally be targeted towards the species that would benefit most. More broadly, conservation stakeholders must consider whether other conservation interventions are worthwhile as a means to maintain the status quo when gains in habitat area cannot be achieved, or whether such efforts are a drain on limited resources for species with no hope of recovery.

# Methods

## Area of Habitat

To identify species whose recovery is limited by habitat area (NES), we calculated Area of Habitat (AOH) following @brooks_measuring_2019. In brief, AOH maps were produced for all threatened, terrestrial species in taxonomic groups that have range maps and have been comprehensively assessed by the IUCN Red List to at least family level (all amphibians, birds and mammals, plus select families in the classes Magnoliopsida and Reptilia). Range maps were downloaded from the websites of the Red List [@iucn_iucn_2020] and BirdLife International [@birdlife_2020], and refined to AOH by retaining only grid cells that fall within the range polygons, and meet the species' habitat and elevation preferences as defined by the Red List [@iucn_iucn_2020]. All spatial analyses were conducted in Google Earth Engine[@gorelick_google_2017].

Following Red List assessment protocol [@iucn_guidelines_2019], we exclude habitats that are defined by the Red List as 'marginal' in their suitability, or 'passage' in their seasonality of use. As we are interested in the minimum habitat area needed to sustain a healthy global population, for migratory species we consider only the smaller of either the breeding or non-breeding range, in keeping with the Red List protocol [@iucn_guidelines_2019]. We do, however, retain habitat types listed with 'unknown' suitability or seasonality, because this likely affects many threatened species that are too rare or poorly known for such classification to be possible.

To identify cells with suitable habitat, we used the 1 km-resolution IUCN habitat types map for the year 2015 [@jung_global_2020]. Level 1 habitat categories refer to the broad categories of Forest, Savanna, Shrubland, Grassland, Wetlands (inland), Rocky Areas, Deserts and Artificial habitats, each of which contains finer level 2 categories, e.g., 'Forest - Boreal'. Where a species' habitat preference is given only to level 1, we conservatively assume that all nested level 2 categories are also suitable. To identify cells with suitable elevation, we use the EarthEnv digital elevation model [@robinson_earthenv-dem90:_2014]. All maps were resampled to 2 km, following @iucn_guidelines_2019, projected to a Mollweide projection to ensure accuracy of area calculations. 

To quantify the extent to which AOH is a limiting factor in decreasing a species' risk of extinction, we compare AOH to the fixed thresholds used by the Red List to assign species to the Red List categories. The Red List use five criteria for this classification, and species are assigned to the highest risk category for which they qualify under any of the criteria [@iucn_guidelines_2019]. Here, we are specifically referring to criterion B, under which species are classified based on their geographic range. The area of a species' range can be measured based on Extent of Occurrence (EOO) or Area of Occupancy (AOO). AOH is analagous to AOO and can be used in the same way, using the same thresholds [@brooks_measuring_2019]. 

For the three 'threatened' Red List categories of Vulnerable, Endangered and Critically Endangered, the AOH thresholds are, respectively: < 2000 km^; < 500 km^2; and, < 10 km^2. We categorise species as 'No Exit' if their AOH falls below any of the three thresholds. We further require that species are at least 1 km^2 from that threshold, based on the assumption that 1 km^2 is a trivial difference that can be easily gained or lost, and is within the expected error of the spatial data. Species that are less than 1 km^2 from the threshold are considered to have met the threshold. We note that when the Red List use these thresholds for classification, they additionally require that two out of three subcriteria are met. However, several additional metrics would be needed to assess these subcriteria (habitat fragmentation, and temporal trends in population size or habitat quantity or quality) [@santini_applying_2019]. Such classification is beyond the scope of this study, as we are interested not in automating Red List classification, but rather in the extent to which habitat quantity limits our ability to improve species' prospects. Here, when a species is classed as 'No Exit', we are referring to situations where no matter what other actions we might take to address threats and improve outcomes for threatened species, their restricted AOH means that NES will always be limited in the extent to which extinction risk can be mitigated.

In downstream analyses, we first summarize the taxonomy and biogeography of NES. Next, we explore the magnitude of species' reliance on habitat by calculating the difference between species' AOH and the threshold for downlisting. This difference is the amount of habitat that a species would need to gain to have any chance of downlisting to a lower category of extinction risk on the Red List, regardless of what other conservation actions we might take. We summarize both the distribution of these differences from the downlisting threshold, as well as the upper quantile of species within each habitat category that are furthest from the threshold - the 'Extreme No Exit' species (ENES). ENES are the species with the most additional habitat required to safeguard them from extinction, and thus the species most likely to require major and concerted effort to recover. We determined the countries and realms in which each species occurs by overlaying species' extant range maps with maps of global administrative areas[@gadm_2018] and zoogeographic realms[@holt_update_2013].  

## Statistical analyses

All statistical analyses were conducted in R version 4.0.3 [@r_core_team_2020]. The degree to which species are constrained from future downlisting by current AOH was captured by the relative distance of AOH from the downlisting threshold. This 'proximity to threshold' measure was calculated as the absolute difference between a species' AOH and the threshold for downlisting to the next lowest Red List category, divided by the maximum possible difference. Thus, proximity to threshold is a standardised measure that accounts for the fact that thresholds between categories do not increase linearly. As an example, a species with 1250 km^2^ AOH has a larger absolute difference to the downlisting threshold (2000 - 1250 = 750 km^2^) than a species with 5 km^2^ AOH (10 - 5 = 5 km^2), but both have a relative proximity to threshold of 0.5. We used a generalized linear model (GLM) with a binomial distribution, using the function `glm` in the `lme4` R  package [@bates_fitting_2015], to test whether proximity to threshold varied by taxonomic class, biogeography and absolute AOH. Model inference was based on Likelihood Ratio Tests, dropping each variable in tern and comparing the reduced model to the full model [@zuur_mixed_2009].

# Results

2,488 species of the 3,869 species with AOH data have habitat areas below the 2000 km^2 threshold. 
Of the 2,488 species that qualify as NES, 1389 are amphibians, 527 birds, 461 mammals, 59 magnoliopsida, and 52 reptiles. Furthermore, 335 fall below the 10 km^2 threshold, 720 fall between the 10 km^2 and 500 km^2 thresholds, and 1433 have between 500 km^2 and 2000 km^2 AOH. Figure 1 shows the threshold and class of the identified NES.

```{r class-and-threshold-graph}
library(readr)
library(ggplot2)
library(stringr)
library(dplyr)

nes <- read_csv('nes.csv')

nes <- nes %>% mutate(Category = str_replace(Category, "Critically Endangered", "AOH < 10 km^2")) %>% 
  mutate(Category = str_replace(Category, "Endangered", "AOH < 500 km^2")) %>% 
  mutate(Category = str_replace(Category, "Vulnerable", "AOH < 2000 km^2"))

# relevel Categories
nes$Category <- factor(nes$Category, levels = c("AOH < 10 km^2", "AOH < 500 km^2", "AOH < 2000 km^2"))

ggplot(nes, aes(x=className, fill=className)) + geom_bar() + facet_wrap(~Category, labeller = label_wrap_gen(width = 12, multi_line = TRUE)) + ggtitle("Class and Threshold of NES") + theme(axis.text.x = element_blank()) + labs(fill="Class") + xlab(" ") + ylab("Count")
```

The three realms with the most NES are Neotropical (1101 species), Afrotropical (571 species), and Indomalayan (452 species). 

Across class and threshold met, there is a right skew towards a large difference to the threshold, in particular for those between 10 km^2 and 500 km^2 (Figure 2).

```{r difference-histogram}
ggplot(nes, aes(x=Difference, fill=className)) + geom_histogram(alpha=0.7) + 
  ggtitle("Distribution of Differences to Next Threshold \n Across Class and Category") +
    facet_wrap(~ Category, scales = "free", labeller = label_wrap_gen(width = 12, multi_line = TRUE)) + 
  labs(fill="Class") + 
  ylab("Count") + xlab("Difference (km^2)") + 
  theme(axis.text.x = element_text(angle = 45))
```

The 25% of NES with the largest difference to the threshold across each threshold category was also subsetted. Of the 623 species in the top 25% differences, 358 are amphibians, 124 birds, 113 mammals, 15 reptiles, and 13 magnoliopsida. The three realms with the most top 25% differences are Neotropical, Afrotropical, and Indomalayan. 

The concentration of NES and top 25% mammals, birds, and amphibians have been mapped by realm and country (Figure 3).

[maps]

The GLM ran on proximity to threshold varying by taxonomic class, biogeography and absolute AOH found that only absolute AOH had a significant effect on proximity to threshold, while taxonomic class and realm had no significance. 

# Discussion


## Data availability 

The datasets used to support the findings of this study are available for download by request from their respective providers. Species assessments can be accessed from the IUCN Red List of Threatened Species website at https://www.iucnredlist.org, and range maps can be requested at https://www.iucnredlist.org/resources/spatial-data-download. Protected area maps can be requested from the World Database on Protected Areas, found at https://www.unep-wcmc.org/resources-and-data/analysis/main/wdpa. 

## Code availability

Custom Python code to calculate species Area of Habitat in Google Earth Engine will be published at a later date, and is available on request from R.A.S.

## Acknowledgements

Funding for this research was provided by the High Meadows Environmental Institute and High Meadows Foundation.

## Author contributions

R.A.S. and D.S.W. conceived the study. R.A.S. collated the data. S.R. carried out the analyses and wrote the first draft of the manuscript, with substantial contribution from R.A.S. and D.S.W. 

## Competing interests

The authors declare no competing interests.

# References
