---
title: "EOO Edits to no_exit_iucn.Rmd" 
author: "Sophia V. Richter"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    code_folding: show
    highlight: default
    keep_md: yes
    theme: journal
    toc: yes
    toc_float:
      collapsed: no
      smooth_scroll: yes
      toc_depth: 3
editor_options: 
  chunk_output_type: console
---

```{=html}
<style>
pre code, pre, code {
  white-space: pre !important;
  overflow-x: scroll !important;
  word-break: keep-all !important;
  word-wrap: initial !important;
}
</style>
```
```{r setup, include=FALSE}
# For width of code chunks and scroll bar 
options(width=250)

knitr::opts_chunk$set(eval = TRUE, 
                      echo = TRUE, 
                      include = TRUE,
                      warning = FALSE,
                      collapse = FALSE,
                      message = FALSE,
                      engine = "R", # Chunks will always have R code, unless noted
                      error = TRUE,
                      fig.align = "center") 

```

```{r load-packages}
# Data wrangling
library(dplyr)
library(tidyr)
library(readr)
# Figures
library(GGally)
library(ggplot2)
# Stats
library(lme4)
```

# Load data

```{r load-sorted-data}
merge_dat <- readRDS("/workdir/sr2352/HMEI-internship/data/HMEI-GDrive/data/sorted/merge_dat.Rds")
realm_cats <- sort(unique(unlist(strsplit(unique(merge_dat$Realms), "[;]"))))
```

```{r eoo-thresholds}
# Modify data frame
eoo_all_df <- merge_dat %>% 
  # remove species with 0 or NA AOH
  filter(!(is.na(areaAOH)), areaAOH > 0) %>%
  mutate(
    # convert AOH to km^2
    AOHkm = areaAOH/1000000,
    # Identify NES
    EOO_NES = AOHkm < 20000,
    # add column for whether AOH falls below 100 km (more than 10 km away from threshold)
    one_hundred = AOHkm < 100, # changed from ten
    # add column for whether AOH falls below more than 1 km below 5000 km threshold
    five_thousand = AOHkm < 5000, # changed from five_hundred
    # add column for whether AOH falls below more than 1 km below 20000 km threshold
    twenty_thousand = AOHkm < 20000, # changed from two_thousand
    # column if one_hundred is lowest threshold
    first_lowest = one_hundred, # was ten_lowest
    # column if 5000 is lowest threshold 
    second_lowest = xor(one_hundred, five_thousand), #was five_lowest
    # column if 20000 is lowest threshold
    third_lowest = xor(five_thousand, twenty_thousand), # was two_lowest
    # record what category should be
    EOO_Category = ifelse(first_lowest == TRUE, "Critically Endangered", 
                      ifelse(second_lowest == TRUE, "Endangered", 
                             ifelse(third_lowest == TRUE, "Vulnerable", 
                                    "Near Threatened"))),
    # Calculate difference to next threshold
    Difference = ifelse(first_lowest == TRUE, 100 - AOHkm, 
                        ifelse(second_lowest == TRUE, 5000 - AOHkm, 
                               ifelse(third_lowest == TRUE, 20000 - AOHkm, 
                                      NA)))) %>% 
  # Calculate Difference between Category and redlistCategory
  mutate(num_Category = ifelse(EOO_Category == "Critically Endangered", 1, 
                               ifelse(EOO_Category == "Endangered", 2, 
                                      ifelse(EOO_Category == "Vulnerable", 3, 
                                             ifelse(EOO_Category == "Near Threatened", 4, 
                                                    NA)))),
         num_redlist = ifelse(redlistCategory == "Critically Endangered", 1, 
                              ifelse(redlistCategory == "Endangered", 2, 
                                     ifelse(redlistCategory == "Vulnerable", 3, 
                                            ifelse(redlistCategory == "Near Threatened", 4, 
                                                   NA)))),
         cat_red_change = num_Category - num_redlist,
         # Standardised difference
         difference_prop = ifelse(EOO_Category == "Critically Endangered", Difference / 100, 
                                  ifelse(EOO_Category == "Endangered", Difference / 4900, 
                                         ifelse(EOO_Category == "Vulnerable", Difference / 15000, 
                                                NA))),
         # Flip the scale so bigger numbers = closer to threshold for downlisting
         proximity = 1 - difference_prop)

# Calculate quantiles
quants <-
    eoo_all_df %>% 
    group_by(EOO_Category) %>%
    dplyr::summarise(diff25 = quantile(Difference, probs = 0.25, na.rm = TRUE), # lower
              diff50 = quantile(Difference, probs = 0.5, na.rm = TRUE), # median
              diff75 = quantile(Difference, probs = 0.75, na.rm = TRUE)) # upper

# Add to df
eoo_all_df <- 
    left_join(eoo_all_df, quants, by = "EOO_Category") %>% 
    mutate(
        # Identify species furthest from downlisting threshold
        diff_top25 = Difference >= diff75,
        # Identify species closest to downlisting threshold
        diff_bottom25 = Difference <= diff25
        )

eoo_nes_df <-
  eoo_all_df %>% 
  # remove Differences that are NA
    # (i.e. species that would not be classed as threatened according to AOH)
  filter(!(is.na(Difference)))

# Sanity check
unique(eoo_nes_df$EOO_NES)
```


```{r realm-long-df}
# Long dataframe
eoo_realms_long_all <- 
 eoo_all_df %>%  
  dplyr::select(-Realms) %>% 
  pivot_longer(cols= all_of(realm_cats), 
               names_to='Realms', values_to='value') %>% 
  dplyr::filter(value == TRUE) %>% 
  dplyr::select(-value)

eoo_realms_long_nes <-
  eoo_realms_long_all %>% 
  dplyr::filter(EOO_NES)
```

There are `r nrow(realms_long_nes)` species with a habitat that qualifies a species for a threatened category in each realm. 

Negative value means Habitat Category is uplisted compared to redlistCategory, 0 means they match, positive means Habitat Category is downlisted compared to redlistCategory.

```{r summarize-category-change}
cat_change <- eoo_all_df %>% 
  group_by(cat_red_change) %>% 
  summarise(Sum = n())

cat_change

nochange <- cat_change %>% filter(cat_red_change == 0) %>% pull(Sum)

decrease <- cat_change %>% filter(cat_red_change < 0) %>% summarize(sum = sum(Sum)) %>% pull(sum)

increase <- cat_change %>% filter(cat_red_change > 0) %>% summarize(sum = sum(Sum)) %>% pull(sum)

change_dat <- 
  eoo_all_df %>% 
  filter(cat_red_change == -2) %>% 
  select(species, className, redlistCategory)
```

`r nochange` species (`r nochange / nrow(eoo_all_df) * 100` %) do not change category. `r decrease` species (`r decrease / nrow(eoo_all_df) * 100` %) have less AOH than their current red list category (- value). `r increase` species (`r increase / nrow(eoo_all_df) * 100` %) species have more AOH than their red list category (+ value).


# No Exit Species by EOO (EOO_NES)

Classes of NES
```{r nes-classes}
# Table
eoo_nes_df %>% 
  group_by(className) %>% 
  summarise(Count = n()) %>% 
  arrange(-Count)
```

Category of NES
```{r nes-category}
#table
eoo_nes_df %>% 
  group_by(EOO_Category) %>% 
  summarize(Count = n()) %>% 
  arrange(-Count)

# graphs
ggplot(eoo_nes_df, aes(x=className, fill=className)) + 
  geom_bar() + 
  facet_wrap(~EOO_Category) + 
  ggtitle("Class and Threshold of NES by EOO") + 
  theme(axis.text.x = element_blank()) + 
  labs(fill="Class") + xlab(" ") + ylab("Count")
```

Realm of Counts of EOO_NES

Realms of each Conservation Reliant Species
```{r nes-realm}
eoo_nes_df %>% 
  group_by(Realms) %>% 
  summarise(Count = n()) %>% 
  arrange(-Count)
```

Number of CRS in each realm
```{r realm-nes-count}
eoo_realms_long_nes %>% 
  group_by(Realms) %>% 
  summarize(Count = n()) %>% 
  arrange(-Count)

ggplot(eoo_realms_long_nes, aes(x=Realms)) + 
  geom_bar() + 
  theme(axis.text.x = element_text(angle = 90)) + 
  ggtitle("Number of CRS in each Realm")

ggplot(eoo_realms_long_nes, aes(x=Realms, fill=Realms)) + 
  geom_bar() + 
  facet_wrap(~EOO_Category) + 
  theme(axis.text.x = element_blank()) + 
  ggtitle("Category and Realm of CRS")

```

# Difference Distribution

Class
```{r diff-density-class}
ggplot(eoo_nes_df, aes(x=Difference, fill=className, alpha = 0.01)) + 
  geom_histogram() + 
  ggtitle("Distribution of Differences to Next Threshold Across Class")

ggplot(eoo_nes_df, aes(x=Difference, fill=className, alpha=0.01)) + 
  geom_density() + 
  ggtitle("Density of Differences to Next Threshold Across Class")

ggplot(eoo_nes_df, aes(x=Difference)) + 
  geom_histogram() + 
  facet_wrap(~className) + 
  ggtitle("Distribution of Differences to Next Threshold Across Class")

ggplot(eoo_nes_df, aes(x=Difference)) + 
  geom_density() + 
  facet_wrap(~className) + 
  ggtitle("Density of Differences to Next Threshold Across Class")
```

Histogram of Differences by Class
```{r diff-hist-class}
ggplot(eoo_nes_df, aes(x=Difference, fill=className)) + 
  geom_histogram(alpha=0.7) + 
  ggtitle("Distribution of Differences to Next Threshold \n Across Class and Category") +
  facet_wrap(~ EOO_Category, scales = "free", 
             labeller = label_wrap_gen(width = 2, multi_line = TRUE)) + 
  labs(fill="Class") + 
  ylab("Count") + 
  xlab("Difference (km^2)")
```

Category
```{r diff-hist-category}
ggplot(eoo_nes_df, aes(x=Difference, fill=EOO_Category, alpha=0.5)) + 
  geom_histogram() + 
  ggtitle("Distribution of Differences to Next Threshold Across Category")
```

Realm
```{r diff-hist-realm}
ggplot(eoo_realms_long_nes, aes(x=Difference, fill=Realms)) + 
  geom_histogram() + 
  ggtitle("Distribution of Differences Across Realm")

ggplot(eoo_realms_long_nes, aes(x=Difference)) + 
  geom_histogram() + 
  facet_wrap(~Realms)

```


# Top 25% Differences

```{r top-25-diff, include=F}
# Filter to highest risk species, furthest from downlisting threshold
# (top 25% of distance from threshold)
diff_top25 <- filter(eoo_nes_df, diff_top25)

# Filter to easy win species, closest to downlisting threshold
# (bottom 25% of distance from threshold)
diff_bottom25 <- filter(eoo_nes_df, diff_bottom25)
```

### Class
```{r top-diff-class}
diff_top25 %>% 
  group_by(className) %>% 
  summarise(Count = n()) %>% 
  arrange(-Count)

ggplot(diff_top25, aes(x=className)) + 
  geom_bar() + 
  ggtitle("Distribution of Top 25% (of each category) AOH Differences - AKA Furthest from Downlisting")

ggplot(diff_top25, aes(x=className, fill=className)) + 
  geom_bar() + 
  facet_wrap(~EOO_Category) + 
  ggtitle("Distribution of Top 25% AOH Differences (furthest from downlisting)") + 
  theme(axis.text.x = element_text(angle = 90))
```

### Category
```{r top-diff-category, include=F}
ggplot(diff_top25, aes(x=EOO_Category, fill=EOO_Category)) + 
  geom_bar() + 
  ggtitle("Distribution of Top 25% (of each category) AOH Differences (furthest from downlisting)") + 
  theme(axis.text.x = element_text(angle = 90))
```

### Realm
```{r top-diff-realm, include=F}
crit_realms <- eoo_realms_long_nes %>% 
    filter(EOO_Category == 'Critically Endangered') %>% 
    filter(Difference >= quants$diff75[quants$EOO_Category == "Critically Endangered"])

endan_realms <- eoo_realms_long_nes %>% 
    filter(EOO_Category == 'Endangered') %>% 
    filter(Difference >= quants$diff75[quants$EOO_Category == "Endangered"])

vul_realms <- eoo_realms_long_nes %>% 
    filter(EOO_Category == 'Vulnerable') %>% 
    filter(Difference >= quants$diff75[quants$EOO_Category == "Vulnerable"])

# near_realms = realms %>% filter(Category == 'Near Threatened') 

# also all combined
top_realm = full_join(crit_realms, endan_realms)
top_realm = full_join(top_realm, vul_realms)
#top_realm = full_join(bottom_realm, near_realms)

ggplot(top_realm, aes(x=Realms)) + 
  geom_bar() + 
  ggtitle("Distribution of Top 25% (of each category) AOH Differences Across Realm") + 
  theme(axis.text.x = element_text(angle = 90))

ggplot(top_realm, aes(x=Realms, fill=Realms)) + 
  geom_bar() + 
  facet_wrap(~EOO_Category) + 
  ggtitle("Distribution of Top 25% AOH Differences Across Realm") + 
  theme(axis.text.x = element_text(angle = 90))
```

# GLM

= column for species name + class + whether NES (0 or 1)
glm(NES ~ class, family = "binomial")


Realm NES

```{r interaction-test}
# I think we should focus on endemic species here, to avoid double-counting
glm_full_nes <- 
  glm(EOO_NES ~ className * Realms + className * EOO_Category + Realms * EOO_Category, 
      data = filter(eoo_realms_long_all, realm_endemic),
      family = "binomial", maxit=100)

# Drop & test each interaction in turn
glm_dropInter1_nes <- update(glm_full_nes, ~.- className:Realms)
anova(glm_dropInter1_nes, glm_full_nes, test = "LRT") # this interaction is not significant

glm_dropInter2_nes <- update(glm_full_nes, ~.- className:EOO_Category)
anova(glm_dropInter2_nes, glm_full_nes, test = "LRT") # this interaction is not significant

glm_dropInter3_nes <- update(glm_full_nes, ~.- Realms:EOO_Category)
anova(glm_dropInter3_nes, glm_full_nes, test = "LRT") # this interaction is not significant

#-------------------------------------------------------------------------------
# All had equally small effect (p-value of 1, small deviance). dropping realm:Category has highest DF
# so make this the new reference model and repeat with remaining interactions
# ------------------------------------------------------------------------------

glm_dropInter4_nes <- update(glm_dropInter3_nes, ~.- className:Realms)
anova(glm_dropInter4_nes, glm_dropInter3_nes, test = "LRT") # this interaction is not significant

glm_dropInter5_nes <- update(glm_dropInter3_nes, ~.- className:EOO_Category)
anova(glm_dropInter5_nes, glm_dropInter3_nes, test = "LRT") # this interaction is not significant

#-------------------------------------------------------------------------------
# Dropping either interaction term had same deviance and p-value effect. Dropping className:Category had higher DF
# so make this the new reference model and repeat with remaining interaction
# ------------------------------------------------------------------------------

glm_dropInter6_nes <- update(glm_dropInter5_nes, ~.- className:Realms)
anova(glm_dropInter6_nes, glm_dropInter5_nes, test = "LRT") # this interaction is not significant

#-------------------------------------------------------------------------------
# Drop ALL interactions - HOORAY
# now test significance of individual variables
# ------------------------------------------------------------------------------

# Test significance of class
glm_dropClass_nes <- update(glm_dropInter6_nes, ~.- className)
anova(glm_dropClass_nes, glm_dropInter6_nes, test = "LRT") # class is not significant

# Test significance of realm
glm_dropRealm_nes <- update(glm_dropInter6_nes, ~.- Realms)
anova(glm_dropRealm_nes, glm_dropInter6_nes, test = "LRT") # realm is not significant

# Test significance of category
glm_dropCat_nes <- update(glm_dropInter6_nes, ~.- EOO_Category)
anova(glm_dropCat_nes, glm_dropInter6_nes, test = "LRT") # category is significant

# Compare AIC as well, for good measure
AIC(glm_dropInter6_nes,glm_dropClass_nes,glm_dropRealm_nes,glm_dropCat_nes)
# all quite similar, except for dropping Category which produces a *much* worse model
```

--> I am not sure if we want to add the plot of model-predicted values? I would guess it is very similar (in that it is not helpful) to the last data set. 

### LM on Difference 
- try difference / max difference for these models 
```{r}
nes_difference <- 
  eoo_nes_df %>% 
  filter(
    # Drop NAs
    !(is.na(Difference)),
    # Restrict to species occurring in one Realm only
    realm_endemic
  )
# Make a standard lm
lm_prop <- lm(difference_prop ~ className, 
              data = eoo_nes_df)

# Model validation plots
par(mfrow = c(2,2))
plot(lm_prop)
summary(lm_prop)

# Make a standard lm
lm_prop_cat <- lm(difference_prop ~ EOO_Category, 
                  data = nes_difference)

# Model validation plots
par(mfrow = c(2,2))
plot(lm_prop_cat)
summary(lm_prop_cat)

# Make a standard lm
lm_prop_realm <- lm(difference_prop ~ Realms, 
                    data = eoo_nes_df)

# Model validation plots
par(mfrow = c(2,2))
plot(lm_prop_realm)
summary(lm_prop_realm)

```
- much smaller p-value for category (close to 0) than classname (0.06501)
- very high p-value for realm

```{r nes-proximity}
# Round negative values to zero (these are the ones that are < 1 km from the threshold)
nes_difference$proximity[nes_difference$proximity < 0] <- 0

# Plot the distribution of standardised differences
ggplot(nes_difference, 
       aes(x = proximity, fill = className)) +
  geom_histogram() +
  facet_wrap(~ EOO_Category) + 
    labs(title = "Proximity to Downlisting by Category and Class")

# GLM with Poisson because relative diff is actually a proportion
eoo_glm_diff <- glm(proximity ~ className + Realms + AOHkm, 
                data = nes_difference,
                family = "binomial") # ignore the non-integer warnings

# Model inference
summary(eoo_glm_diff)
anova(eoo_glm_diff, test = "LRT")

# Drop class
eoo_class_results <- anova(update(eoo_glm_diff, ~.- className), eoo_glm_diff, test = "LRT")
# Drop realm
eoo_realm_results <- anova(update(eoo_glm_diff, ~.- Realms), eoo_glm_diff, test = "LRT")
# Drop AOH
eoo_aoh_results <- anova(update(eoo_glm_diff, ~.- AOHkm), eoo_glm_diff, test = "LRT")

### Only absolute AOH is a strong predictor or proximity to threshold

#############################
# Plot model-predicted values
#############################

# Make up some new data
newdat <- expand.grid(
  # All levels of taxonomic class
  className = unique(eoo_glm_diff$data$className),
  # All levels of realm
  Realms = unique(eoo_glm_diff$data$Realms),
  # Sequence of 100 made up values for AOH, bounded by observed data
  AOHkm = seq(min(eoo_glm_diff$data$AOHkm),
              max(eoo_glm_diff$data$AOHkm),
              length.out = 100))

# Use model to predict distance values
predvals <- predict(eoo_glm_diff, newdata = newdat, type = "link", se.fit = TRUE)

# Function to calculate inverse logit
inv.logit <- function(x) exp(x)/(1 + exp(x))

# Function to calculate critical significance value
ci.critical <- 
  function(siglevel) qnorm((100 - siglevel) / 100 / 2, lower.tail = FALSE)
ci_const = ci.critical(95)

# Add vals to newdat dataframe
newdat <- 
  newdat %>% 
  mutate(proximity = predvals$fit,
         se = predvals$se.fit,
         CI_lo = proximity - (ci_const * se),
         CI_hi = proximity + (ci_const * se)) 
# Back transform
newdat[,c("proximity", "CI_lo", "CI_hi")] <- 
  inv.logit(newdat[,c("proximity", "CI_lo", "CI_hi")])

# Plot model-predicted values
ggplot(newdat, 
       aes(x = AOHkm, y = proximity)) +
  # # Add confidence intervals
  # geom_ribbon(aes(ymin = CI_lo,
  #                 ymax = CI_hi,
  #                 fill = realm),
  #             alpha = 0.3) +
  # Add fitted line
  geom_line(aes(colour = Realms)) +
  # Facet by taxonomic class
  facet_grid(~ className) +
  theme_classic()

# Since class and realm aren't relevant, we can also just pick a reference level
# or average across everything, and plot one line only
eoo_newdat_avg <- 
  group_by(newdat, AOHkm) %>% 
  summarise(proximity = median(proximity),
            CI_lo = median(CI_lo),
            CI_hi = median(CI_hi))
ggplot(eoo_newdat_avg, 
       aes(x = AOHkm, y = proximity)) +
  # Add confidence intervals
  geom_ribbon(aes(ymin = CI_lo,
                  ymax = CI_hi),
              alpha = 0.1) +
  # Add fitted line
  geom_line() +
  theme_classic()

# Proximity to threshold increases as absolute AOH increases
```

### Correlation between className and Category
```{r class-category-corr}
ggplot(eoo_nes_df, aes(className, EOO_Category)) + 
  geom_count()

class_cat <- table(eoo_nes_df$className, eoo_nes_df$EOO_Category) 

chisq <- chisq.test(class_cat)
chisq
```
p-value is very small (< 2.2e-16), therefore there is a strong correlation between Category and class


# Spatial Data
Save data frame for spatial analysis

```{r nes-top25-spatial}
nes_top25 <- 
  eoo_all_df %>%
  rename(scientificName = species) %>%
  mutate(
    # Match species name to how it appears in occurrence dataframe
    scientificName = gsub(" ", "_", scientificName)) 
```

```{r nes-top25-prop}
# Summarize NES by realm
realm_nes <-
  nes_top25 %>%
  filter(!(is.na(Realms))) %>% 
  # Count species in each realm
  group_by(Realms, className) %>% 
  summarise(n = n(),
            n_NES = length(which(EOO_NES)),
            n_nonNES = length(which(!(EOO_NES))),
            prop_NES = n_NES / n)
```

```{r save-data}
# Inspect the loaded objects - what might we need in the manuscript?
ls()
save(eoo_all_df,eoo_nes_df,eoo_realms_long_all,eoo_realms_long_nes,
     eoo_glm_diff,eoo_class_results,eoo_realm_results,eoo_aoh_results,eoo_newdat_avg,
     file = "data/HMEI-GDrive/data/sorted/eoo_msdata.Rdata")
```



# EDGE species analysis
load files
```{r load-edge}
edge_amph <- read.csv('data/EDGE_data/EDGE_amphibia.csv')
edge_aves <- read.csv('data/EDGE_data/EDGE_aves.csv')
edge_mamm <- read.csv('data/EDGE_data/EDGE_mammalia.csv')
edge_rept <- read.csv('data/EDGE_data/EDGE_reptilia.csv')

# keep only top 100 in each
edge_amph <- edge_amph %>% filter(EDGE.Rank <= 100)
edge_aves <- edge_aves %>% filter(EDGE.Rank <= 100)
edge_mamm <- edge_mamm %>% filter(EDGE.Rank <= 100)
edge_rept <- edge_rept %>% filter(EDGE.Rank <= 100)
```

compare edge to nes
```{r edge-nes-comp}
amph <- inner_join(eoo_nes_df, edge_amph, by=c('species'='Species'))
aves <- inner_join(eoo_nes_df, edge_aves, by=c('species'='Species'))
mamm <- inner_join(eoo_nes_df, edge_mamm, by=c('species'='Species'))
rept <- inner_join(eoo_nes_df, edge_rept, by=c('species'='Species'))
```

There are `r nrow(amph)` amphibians, `r nrow(aves)` birds, `r nrow(mamm)` mammals, and `r nrow(rept)` reptile NES by EOO that are also EDGE species. 