---
title: "IUCN and Habitat Data Merge"
author: "Sophia Richter"
date: "6/10/2021"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(warning=F)
knitr::opts_chunk$set(message=F)
```

Read Data (code and files obtained from Rebecca)

```{r}
library(dplyr)
library(readr)
library(GGally)
library(ggplot2)

# Get a vector of all the filenames
files <- list.files(path = "aohareas_historic", 
                    # Keep the full file path
                    full.names = TRUE,
                    # Limit to CSV files
                    pattern = "\\.csv$")

# Read each one in with lapply & bind all
areadat <- 
    lapply(files, function(file_i){
        suppressMessages(read_csv(file_i))
    }) %>% 
    # Collapse list into dataframe
    bind_rows()

# test to make sure correct
areadat %>% distinct(species)
```

Merge area data and IUCN data
```{r}
# read iucn data
iucndat <- read_csv('iucn_dat.csv')

merge_dat <- inner_join(areadat, iucndat, by=c('species'='scientificName'))
```


## For each year and each species, identify whether the species' AOH falls below any of the IUCN Red List category thresholds (10, 500 and 2000 km^2) 

```{r}
# look at data
# 8834 year/species have 0 AOH
merge_dat %>% group_by(areaAOH.x) %>% summarize(Sum = n()) %>% arrange(areaAOH.x)

# add column for AOH in km^2
merge_dat_km <- merge_dat %>% mutate(AOHkm = areaAOH.x/1000000) %>% mutate(PAkm = areaAOHinPA.x/1000000)

# add column for whether AOH falls below 10 km
merge_dat_km <- merge_dat_km %>% mutate(ten = AOHkm < 10)

merge_dat_km %>% group_by(ten) %>% summarize(Sum = n()) %>% arrange(-Sum) %>% select(ten, Sum)

# add column for whether AOH falls below 500
merge_dat_km <- merge_dat_km %>% mutate(five_hundred = AOHkm < 500)

merge_dat_km %>% group_by(five_hundred) %>% summarize(Sum = n()) %>% arrange(-Sum) %>% select(five_hundred, Sum)

# add column for whether AOH falls below 2000
merge_dat_km <- merge_dat_km %>% mutate(two_thousand = AOHkm < 2000)

merge_dat_km %>% group_by(two_thousand) %>% summarize(Sum = n()) %>% arrange(-Sum) %>% select(two_thousand, Sum)
```


## Identify the lowest threshold that AOH falls under - this will define what the species' Red List category ought to be

```{r}
# column if ten is lowest threshold
merge_dat_km_thresh <- merge_dat_km %>% mutate(ten_lowest = ten)

# column if 500 is lowest threshold
merge_dat_km_thresh <- merge_dat_km_thresh %>% mutate(five_lowest = xor(ten, five_hundred))

merge_dat_km_thresh %>% select(ten, ten_lowest, five_hundred, five_lowest, two_thousand) 

# column if 2000 is lowest threshold
merge_dat_km_thresh <- merge_dat_km_thresh %>% mutate(two_lowest = xor(five_hundred, two_thousand))
```


## Record what the species' Red List category ought to be, and record whether this is the same as its actual Red List category

```{r}
# record what category should be
merge_dat_cat <- merge_dat_km_thresh %>% mutate(Category = 'Not Threatened') 

merge_dat_cat <- merge_dat_cat %>% mutate(Category = ifelse(ten_lowest == TRUE, "Critically Endangered", ifelse(five_lowest == TRUE, "Endangered", ifelse(two_lowest == TRUE, "Vulnerable", "Near Threatened"))))

# record if category matches IUCN red list data
merge_dat_match <- merge_dat_cat %>% mutate(Match = Category == redlistCategory)

check <- merge_dat_match %>% select(redlistCategory, Category, Match)
check
```


## For all species falling under a threshold, calculate the difference between AOH and the next highest threshold.

```{r}
merge_dat_diff <- merge_dat_match %>% mutate(Difference = ifelse(ten_lowest == TRUE, 10 - AOHkm, ifelse(five_lowest == TRUE, 500 - AOHkm, ifelse(two_lowest == TRUE, 2000 - AOHkm, NA))))

check <- merge_dat_diff %>% select(AOHkm, Difference, ten_lowest, five_lowest, two_lowest)

# there are a lot of species with only a small difference !
check %>% arrange(Difference)

```



# Difference in threshold - AOH Explored

## Explore the values calculated above (i.e. how much additional habitat is needed) to see what their distribution is, whether there are any 'easy wins' where the difference is relatively small, and if there are any patterns in species taxonomy/geography/traits that correlate with the amount of additional habitat required

Distribution of Differences
```{r}
# remove values with NA and extract the last year for each species
merge_diff = merge_dat_diff %>% filter(`system:index` == 23) %>%  filter(Difference != "NA")

ggplot(merge_diff, aes(x=Difference)) + geom_histogram(binwidth=50) + ggtitle("Distribution of Differences (km^2) to Next Threshold Value")

```

A large portion of the differences are near 0, and the majority are below 500 km^2. 


## Easy Wins


### Critically Endangered (< 10 km^2)
```{r}
# there are 701 species that have an AOH less than 10 km^2
crit_diff <- merge_diff %>% filter(Category == "Critically Endangered")

# Distribution
ggplot(crit_diff, aes(x=Difference)) + geom_histogram(binwidth=1) + ggtitle("Distribution of Difference (km^2) to Next Threshold Value \n for Critically Endangered Species")

# Number of species in the lowest 5% of differences and that average
701 * 0.05
crit_diff %>% arrange(Difference) %>% slice(1:35) %>% select(species, Difference)

crit_diff %>% arrange(Difference) %>% slice(1:35) %>% summarize(Mean = mean(Difference), Median = median(Difference))

# Number of species in the lowest 10% of differences and that average
701*0.1

crit_diff %>% arrange(Difference) %>% slice(1:70) %>% select(species, Difference)

crit_diff %>% arrange(Difference) %>% slice(1:70) %>% summarize(Mean = mean(Difference), Median = median(Difference))

# Number of species in the lowest 25% of differences and that average
701*0.25

crit_diff %>% arrange(Difference) %>% slice(1:175) %>% select(species, Difference)

crit_diff %>% arrange(Difference) %>% slice(1:175) %>% summarize(Mean = mean(Difference), Median = median(Difference))

```

The distribution of the species that are Critically Endangered skews right, with a high proportion needing the full 10 km^2 restored. The mean and median of the bottom 5% difference is around 1. For 10% it is around 1.5, and for 25% it is around 4.  


### Endangered

```{r}
# There are 1404 endangered species
endan_diff <- merge_diff %>% filter(Category == "Endangered")

# Distribution
ggplot(endan_diff, aes(x=Difference)) + geom_histogram(binwidth=20) + ggtitle("Distribution of Difference (km^2) to Next Threshold Value \n for Endangered Species")


# Number of species in the lowest 5% of differences and that average
1404 * 0.05

endan_diff %>% arrange(Difference) %>% slice(1:70) %>% select(species, Difference)

endan_diff %>% arrange(Difference) %>% slice(1:70) %>% summarize(Mean = mean(Difference), Median = median(Difference))

# Number of species in the lowest 10% of differences and that average
1404 * 0.1

endan_diff %>% arrange(Difference) %>% slice(1:140) %>% select(species, Difference)

endan_diff %>% arrange(Difference) %>% slice(1:140) %>% summarize(Mean = mean(Difference), Median = median(Difference))

# Number of species in the lowest 25% of differences and that average
1404 * 0.25

endan_diff %>% arrange(Difference) %>% slice(1:351) %>% select(species, Difference)

endan_diff %>% arrange(Difference) %>% slice(1:351) %>% summarize(Mean = mean(Difference), Median = median(Difference))
    
```

The distribution of the species that are Endangered skews right, with a left tail. For the 5% of the species with the lowest Difference, the mean and median is around 29 and 28 (respectively). For 10%, the mean and median is 59 and 60, and for 25% it is 138 and 142. 


### Vulnerable

```{r}
# There are 753 endangered species
vul_diff <- merge_diff %>% filter(Category == "Vulnerable")

# Distribution
ggplot(vul_diff, aes(x=Difference)) + geom_histogram(binwidth=75) + ggtitle("Distribution of Difference (km^2) to Next Threshold Value \n for Vulnerable Species")


# Number of species in the lowest 5% of differences and that average
753 * 0.05

vul_diff %>% arrange(Difference) %>% slice(1:37) %>% select(species, Difference)

vul_diff %>% arrange(Difference) %>% slice(1:37) %>% summarize(Mean = mean(Difference), Median = median(Difference))


# Number of species in the lowest 10% of differences and that average
753 * 0.1

vul_diff %>% arrange(Difference) %>% slice(1:75) %>% select(species, Difference)

vul_diff %>% arrange(Difference) %>% slice(1:75) %>% summarize(Mean = mean(Difference), Median = median(Difference))


# Number of species in the lowest 25% of differences and that average
753 * 0.25

vul_diff %>% arrange(Difference) %>% slice(1:188) %>% select(species, Difference)

vul_diff %>% arrange(Difference) %>% slice(1:188) %>% summarize(Mean = mean(Difference), Median = median(Difference))
```

For Vulnerable species, the distribution in difference is relatively flat and only a slight left tail. The mean and median for the bottom 5% is around 53. For 10%, it is 111, and for 25% the mean and median is 291 and 299. 


## Patterns 
taxonomy/geography/traits

### Taxonomy

order and class across all red list categories
```{r}
ggplot(merge_diff, aes(x=orderName, y=Difference)) + geom_boxplot() + theme(axis.text.x = element_text(angle = 90)) + ggtitle("Difference (km^2) to Next Threshold Value Across Order")

ggplot(merge_diff, aes(x=className, y=Difference)) + geom_boxplot() + theme(axis.text.x = element_text(angle = 90)) + ggtitle("Difference (km^2) to Next Threshold Value Across Class")
```

Separated by Red List Category
```{r}
# Critically Endangered
ggplot(crit_diff, aes(x=orderName, y=Difference)) + geom_violin() + theme(axis.text.x = element_text(angle = 90)) + ggtitle("Difference (km^2) to Next Threshold Value Across Order \n for Critically Endangered Species")

ggplot(crit_diff, aes(x=className, y=Difference)) + geom_violin() + theme(axis.text.x = element_text(angle = 90)) + ggtitle("Difference (km^2) to Next Threshold Value Across Class \n for Critically Endangered Species")

ggplot(crit_diff, aes(x=className, y=Difference)) + geom_boxplot() + theme(axis.text.x = element_text(angle = 90)) + ggtitle("Difference (km^2) to Next Threshold Value Across Class \n for Critically Endangered Species")


# Endangered
ggplot(endan_diff, aes(x=orderName, y=Difference)) + geom_violin() + theme(axis.text.x = element_text(angle = 90)) + ggtitle("Difference (km^2) to Next Threshold Value Across Order \n for Endangered Species")

ggplot(endan_diff, aes(x=className, y=Difference)) + geom_boxplot() + theme(axis.text.x = element_text(angle = 90)) + ggtitle("Difference (km^2) to Next Threshold Value Across Class \n for Endangered Species")


# Vulnerable
ggplot(vul_diff, aes(x=orderName, y=Difference)) + geom_violin() + theme(axis.text.x = element_text(angle = 90)) + ggtitle("Difference (km^2) to Next Threshold Value Across Order \n for Vulnerable Species")

ggplot(vul_diff, aes(x=className, y=Difference)) + geom_boxplot() + theme(axis.text.x = element_text(angle = 90)) + ggtitle("Difference (km^2) to Next Threshold Value Across Class \n for Vulnerable Species")

```

Across each of the red list categories and classes, birds have a consistently lower median (and often lower bound of IQR) for Difference. In addition, amphibians and mammals both have a larger proportion of species with a Difference skewed towards the upper values. 


### Geography

```{r}
ggplot(merge_diff, aes(x=systems, y=Difference)) + geom_boxplot() + theme(axis.text.x = element_text(angle = 90)) + ggtitle("Difference (km^2) to Next Threshold Value for each System")

ggplot(merge_diff, aes(x=realm, y=Difference)) + geom_boxplot() + theme(axis.text.x = element_text(angle = 90)) + ggtitle("Difference (km^2) to Next Threshold Value Across Realms")

```

There is little difference across system, but certain realms have a significantly higher median and IQR Difference (e.g. Indomalayn|Palearctic).


Geography across Red List Category
```{r}
# Critically Endangered
ggplot(crit_diff, aes(x=systems, y=Difference)) + geom_violin() + theme(axis.text.x = element_text(angle = 90)) + ggtitle("Difference (km^2) to Next Threshold Value Across Systems \n for Critically Endangered Species")

ggplot(crit_diff, aes(x=realm, y=Difference)) + geom_boxplot() + theme(axis.text.x = element_text(angle = 90)) + ggtitle("Difference (km^2) to Next Threshold Value Across Realms \n for Critically Endangered Species")


# Endangered
ggplot(endan_diff, aes(x=systems, y=Difference)) + geom_boxplot() + theme(axis.text.x = element_text(angle = 90)) + ggtitle("Difference (km^2) to Next Threshold Value Across Systems \n for Endangered Species")

ggplot(endan_diff, aes(x=realm, y=Difference)) + geom_boxplot() + theme(axis.text.x = element_text(angle = 90)) + ggtitle("Difference (km^2) to Next Threshold Value Across Realms \n for Endangered Species")


# Vulnerable
ggplot(vul_diff, aes(x=systems, y=Difference)) + geom_boxplot() + theme(axis.text.x = element_text(angle = 90)) + ggtitle("Difference (km^2) to Next Threshold Value Across Systems \n for Vulnerable Species")

ggplot(vul_diff, aes(x=realm, y=Difference)) + geom_boxplot() + theme(axis.text.x = element_text(angle = 90)) + ggtitle("Difference (km^2) to Next Threshold Value Across Realms \n for Vulnerable Species")

```

Critically Endangered: Terrestrial is more evenly distributed, whereas inland waters have a large proportion with a 10 km^2 difference. Of the realms,  afrotropical|palearctic, indomalayan|oceanian, indomalayan|palearctic have only a few species in this category. Afrotropical, australasion, and nearctic|neotropical (in particular) have lower medians for difference, and the rest have medians at 10 km^2. 

Endangered: There is no significant difference between systems. Indomalayn|Palearctic has a significantly larger IQR but also the highest median. Afrotropical|Palearctic has a small and low IQR and median. 

Vulnerable: There is no significant difference between systems. Oceanian has a lower median when compared to other realms, whereas nearctic and Nearctic|Neotropical have small IQRs and higher medians. 


### Traits
There was no significant correlation between difference and any of the following traits: body mass, congregatory, elevation, area restricted, fragmentation, population trend, or cat_change_bi. 

Body Mass
```{r}
pair = merge_diff %>% select(BodyMass, Difference)

ggpairs(pair)
```

There is no correlation between body mass and difference. 


Congregatory
```{r}
ggplot(merge_diff, aes(x=Congregatory.value,  y=Difference)) + geom_boxplot() + ggtitle("Difference (km^2) to Next Threshold Value Across \n Congregatory Patterns")

# Numbers
merge_diff %>% group_by(Category, Congregatory.value) %>% summarize(Sum = n())
```

There is not a significant difference among congregatory patterns overall, also there is not enough data to really gather anything significant. 


Elevation (lower and upper limit)
```{r}
elevation = merge_diff %>% select(ElevationLower.limit, ElevationUpper.limit, Difference)

ggpairs(elevation)

```

There is no significant correlation between elevation and difference. Unsurprisingly, there is a strong correlation between lower and upper limit. 


Area Restricted
```{r}
ggplot(merge_diff, aes(x=AreaRestricted.isRestricted, y=Difference)) + geom_boxplot() + ggtitle("Restricted Area v. Difference (km^2) to Next Threshold Value")
```

Between 'Yes' and 'No' for whether the area is restricted, the IQR and median for 'Yes' is a lower Difference than 'No'. However it does not seem to be significant. 


Fragmentation
```{r}
ggplot(merge_diff, aes(x=SevereFragmentation.isFragmented, y=Difference)) + geom_boxplot() + ggtitle("Fragmented v. Difference (km^2) to Next Threshold Value")
```

There does not seem to be a significant correlation. 


Population Trend
```{r}
ggplot(merge_diff, aes(x=populationTrend, y=Difference)) + geom_boxplot() + ggtitle("Population Trend v. Difference (km^2) to Next Threshold Value")
```

No significant difference. 


cat_change_bi
```{r}
ggplot(merge_diff, aes(x=cat_change_bi, y=Difference)) + geom_boxplot() + ggtitle("Red List Category Change v. Difference (km^2) to \n Next Threshold Value")
```

No significance. 