---
title: "Conservation Reliance in IUCN Red List"
author: "Sophia Richter"
date: "6/23/2021"
output:
    html_document: default
    pdf_document: default
---
    
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F)
knitr::opts_chunk$set(warning=F)
knitr::opts_chunk$set(message=F)
```


```{r}
# Libraries and read data set
library(dplyr)
library(readr)
library(GGally)
library(ggplot2)

iucn <- read.csv('iucn_dat.csv')
```


```{r}
# Modify data frame

# remove species with 0 or NA AOH
data <- iucn %>% filter(!(is.na(areaAOH)), areaAOH > 0)

# convert AOH to km^2
data <- data %>% mutate(AOHkm = areaAOH/1000000)


# add column for whether AOH falls below 10 km
data <- data %>% mutate(ten = AOHkm < 10)
# add column for whether AOH falls below 500
data <- data %>% mutate(five_hundred = AOHkm < 500)
# add column for whether AOH falls below 2000
data <- data %>% mutate(two_thousand = AOHkm < 2000)


# column if ten is lowest threshold
data <- data %>% mutate(ten_lowest = ten)
# column if 500 is lowest threshold
data <- data %>% mutate(five_lowest = xor(ten, five_hundred))
# column if 2000 is lowest threshold
data <- data %>% mutate(two_lowest = xor(five_hundred, two_thousand))


# record what category should be
data <- data %>% mutate(Category = ifelse(ten_lowest == TRUE, "Critically Endangered", ifelse(five_lowest == TRUE, "Endangered", ifelse(two_lowest == TRUE, "Vulnerable", "Near Threatened"))))


# Calculate difference to next threshold
data <- data %>% mutate(Difference = ifelse(ten_lowest == TRUE, 10 - AOHkm, ifelse(five_lowest == TRUE, 500 - AOHkm, ifelse(two_lowest == TRUE, 2000 - AOHkm, NA))))



# remove differences less than 1 km^2 and Differences that are NA
test2 <- data %>% filter(Difference > 1)



# Calculate Difference between Category and redlistCategory
test <- test2 %>% mutate(num_Category = ifelse(Category == "Critically Endangered", 1, ifelse(Category == "Endangered", 2, ifelse(Category == "Vulnerable", 3, ifelse(Category == "Near Threatened", 4, NA)))))

test <- test %>% mutate(num_redlist = ifelse(redlistCategory == "Critically Endangered", 1, ifelse(redlistCategory == "Endangered", 2, ifelse(redlistCategory == "Vulnerable", 3, ifelse(redlistCategory == "Near Threatened", 4, NA)))))

test <- test %>% mutate(cat_red_change = num_Category - num_redlist)
```

There are 2612 species with a difference to habitat threshold greater than 1 km^2. 

Negative value means Habitat Category is uplisted compared to redlistCategory, 0 means they match, positive means Habitat Category is downlisted compared to redlistCategory. 
```{r}
test %>% group_by(cat_red_change) %>% summarize(Sum = n())
```
1322 species (50.6%) do not change category. 581 species (22.2%) have less habitat than redlistCategory (- value). 709 species (27.1%) have more habitat than redlistCategory (+ value).


The largest category of species with AOH data is endangered, neotropical amphibians. The only species with AOH data are vulnerable, endangered, or critically endangered. Of all species that are in these three red list categories, the largest category is endangered neotropical amphibans.
```{r}
data %>% group_by(Category, realm, className) %>% summarize(Sum = n()) %>% arrange(-Sum)

data %>% group_by(redlistCategory, realm, className) %>% summarize(Sum = n()) %>% arrange(-Sum)

data %>% distinct(redlistCategory)

iucn %>% filter(redlistCategory == "Vulnerable" | redlistCategory == "Endangered" | redlistCategory == "Critically Endangered") %>% group_by(redlistCategory, realm, className) %>% summarize(Sum = n()) %>% arrange(-Sum)
```


# Conservation Reliant Species

Classes of Conservation Reliant Species
```{r}
# Table
test %>% group_by(className) %>% summarize(Count = n()) %>% arrange(-Count)
```

Category of Conservation Reliant Species
```{r}
#table
test %>% group_by(Category) %>% summarize(Count = n()) %>% arrange(-Count)

# graphs
ggplot(test, aes(x=className, fill=className)) + geom_bar() + facet_wrap(~Category) + ggtitle("Class and Category of CRS") + theme(axis.text.x = element_blank())
```

Realm
```{r, include=F}
# separate species into realms
neo = test %>% filter(realm == "Neotropical" | realm == "Nearctic|Neotropical" | realm == "Neotropical|Oceanian") %>% mutate(realm = "Neotropical")

afro = test %>% filter(realm == "Afrotropical" | realm == "Afrotropical|Palearctic" | realm == "Afrotropical|Indomalayan|Palearctic") %>% mutate(realm = "Afrotropical")

ocean = test %>% filter(realm == "Oceanian" | realm == "Neotropical|Oceanian" | realm == "Indomalayan|Oceanian" | realm == "Australasian|Oceanian" | realm == "Oceanian|Palearctic" | realm == "Indomalayan|Oceanian|Palearctic") %>% mutate(realm = "Oceanian")

indo = test %>% filter(realm == "Indomalayan" | realm == "Indomalayan|Palearctic" | realm == "Afrotropical|Indomalayan|Palearctic" | realm == "Indomalayan|Nearctic|Palearctic" | realm == "Indomalayan|Oceanian" | realm == "Australasian|Indomalayan" | realm == "Indomalayan|Oceanian|Palearctic" | realm == "Australasian|Indomalayan|Palearctic") %>% mutate(realm = "Indomalayan")

austra = test %>% filter(realm == "Australasian" | realm == "Australasian|Oceanian" | realm == "Australasian|Indomalayan" | realm == "Australasian|Indomalayan|Palearctic") %>% mutate(realm = "Australasian")

palea = test %>% filter(realm == "Palearctic" | realm == "Afrotropical|Palearctic" | realm == "Indomalayan|Palearctic" | realm == "Afrotropical|Indomalayan|Palearctic" | realm == "Indomalayan|Nearctic|Palearctic" | realm == "Oceanian|Palearctic" | realm == "Australasian|Indomalayan|Palearctic" | realm == "Nearctic|Palearctic" | realm == "Indomalayan|Oceanian|Palearctic") %>% mutate(realm = "Palearctic")

nea = test %>% filter(realm == "Nearctic" | realm == "Nearctic|Neotropical" | realm == "Indomalayan|Nearctic|Palearctic" | realm == "Nearctic|Palearctic") %>% mutate(realm = "Nearctic")

# combine with duplicates
realms = bind_rows(neo, afro)
realms = bind_rows(realms, ocean)
realms = bind_rows(realms, indo)
realms = bind_rows(realms, austra)
realms = bind_rows(realms, palea)
realms = bind_rows(realms, nea)
```

Realms of each Conservation Reliant Species
```{r}
test %>% group_by(realm) %>% summarize(Count = n()) %>% arrange(-Count)
```

Number of CRS in each realm
```{r}
realms %>% group_by(realm) %>% summarize(Count = n()) %>% arrange(-Count)

ggplot(realms, aes(x=realm)) + geom_bar() + theme(axis.text.x = element_text(angle = 90)) + ggtitle("Realms of CRS")

ggplot(realms, aes(x=realm, fill=realm)) + geom_bar() + facet_wrap(~Category) + theme(axis.text.x = element_blank()) + ggtitle("Category and Realm of CRS")

```


# Difference Distribution

Class
```{r}
ggplot(test, aes(x=Difference, fill=className, alpha=0.01)) + geom_histogram() + ggtitle("Distribution of Differences to Next Threshold Across Class")


ggplot(test, aes(x=Difference, fill=className, alpha=0.01)) + geom_density() + ggtitle("Density of Differences to Next Threshold Across Class")

ggplot(crit_top, aes(x=Difference, fill=className, alpha=0.01)) + geom_density() + ggtitle("Distribution of Differences to Next Threshold Across Class \n for Critically Endangered Species")

ggplot(test, aes(x=Difference)) + geom_histogram() + facet_wrap(~className) + ggtitle("Distribution of Differences to Next Threshold Across Class")

ggplot(test, aes(x=Difference)) + geom_density() + facet_wrap(~className) + ggtitle("Density of Differences to Next Threshold Across Class")
```

Category
```{r}
ggplot(test, aes(x=Difference, fill=Category, alpha=0.5)) + geom_histogram() + ggtitle("Distribution of Differences to Next Threshold Across Category")
```

Realm
```{r}
ggplot(realms, aes(x=Difference, fill=realm)) + geom_histogram() + ggtitle("Distribution of Differences Across Realm")

ggplot(realms, aes(x=Difference)) + geom_histogram() + facet_wrap(~realm)
ggplot(realms, aes(x=Difference)) + geom_density() + facet_wrap(~realm)  + ggtitle("Distribution of Differences Across Realm")

ggplot(realms, aes(x=Difference)) + geom_density() + facet_wrap(~realm)  + ggtitle("Density of Differences Across Realm")
```


# Top 25% Differences

```{r, include=F}
# Top 25% of each category
# NOTE: may need to add Near Threatened if want to include 
crit_top = test %>% filter(Category == 'Critically Endangered') %>% filter(Difference >= 9.231000)

endan_top = test %>% filter(Category == 'Endangered') %>% filter(Difference >= 457.429708)

vul_top = test %>% filter(Category == 'Vulnerable') %>% filter(Difference >= 1263.658855)

# near_top = test %>% filter(Category == 'Near Threatened')

# also all combined
top_diff = full_join(crit_top, endan_top)
top_diff = full_join(top_diff, vul_top)
# bottom_PA = full_join(bottom_PA, near_PA)

# bottom 25% of all Differences
all_top = test %>% filter(Difference >= 485.591354)
```



### Class
```{r}
top_diff %>% group_by(className) %>% summarize(Count = n()) %>% arrange(-Count)

ggplot(top_diff, aes(x=className)) + geom_bar() + ggtitle("Distribution of Top 25% (of each category) AOH Differences")

ggplot(top_diff, aes(x=className, fill=className)) + geom_bar() + facet_wrap(~Category) + ggtitle("Distribution of Top 25% AOH Differences") + theme(axis.text.x = element_text(angle = 90))
```


### Realm
```{r, include=F}
crit_realms = realms %>% filter(Category == 'Critically Endangered') %>% filter(Difference >= 9.231000)

endan_realms = realms %>% filter(Category == 'Endangered') %>% filter(Difference >= 457.821961)

vul_realms = realms %>% filter(Category == 'Vulnerable') %>% filter(Difference >= 1265.166894)

# near_realms = realms %>% filter(Category == 'Near Threatened') 

# also all combined
top_realm = full_join(crit_realms, endan_realms)
top_realm = full_join(top_realm, vul_realms)
#top_realm = full_join(bottom_realm, near_realms)
```


```{r}
ggplot(top_realm, aes(x=realm)) + geom_bar() + ggtitle("Distribution of Top 25% (of each category) AOH Differences Across Realm") + theme(axis.text.x = element_text(angle = 90))

ggplot(top_realm, aes(x=realm, fill=realm)) + geom_bar() + facet_wrap(~Category) + ggtitle("Distribution of Top 25% AOH Differences Across Realm") + theme(axis.text.x = element_text(angle = 90))
```

NOTE: I did proportion analysis and it was definitely not significant. 

### Category
```{r, include=F}
ggplot(top_diff, aes(x=Category, fill=Category)) + geom_bar() + ggtitle("Distribution of Top 25% (of each category) AOH Differences") + theme(axis.text.x = element_text(angle = 90))
```
