---
title: "Conservation Reliance in IUCN Red List"
author: "Sophia Richter"
date: "6/23/2021"
output:
    html_document: default
    pdf_document: default
---
    
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F)
knitr::opts_chunk$set(warning=F)
knitr::opts_chunk$set(message=F)
```


```{r libraries}
# Libraries and read data set
library(dplyr)
library(readr)
library(GGally)
library(ggplot2)

iucn <- read.csv('HMEI-GDrive/data/iucn_dat.csv')
```


```{r sort-data}
# Modify data frame
crs <- iucn %>% 
    # remove species with 0 or NA AOH
    filter(!(is.na(areaAOH)), areaAOH > 0) %>%
    mutate(
        # convert AOH to km^2
        AOHkm = areaAOH/1000000,  
        # add column for whether AOH falls below 10 km
        ten = AOHkm < 10,
        # add column for whether AOH falls below 500
        five_hundred = AOHkm < 500,
        # add column for whether AOH falls below 2000
        two_thousand = AOHkm < 2000,
        # column if ten is lowest threshold
        ten_lowest = ten,
        # column if 500 is lowest threshold
        five_lowest = xor(ten, five_hundred),
        # column if 2000 is lowest threshold
        two_lowest = xor(five_hundred, two_thousand),
        # record what category should be
        Category = ifelse(ten_lowest == TRUE, "Critically Endangered", 
                          ifelse(five_lowest == TRUE, "Endangered", 
                                 ifelse(two_lowest == TRUE, "Vulnerable", "Near Threatened"))),
        # Calculate difference to next threshold
        Difference = ifelse(ten_lowest == TRUE, 10 - AOHkm, 
                            ifelse(five_lowest == TRUE, 500 - AOHkm, 
                                   ifelse(two_lowest == TRUE, 2000 - AOHkm, NA))),
        # Calculate Difference between Category and redlistCategory
        num_Category = ifelse(Category == "Critically Endangered", 1, 
                              ifelse(Category == "Endangered", 2, 
                                     ifelse(Category == "Vulnerable", 3, 
                                            ifelse(Category == "Near Threatened", 4, NA)))),
        num_redlist = ifelse(redlistCategory == "Critically Endangered", 1, 
                             ifelse(redlistCategory == "Endangered", 2, 
                                    ifelse(redlistCategory == "Vulnerable", 3, 
                                           ifelse(redlistCategory == "Near Threatened", 4, NA)))),
        cat_red_change = num_Category - num_redlist) %>% 
    # Select variables of interest
    dplyr::select(scientificName,className, realms,
                  redlistCategory,Category,
                  AOHkm,Difference,
                  num_Category,num_redlist,cat_red_change) %>% 
    # Rename some variables for simplicity
    rename(
        # True Red List category 
        "category_true" = "redlistCategory",
        "category_true_num" = "num_redlist",
        # Red List category based on AOH only
        "category_aoh" = "Category",
        "category_aoh_num" = "num_Category",
        # Biogeographic realm
        realm = realms)
```

There are 2612 species with a difference to habitat threshold greater than 1 km^2. 

Negative value means Habitat Category is uplisted compared to redlistCategory, 0 means they match, positive means Habitat Category is downlisted compared to redlistCategory. 
```{r}
crs %>% 
    group_by(cat_red_change) %>% 
    summarize(Sum = n())

change2 <- 
    crs %>% 
    filter(cat_red_change == -2) %>% 
    select(scientificName, className)
```
1322 species (50.6%) do not change category. 581 species (22.2%) have less habitat than redlistCategory (- value). 709 species (27.1%) have more habitat than redlistCategory (+ value).
<!-- It is notable that so many species should probably be in a higher risk category! -->


The largest category of species with AOH data is endangered, neotropical amphibians. The only species with AOH data are vulnerable, endangered, or critically endangered. Of all species that are in these three red list categories, the largest category is endangered neotropical amphibans.
```{r}
crs %>% 
    group_by(category_aoh, realms, className) %>% 
    summarize(Sum = n()) %>% 
    arrange(-Sum)

crs %>% 
    group_by(category_true, realms, className) %>% 
    summarize(Sum = n()) %>% arrange(-Sum)

crs %>% distinct(category_true)

iucn %>% 
    filter(redlistCategory == "Vulnerable" | 
               redlistCategory == "Endangered" | 
               redlistCategory == "Critically Endangered") %>% 
    group_by(redlistCategory, realms, className) %>% 
    summarize(Sum = n()) %>% 
    arrange(-Sum)
```


# Conservation Reliant Species

Classes of Conservation Reliant Species
```{r}
# Table
crs %>% 
    group_by(className) %>% 
    summarize(Count = n()) %>% 
    arrange(-Count)
```

Category of Conservation Reliant Species
```{r}
#table
crs %>% 
    group_by(category_aoh) %>% 
    summarize(Count = n()) %>% 
    arrange(-Count)

# graphs
ggplot(crs, aes(x=className, fill=className)) + 
    geom_bar() + 
    facet_wrap(~ category_aoh) + 
    ggtitle("Class and Category of CRS") + 
    theme(axis.text.x = element_blank()) + 
    labs(fill="Class") + 
    xlab(" ") + ylab("Count")
```
<!-- Wow - it's pretty mad that there are as many CR birds & mammals as there are. Amphibians not so surprising, since their ranges are naturally much smaller -->

Realm

<!-- The code chunk below is a simpler way to identify which realms a species occurs in -->
```{r realms, include = FALSE, eval = FALSE}
# Figure out what the unique realms are
sort(unique(unlist(strsplit(unique(crs$realm), "[|]"))))

# Define wee function to reformat realms variable
gsubGeo <- function(x){gsub(", ",";",gsub("[]]","",gsub("[[]","",x)))}

# Use 'grepl' to see if string is contained in another string
realms <-
    crs %>% 
    # Get rid of square brackets in realm variable
    mutate(realm = gsubGeo(realm)) %>% 
    mutate(
           Afrotropical = grepl("Afrotropical", realm),
           Australasian = grepl("Australasian", realm),
           Indomalayan = grepl("Indomalayan", realm),
           Nearctic = grepl("Nearctic", realm),
           Neotropical = grepl("Neotropical", realm),
           Oceanian = grepl("Oceanian", realm),
           Palearctic = grepl("Palearctic", realm))
```

Realms of each Conservation Reliant Species
```{r}
crs %>% 
    group_by(realm) %>% 
    summarize(Count = n()) %>% 
    arrange(-Count)
```

Number of CRS in each realm
```{r}
realms %>% 
    group_by(realm) %>% 
    summarize(Count = n()) %>% 
    arrange(-Count)

ggplot(realms, aes(x=realm)) + 
    geom_bar() + 
    theme(axis.text.x = element_text(angle = 90)) + 
    ggtitle("Number of CRS in each Realm")

ggplot(realms, aes(x=realm, fill=realm)) + 
    geom_bar() + 
    facet_wrap(~ category_aoh) + 
    theme(axis.text.x = element_blank()) + 
    ggtitle("Category and Realm of CRS")

```


# Difference Distribution

Class
```{r}
ggplot(crs, aes(x=Difference, fill=className, alpha=0.01)) + 
    geom_histogram() + 
    ggtitle("Distribution of Differences to Next Threshold Across Class")

# I'm adding this one below with free scales so we can see what's going on in
# EN and VU without the distribution being squashed by CR
ggplot(crs, aes(x=Difference, fill=className)) + 
    geom_histogram(alpha=0.7) + 
    ggtitle("Distribution of Differences to Next Threshold \n Across Class and Category") +
    facet_wrap(~ category_aoh, scales = "free", 
               labeller = label_wrap_gen(width = 2, multi_line = TRUE)) + 
    labs(fill="Class") + 
    ylab("Count") + 
    xlab("Difference (km^2)")

ggplot(crs, aes(x=Difference, fill=className, alpha=0.01)) + 
    geom_density() + 
    ggtitle("Density of Differences to Next Threshold Across Class")

ggplot(crs, aes(x=Difference)) + 
    geom_histogram() + 
    facet_wrap(~className) + 
    ggtitle("Distribution of Differences to Next Threshold Across Class")

ggplot(crs, aes(x=Difference)) + 
    geom_density() + 
    facet_wrap(~className) + 
    ggtitle("Density of Differences to Next Threshold Across Class")
```
 <!-- Cool - looks like most things are close to the threshold -->

Category
```{r}
ggplot(crs, aes(x=Difference, fill=category_aoh), alpha=0.5) + 
    geom_histogram() + 
    ggtitle("Distribution of Differences to Next Threshold Across Category")
```

Realm
```{r}
ggplot(realms, aes(x=Difference, fill=realm)) + 
    geom_histogram() + 
    ggtitle("Distribution of Differences Across Realm")

ggplot(realms, aes(x=Difference)) + 
    geom_histogram() + 
    facet_wrap(~realm)

ggplot(realms, aes(x=Difference)) + 
    geom_density() + 
    facet_wrap(~realm)  + 
    ggtitle("Distribution of Differences Across Realm")

ggplot(realms, aes(x=Difference)) + 
    geom_density() + 
    facet_wrap(~realm)  + 
    ggtitle("Density of Differences Across Realm")
```


# Top 25% Differences

```{r, include=F}
# Calculate quantiles
quants <-
    crs %>% 
    group_by(category_aoh) %>%
    summarize(diff25 = quantile(Difference, probs = 0.25, na.rm = TRUE), # lower
              diff50 = quantile(Difference, probs = 0.5, na.rm = TRUE), # median
              diff75 = quantile(Difference, probs = 0.75, na.rm = TRUE)) # upper

# Add to df
crs <- 
    left_join(crs, quants, by = "category_aoh") %>% 
    mutate(
        # Identify species furthest from downlisting threshold
        diff_top25 = Difference >= diff75,
        # Identify species closest to downlisting threshold
        diff_bottom25 = Difference <= diff25
        )

# Filter to highest risk species, furthest from downlisting threshold
# (top 25% of distance from threshold)
diff_top25 <- filter(crs, diff_top25)

# Filter to easy win species, closest to downlisting threshold
# (bottom 25% of distance from threshold)
diff_bottom25 <- filter(crs, diff_bottom25)
```


### Class
```{r}
top_diff %>% group_by(className) %>% summarize(Count = n()) %>% arrange(-Count)

ggplot(top_diff, aes(x=className)) + geom_bar() + ggtitle("Distribution of Top 25% (of each category) AOH Differences")

ggplot(top_diff, aes(x=className, fill=className)) + geom_bar() + facet_wrap(~Category) + ggtitle("Distribution of Top 25% AOH Differences") + theme(axis.text.x = element_text(angle = 90))
```


### Realm
```{r, include=F}
crit_realms = realms %>% filter(Category == 'Critically Endangered') %>% filter(Difference >= 9.231000)

endan_realms = realms %>% filter(Category == 'Endangered') %>% filter(Difference >= 457.821961)

vul_realms = realms %>% filter(Category == 'Vulnerable') %>% filter(Difference >= 1265.166894)

# near_realms = realms %>% filter(Category == 'Near Threatened') 

# also all combined
top_realm = full_join(crit_realms, endan_realms)
top_realm = full_join(top_realm, vul_realms)
#top_realm = full_join(bottom_realm, near_realms)
```


```{r}
ggplot(top_realm, aes(x=realm)) + geom_bar() + ggtitle("Distribution of Top 25% (of each category) AOH Differences Across Realm") + theme(axis.text.x = element_text(angle = 90))

ggplot(top_realm, aes(x=realm, fill=realm)) + geom_bar() + facet_wrap(~Category) + ggtitle("Distribution of Top 25% AOH Differences Across Realm") + theme(axis.text.x = element_text(angle = 90))
```

NOTE: I did proportion analysis and it was definitely not significant. 

### Category
```{r, include=F}
ggplot(top_diff, aes(x=Category, fill=Category)) + geom_bar() + ggtitle("Distribution of Top 25% (of each category) AOH Differences") + theme(axis.text.x = element_text(angle = 90))
```









# GLM
```{r}
library(lme4)
```

### CRS 
= column for species name + class + whether CRS (0 or 1)
glm(CRS ~ class, family = "binomial")
```{r}
crs = df %>% mutate(CRS = ifelse(is.na(Difference), 0, ifelse(Difference > 1, 1, 0)))

ggplot(crs, aes(x = as.factor(CRS))) +
    geom_bar() +
    facet_grid(Category ~ className) 
```
Near Threatened species are not CRS, all other categories have majority CRS. Amphibians have highest counts of CRS across categories. 

Create GLM
- Really high deviance --> class does not correlate with CRS 
```{r}
glm = glm(CRS ~ className, data = crs, family = "binomial")
summary(glm)
anova(glm, test = "LRT")
```

```{r}
glm5 = glm(CRS ~ className * Category, data = crs, family = "binomial")
summary(glm5)

# can remove interaction term
glm6 <- update(glm5, ~.- className:Category)
anova(glm5, glm6, test = "LRT")

# dropping category makes much worse
glm7 <- update(glm6, ~.- Category)
anova(glm6, glm7, test = "LRT")

# worse but still okay 
glm8 <- update(glm6, ~.- className)
anova(glm6, glm8, test = "LRT")

# highest AIC score is removing Category, lowest is removing className and interaction
AIC(glm5, glm6, glm7, glm8)
```
-- Category is most important variable
-- class is not 

### CRS_realm 
= species + class + whether CRS + realms
```{r}
sort(unique(unlist(strsplit(unique(crs$realm), "[|]"))))

# Use 'grepl' to see if string is contained in another string
realms_v2 = 
    mutate(crs,
           Afrotropical = grepl("Afrotropical", realm),
           Australasian = grepl("Australasian", realm),
           Indomalayan = grepl("Indomalayan", realm),
           Nearctic = grepl("Nearctic", realm),
           Neotropical = grepl("Neotropical", realm),
           Oceanian = grepl("Oceanian", realm),
           Palearctic = grepl("Palearctic", realm), 
           Antarctic = grepl("Antarctic", realm))

# long
realms_long <- realms_v2 %>%  pivot_longer(cols=c(Afrotropical, Australasian, Indomalayan, Nearctic, Neotropical, Oceanian, Palearctic, Antarctic), names_to='Realms', values_to='value')

realms_long_test <- realms_long %>% filter(value == TRUE)
```

GLM

```{r interaction-test}
glm_full <- 
    glm(CRS ~ className * Realms + className * Category + Realms * Category, 
        data = realms_long_test, 
        family = "binomial")

# Drop & test each interaction in turn
glm_dropInter1 <- update(glm_full, ~.- className:Realms)
anova(glm_dropInter1, glm_full, test = "LRT") # this interaction is not significant

glm_dropInter2 <- update(glm_full, ~.- className:Category)
anova(glm_dropInter2, glm_full, test = "LRT") # this interaction is not significant

glm_dropInter3 <- update(glm_full, ~.- Realms:Category)
anova(glm_dropInter3, glm_full, test = "LRT") # this interaction is not significant

#-------------------------------------------------------------------------------
# Dropping Realms:Category had the least effect (smallest change in Deviance, biggest p value)
# so make this the new reference model and repeat with remaining interactions
# ------------------------------------------------------------------------------

glm_dropInter4 <- update(glm_dropInter3, ~.- className:Realms)
anova(glm_dropInter4, glm_dropInter3, test = "LRT") # this interaction is not significant

glm_dropInter5 <- update(glm_dropInter3, ~.- className:Category)
anova(glm_dropInter5, glm_dropInter3, test = "LRT") # this interaction is not significant

#-------------------------------------------------------------------------------
# Dropping className:Category had the least effect (smallest change in Deviance, biggest p value)
# so make this the new reference model and repeat with remaining interaction
# ------------------------------------------------------------------------------

glm_dropInter6 <- update(glm_dropInter5, ~.- className:Realms)
anova(glm_dropInter6, glm_dropInter5, test = "LRT") # this interaction is not significant

#-------------------------------------------------------------------------------
# Drop ALL interactions - HOORAY
# now test significance of individual variables
# ------------------------------------------------------------------------------

# Test significance of class
glm_dropClass <- update(glm_dropInter6, ~.- className)
anova(glm_dropClass, glm_dropInter6, test = "LRT") # class is not significant
anova(glm_dropInter6, glm_dropClass, test = "LRT")

# Test significance of realm
glm_dropRealm <- update(glm_dropInter6, ~.- Realms)
anova(glm_dropRealm, glm_dropInter6, test = "LRT") # realm is not significant

# Test significance of category
glm_dropCat <- update(glm_dropInter6, ~.- Category)
anova(glm_dropCat, glm_dropInter6, test = "LRT") # category is significant

# Compare AIC as well, for good measure
AIC(glm_dropInter6,glm_dropClass,glm_dropRealm,glm_dropCat)
# all quite similar, except for dropping Category which produces a *much* worse model
```

```{r plot-model-output}
# Get model-predicted values --------------------------------------------------

# First we need to create a new dataframe with all possible values
newdat <-  
    expand.grid(Category = unique(realms_long_test$Category),
                className = unique(realms_long_test$className))

# Now we predict y values, using our model (here, just with class and category)
predvals <- predict(glm_dropRealm, newdata = newdat, 
                    # Return y values on the scale of the link function (logistic)
                    type = "link", 
                    # Return standard errors of model-predicted y values
                    se.fit = TRUE)

# Function to calculate critical significance value, used to convert
# standard error to 95% confidence intervals
ci.critical <- 
    function(siglevel) qnorm((100 - siglevel) / 100 / 2, lower.tail = FALSE)
ci_const <- ci.critical(95)

# Assign to dataframe
newdat <-
    newdat %>% 
    mutate(crs_prob = predvals$fit, 
           se = predvals$se.fit,
           CI_lo = crs_prob - (ci_const * se),
           CI_hi = crs_prob + (ci_const * se))
head(newdat)

# You'll see that these values appear to be crazy at first - this is because
# we specified a binomial error family, meaning that the values by default are 
# on a logit scale and need to be back-transformed before plotting:

# Function to calculate inverse logit
inv.logit <- function(x){
    exp(x)/(1 + exp(x))
}

# Now we can back transform the y values
newdat <-
    newdat %>% 
    mutate(crs_prob = inv.logit(crs_prob),
           se = inv.logit(se),
           CI_lo = inv.logit(CI_lo),
           CI_hi = inv.logit(CI_hi))
head(newdat)
# Values now range between zero and one

# Plot model-predicted values ---------------------------------------------

ggplot(newdat, aes(x = className, y = crs_prob, 
                   colour = Category)) +
    geom_point(position = position_dodge(width = 0.5)) +
    # 95% confidence intervals
    geom_errorbar(aes(ymin = CI_lo,
                      ymax = CI_hi),
                  position = position_dodge(width = 0.5),
                  width = 0) +
    ylab("Probability of being CRS")

# Generally similar probabilities among taxa, but amphibians a little lower

# Increasing probability of improvement if old category was higher risk of extinction

# No errorbars for LC/CR because these categories can only go in one direction

```



```{r}
glm9 = glm(CRS ~ Category * Realms, data = realms_long_test, family = "binomial")
summary(glm9)
anova(glm9, test = "LRT")

# don't remove interaction term, deviance is higher
glm10 <- update(glm9, ~.- className:Realms)
anova(glm9, glm10, test = "LRT")

# dropping Realms makes it worse
glm11 <- update(glm10, ~.- Realms)
anova(glm10, glm11, test = "LRT")

# dropping className makes it worse
glm12 <- update(glm11, ~.- className)
anova(glm11, glm12, test = "LRT")

# lowest AIC is glm9 with both categories and interaction term (still very high though)
AIC(glm9, glm10, glm11, glm12)



# Best model is still glm6 --> Category and className w/ no interaction or realm
glm13 = glm(CRS ~ className * Realms * Category, data = realms_long_test, family = "binomial")
anova(glm9, glm13, test = "LRT")
AIC(glm9, glm6, glm13)
```

Realm of each species
- get rid of species in multiple categories ? 
```{r}
glm14 = glm(CRS ~ className * realm, data = crs, family = "binomial")
anova(glm14, test = "LRT")
AIC(glm6, glm14)
```


Run lm for difference and class 
```{r}
# Make a standard lm
lm <- lm(Difference ~ className, 
          data = df)

# Model validation plots
par(mfrow = c(2,2))
plot(lm)
summary(lm)

# compare to lm for category
lm_cat <- lm(Difference ~ Category, 
          data = df)
par(mfrow = c(2,2))
plot(lm_cat)
summary(lm_cat)

```
className and Difference has a p-value of 0.0006433, but a low adjusted R-squared of 0.005839. Indicates that className is statistically significant.

- try difference / max difference for these models 
```{r}
crs_difference <- crs %>% mutate(difference_prop = ifelse(Category == "Critically Endangered", Difference / 10, ifelse(Category == "Endangered", Difference / 490, ifelse(Category == "Vulnerable", Difference / 1500, NA))))

# Make a standard lm
lm_prop <- lm(difference_prop ~ className, 
          data = crs_difference)

# Model validation plots
par(mfrow = c(2,2))
plot(lm_prop)
summary(lm_prop)

# Make a standard lm
lm_prop_cat <- lm(difference_prop ~ Category, 
          data = crs_difference)

# Model validation plots
par(mfrow = c(2,2))
plot(lm_prop_cat)
summary(lm_prop_cat)

# Make a standard lm
lm_prop_realm <- lm(difference_prop ~ realm, 
          data = crs_difference)

# Model validation plots
par(mfrow = c(2,2))
plot(lm_prop_realm)
summary(lm_prop_realm)
    
```
- much smaller p-value for category (close to 0) than classname (0.07748)
- very high p-value for realm (not good model)


```{r}
crs_top25 <- crs %>% mutate(top25 = ifelse(Category == 'Critically Endangered' & Difference >= 9.231000, 1, ifelse(Category == 'Endangered' & Difference >= 457.429708, 1, ifelse(Category == 'Vulnerable' & Difference >= 1263.658855, 1, 0))))

write.csv(crs_top25,"Desktop\\crs_top25.csv")
```


### Poisson
- group_by class and realm, summarize for sum, that sum is response variable for poisson distribution
- change family from binary to poisson 
```{r}
class_realm <- crs %>% group_by(className, realm) %>% summarize(Sum = n())

poisson <- 
    glm(Sum ~ className * realm, 
        data = class_realm, 
        family = "poisson")

anova(poisson)
summary(poisson)
```

Using Many GLM
```{r}
library(mvabund)
poisson2 <- manyglm(Sum ~ className * realm, data = class_realm, family = "poisson")

plot(poisson2)

anova(poisson2)
summary(poisson2)
```
P-value: 0.001


### Correlation between className and Category
```{r}
ggplot(crs, aes(className, Category)) + geom_count()

class_cat = table(crs$className, crs$Category) 

chisq <- chisq.test(class_cat)
chisq
```
p-value is very small (< 2.2e-16), therefore there is a strong correlation between Category and class
