---
title: "Big Q Data Analysis"
author: "Sophia Richter"
date: "6/22/2021"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(warning=F)
knitr::opts_chunk$set(message=F)
```


Libraries and read data set. 
```{r, include=F}
library(dplyr)
library(readr)
library(GGally)
library(ggplot2)

data <- read_csv('iucn_aoh_data.csv')

# add cat_change_PA
data <- data %>% mutate(cat_change_PA = ifelse(Category == "Near Threatened" & Category_PA == "Vulnerable", 1, ifelse(Category == "Near Threatened" & Category_PA == "Endangered", 2, ifelse(Category == "Near Threatened" & Category_PA == "Critically Endangered", 3,
ifelse(Category == "Vulnerable" & Category_PA == "Endangered", 1, ifelse(Category == "Vulnerable" & Category_PA == "Critically Endangered", 2,
ifelse(Category == "Endangered" & Category_PA == "Critically Endangered", 1, 0))))))) 

last_data <- data %>% filter(`system:index` == 23)
```

Restructure realm data so that a species can be in two realms
```{r, include=F}
neo = data %>% filter(realm == "Neotropical" | realm == "Nearctic|Neotropical" | realm == "Neotropical|Oceanian") %>% mutate(realm = "Neotropical")

afro = data %>% filter(realm == "Afrotropical" | realm == "Afrotropical|Palearctic" | realm == "Afrotropical|Indomalayan|Palearctic") %>% mutate(realm = "Afrotropical")

ocean = data %>% filter(realm == "Oceanian" | realm == "Neotropical|Oceanian" | realm == "Indomalayan|Oceanian" | realm == "Australasian|Oceanian" | realm == "Oceanian|Palearctic" | realm == "Indomalayan|Oceanian|Palearctic") %>% mutate(realm = "Oceanian")

indo = data %>% filter(realm == "Indomalayan" | realm == "Indomalayan|Palearctic" | realm == "Afrotropical|Indomalayan|Palearctic" | realm == "Indomalayan|Nearctic|Palearctic" | realm == "Indomalayan|Oceanian" | realm == "Australasian|Indomalayan" | realm == "Indomalayan|Oceanian|Palearctic" | realm == "Australasian|Indomalayan|Palearctic") %>% mutate(realm = "Indomalayan")

austra = data %>% filter(realm == "Australasian" | realm == "Australasian|Oceanian" | realm == "Australasian|Indomalayan" | realm == "Australasian|Indomalayan|Palearctic") %>% mutate(realm = "Australasian")

palea = data %>% filter(realm == "Palearctic" | realm == "Afrotropical|Palearctic" | realm == "Indomalayan|Palearctic" | realm == "Afrotropical|Indomalayan|Palearctic" | realm == "Indomalayan|Nearctic|Palearctic" | realm == "Oceanian|Palearctic" | realm == "Australasian|Indomalayan|Palearctic" | realm == "Nearctic|Palearctic" | realm == "Indomalayan|Oceanian|Palearctic") %>% mutate(realm = "Palearctic")

nea = data %>% filter(realm == "Nearctic" | realm == "Nearctic|Neotropical" | realm == "Indomalayan|Nearctic|Palearctic" | realm == "Nearctic|Palearctic") %>% mutate(realm = "Nearctic")

# combine with duplicates
realms = bind_rows(neo, afro)
realms = bind_rows(realms, ocean)
realms = bind_rows(realms, indo)
realms = bind_rows(realms, austra)
realms = bind_rows(realms, palea)
realms = bind_rows(realms, nea)

# checked to make sure species were divided correctly with a 'manual' count
```


# Which species require just a small increase in habitat protection to safeguard them from potential future uplisting? 

Look for patterns that indicate a species would have a small PA difference to next threshold. 
```{r, include=F}
# remove values with NA and extract the last year for each species
merge_diff_PA = data %>% filter(`system:index` == 23) %>%  filter(Difference_PA != "NA")

# Boxplots of each category
ggplot(merge_diff_PA, aes(x=className, y=Difference_PA)) + geom_boxplot() + ggtitle("PA Difference Across Class")

ggplot(merge_diff_PA, aes(x=realm, y=Difference_PA)) + geom_boxplot() + theme(axis.text.x = element_text(angle = 90)) + ggtitle("PA Difference Across Realm")

ggplot(merge_diff_PA, aes(x=Category, y=Difference_PA)) + geom_boxplot() + ggtitle("PA Difference Across Category")
```


```{r, include=F}
# Bottom 25% of each category
# NOTE: I included species that met the same number as the 25% quantile, so there are actually more than 25% species. I can change this if needed. 
crit_PA = merge_diff_PA %>% filter(Category == 'Critically Endangered') %>% filter(Difference_PA <= 8)

endan_PA = merge_diff_PA %>% filter(Category == 'Endangered') %>% filter(Difference_PA <= 10)

vul_PA = merge_diff_PA %>% filter(Category == 'Vulnerable') %>% filter(Difference_PA <= 164)

near_PA = merge_diff_PA %>% filter(Category == 'Near Threatened') %>% filter(Difference_PA <= 204.4029)

# also all combined
bottom_PA = full_join(crit_PA, endan_PA)
bottom_PA = full_join(bottom_PA, vul_PA)
bottom_PA = full_join(bottom_PA, near_PA)

# bottom 25% of all Differences
all_bottom_PA = merge_diff_PA %>% filter(Difference_PA <= 10)

```

### Class
```{r}
ggplot(bottom_PA, aes(x=className)) + geom_bar() + ggtitle("Distribution of Bottom 25% (of each category) PA Differences")

ggplot(bottom_PA, aes(x=className, fill=className)) + geom_bar() + facet_wrap(~Category) + ggtitle("Distribution of Bottom 25% PA Differences") + theme(axis.text.x = element_text(angle = 90))
```


### Realm
```{r, include=F}
merge_diff_realm = realms %>% filter(`system:index` == 23) %>%  filter(Difference_PA != "NA")

crit_realms = merge_diff_realm %>% filter(Category == 'Critically Endangered') %>% filter(Difference_PA <= 8)

endan_realms = merge_diff_realm %>% filter(Category == 'Endangered') %>% filter(Difference_PA <= 10)

vul_realms = merge_diff_realm %>% filter(Category == 'Vulnerable') %>% filter(Difference_PA <= 164)

near_realms = merge_diff_realm %>% filter(Category == 'Near Threatened') %>% filter(Difference_PA <= 204.4029)

# also all combined
bottom_realm = full_join(crit_realms, endan_realms)
bottom_realm = full_join(bottom_realm, vul_realms)
bottom_realm = full_join(bottom_realm, near_realms)
```

```{r}
ggplot(bottom_realm, aes(x=realm)) + geom_bar() + ggtitle("Distribution of Bottom 25% (of each category) PA Differences Across Realm") + theme(axis.text.x = element_text(angle = 90))

ggplot(bottom_realm, aes(x=realm, fill=realm)) + geom_bar() + facet_wrap(~Category) + ggtitle("Distribution of Bottom 25% PA Differences Across Realm") + theme(axis.text.x = element_text(angle = 90))

# Proportion
# NOTE: I could do proportion of all species or proportion of species with differences. I can also change it to bottom 25% of entire dataset
last_realm <- realms %>% filter(`system:index` == 23)

prop1 = last_realm %>% group_by(realm) %>% summarize(Tot_Sum = n()) %>% arrange(-Tot_Sum)
prop2 = bottom_realm %>% group_by(realm) %>% summarize(PA_Sum = n()) %>% arrange(-PA_Sum)
prop3 = inner_join(prop2, prop1, by=c('realm'='realm'))
prop3 = prop3 %>% mutate(Proportion = PA_Sum / Tot_Sum)

ggplot(prop3, aes(x=realm, y=Proportion)) + geom_col() + theme(axis.text.x = element_text(angle = 90)) + ggtitle("Proportion of species in bottom 25% across realm")

prop3 %>% dplyr::select(realm, Proportion, Tot_Sum, PA_Sum) %>% arrange(-Proportion)
```


### Category
```{r, eval=F}
ggplot(bottom_PA, aes(x=Category, fill=Category)) + geom_bar() + ggtitle("Distribution of Bottom 25% (of each category) PA Differences") + theme(axis.text.x = element_text(angle = 90))

# Distribution of bottom 25% of overall data
ggplot(all_bottom_PA, aes(x=Category, fill=Category)) + geom_bar() + ggtitle("Distribution of Bottom 25% (of each category) PA Differences") + theme(axis.text.x = element_text(angle = 90))
```


# Which species require just a small increase in overall habitat to potentially qualify for downlisting? 

Look for patterns that indicate a species would have a small AOH difference to next threshold. 
```{r, eval=F}
merge_diff = data %>% filter(`system:index` == 23) %>%  filter(Difference != "NA")
crit_diff <- merge_diff %>% filter(Category == "Critically Endangered")
endan_diff <- merge_diff %>% filter(Category == "Endangered")
vul_diff <- merge_diff %>% filter(Category == "Vulnerable")

# taxonomy (class)
ggplot(crit_diff, aes(x=className, y=Difference)) + geom_boxplot() + theme(axis.text.x = element_text(angle = 90)) + ggtitle("Difference (km^2) to Next Threshold Value Across Class \n for Critically Endangered Species")

ggplot(endan_diff, aes(x=className, y=Difference)) + geom_boxplot() + theme(axis.text.x = element_text(angle = 90)) + ggtitle("Difference (km^2) to Next Threshold Value Across Class \n for Endangered Species")

# ggplot(vul_diff, aes(x=className, y=Difference)) + geom_boxplot() + theme(axis.text.x = element_text(angle = 90)) + ggtitle("Difference (km^2) to Next Threshold Value Across Class \n for Vulnerable Species")


# Geography
ggplot(merge_diff, aes(x=realm, y=Difference)) + geom_boxplot() + theme(axis.text.x = element_text(angle = 90)) + ggtitle("Difference (km^2) to Next Threshold Value Across Realms")

# Population trend
ggplot(merge_diff, aes(x=populationTrend, y=Difference)) + geom_boxplot() + ggtitle("Population Trend v. Difference (km^2) to Next Threshold Value")
```


# Which species are somewhat doomed?
i.e. very little overall habitat + very little protected habitat = very far from being able to help these species

```{r, eval=F}
no_PA <- last_data %>% filter(PAkm == 0)

ggplot(last_data, aes(x=className, y=PAkm)) + geom_boxplot(outlier.shape = NA) + coord_cartesian(ylim = quantile(last_data$PAkm, c(0.1, 0.93))) + ggtitle("Distribution of Species across PAkm (outliers removed)")

ggplot(no_PA, aes(x=className)) + geom_bar() + ggtitle("Classes of species with 0 km^2 PA")

ggplot(last_data, aes(x=redlistCategory, y=PAkm)) + geom_boxplot(outlier.shape = NA) + coord_cartesian(ylim = quantile(last_data$PAkm, c(0.1, 0.87))) + ggtitle("Distribution of Red List Category across PAkm (outliers removed)")

ggplot(last_data, aes(x=Category, y=PAkm)) + geom_boxplot(outlier.shape = NA) + coord_cartesian(ylim = quantile(last_data$PAkm, c(0.1, 0.89))) + ggtitle("Distribution of Habitat Category across PAkm (outliers removed)")
```

Amphibians have lowest PA habitat. Threatened categories have lowest PA habitat. 
