---
title: "No Exit Species on the IUCN Red List"
author: "Sophia Richter"
date: "7/26/2021"
output:
    html_document: default
    pdf_document: default
---
    
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F)
knitr::opts_chunk$set(warning=F)
knitr::opts_chunk$set(message=F)
```


```{r}
# Libraries and read data set
library(dplyr)
library(tidyr)
library(readr)
library(GGally)
library(ggplot2)
```

```{r}
# Get a vector of all the filenames
files <- list.files(path = "jung2015_RL", 
                    # Keep the full file path
                    full.names = TRUE,
                    # Limit to CSV files
                    pattern = "\\.csv$")

# Read each one in with lapply & bind all
areadat <- 
    lapply(files, function(file_i){
        suppressMessages(read_csv(file_i))
    }) %>% 
    # Collapse list into dataframe
    bind_rows()
```

Merge area data and IUCN data
```{r}
# read iucn data
iucndat <- read_csv('iucn_dat.csv') %>% 
  # Rename the AOH variables to avoid confusion
  # (Jung refers to the land cover map by Jung et al. 2015)
  rename(areaAOH_jung = areaAOH,
         areaAOHinPA_jung = areaAOHinPA)

merge_dat <- inner_join(areadat, iucndat, by=c('species'='scientificName'))
```

# Data Frame Modifications
```{r}
# Modify data frame

# remove species with 0 or NA AOH
data <- merge_dat %>% filter(!(is.na(areaAOH)), areaAOH > 0)

# convert AOH to km^2
data <- data %>% mutate(AOHkm = areaAOH/1000000)


# add column for whether AOH falls below 9 km (more than 1 km away from threshold)
data <- data %>% mutate(ten = AOHkm < 9)
# add column for whether AOH falls below more than 1 km below 500 km threshold
data <- data %>% mutate(five_hundred = AOHkm < 499)
# add column for whether AOH falls below more than 1 km below 2000 km threshold
data <- data %>% mutate(two_thousand = AOHkm < 1999)


# column if ten is lowest threshold
data <- data %>% mutate(ten_lowest = ten)
# column if 500 is lowest threshold
data <- data %>% mutate(five_lowest = xor(ten, five_hundred))
# column if 2000 is lowest threshold
data <- data %>% mutate(two_lowest = xor(five_hundred, two_thousand))


# record what category should be
data <- data %>% mutate(Category = ifelse(ten_lowest == TRUE, "Critically Endangered", ifelse(five_lowest == TRUE, "Endangered", ifelse(two_lowest == TRUE, "Vulnerable", "Near Threatened"))))


# Calculate difference to next threshold
data <- data %>% mutate(Difference = ifelse(ten_lowest == TRUE, 10 - AOHkm, ifelse(five_lowest == TRUE, 500 - AOHkm, ifelse(two_lowest == TRUE, 2000 - AOHkm, NA))))


# remove Differences that are NA
test2 <- data %>% filter(!(is.na(Difference)))


# Calculate Difference between Category and redlistCategory
test <- test2 %>% mutate(num_Category = ifelse(Category == "Critically Endangered", 1, ifelse(Category == "Endangered", 2, ifelse(Category == "Vulnerable", 3, ifelse(Category == "Near Threatened", 4, NA)))))

test <- test %>% mutate(num_redlist = ifelse(redlistCategory == "Critically Endangered", 1, ifelse(redlistCategory == "Endangered", 2, ifelse(redlistCategory == "Vulnerable", 3, ifelse(redlistCategory == "Near Threatened", 4, NA)))))

test <- test %>% mutate(cat_red_change = num_Category - num_redlist)
```

There are 2488 species with a habitat that qualifies a species for a threatened category. 

Negative value means Habitat Category is uplisted compared to redlistCategory, 0 means they match, positive means Habitat Category is downlisted compared to redlistCategory. 
```{r}
test %>% group_by(cat_red_change) %>% summarize(Sum = n())

change2 <- test %>% filter(cat_red_change == -2) %>% select(species, className, redlistCategory)
```
1313 species (52.8%) do not change category. 549 species (22.1%) have less habitat than redlistCategory (- value). 626 species (25.2%) have more habitat than redlistCategory (+ value).

# No Exit Species (NES)

Classes of NES
```{r}
# Table
test %>% group_by(className) %>% summarize(Count = n()) %>% arrange(-Count)
```

Category of NES
```{r}
#table
test %>% group_by(Category) %>% summarize(Count = n()) %>% arrange(-Count)

# graphs
ggplot(test, aes(x=className, fill=className)) + geom_bar() + facet_wrap(~Category) + ggtitle("Class and Category of NES") + theme(axis.text.x = element_blank()) + labs(fill="Class") + xlab(" ") + ylab("Count")
```

Realm of Counts of NES
```{r}
sort(unique(unlist(strsplit(unique(test$realm), "[|]"))))

# Use 'grepl' to see if string is contained in another string
realms_v3 = 
    mutate(test,
           Afrotropical = grepl("Afrotropical", realm),
           Australasian = grepl("Australasian", realm),
           Indomalayan = grepl("Indomalayan", realm),
           Nearctic = grepl("Nearctic", realm),
           Neotropical = grepl("Neotropical", realm),
           Oceanian = grepl("Oceanian", realm),
           Palearctic = grepl("Palearctic", realm))

# long
realms_long <- realms_v3 %>%  pivot_longer(cols=c(Afrotropical, Australasian, Indomalayan, Nearctic, Neotropical, Oceanian, Palearctic), names_to='Realms', values_to='value')

realms_long <- realms_long %>% filter(value == TRUE)
```

Realms of each Conservation Reliant Species
```{r}
test %>% group_by(realm) %>% summarize(Count = n()) %>% arrange(-Count)
```

Number of CRS in each realm
```{r}
realms_long %>% group_by(Realms) %>% summarize(Count = n()) %>% arrange(-Count)

ggplot(realms_long, aes(x=Realms)) + geom_bar() + theme(axis.text.x = element_text(angle = 90)) + ggtitle("Number of CRS in each Realm")

ggplot(realms_long, aes(x=Realms, fill=Realms)) + geom_bar() + facet_wrap(~Category) + theme(axis.text.x = element_blank()) + ggtitle("Category and Realm of CRS")

```


# Difference Distribution

Class
```{r}
ggplot(test, aes(x=Difference, fill=className, alpha=0.01)) + geom_histogram() + ggtitle("Distribution of Differences to Next Threshold Across Class")

ggplot(test, aes(x=Difference, fill=className, alpha=0.01)) + geom_density() + ggtitle("Density of Differences to Next Threshold Across Class")

ggplot(test, aes(x=Difference)) + geom_histogram() + facet_wrap(~className) + ggtitle("Distribution of Differences to Next Threshold Across Class")

ggplot(test, aes(x=Difference)) + geom_density() + facet_wrap(~className) + ggtitle("Density of Differences to Next Threshold Across Class")
```

Histogram pf Differences by Class
```{r}
# I'm adding this one below with free scales so we can see what's going on in
# EN and VU without the distribution being squashed by CR
ggplot(test, aes(x=Difference, fill=className)) + geom_histogram(alpha=0.7) + ggtitle("Distribution of Differences to Next Threshold \n Across Class and Category") +
    facet_wrap(~ Category, scales = "free", labeller = label_wrap_gen(width = 2, multi_line = TRUE)) + labs(fill="Class") + ylab("Count") + xlab("Difference (km^2)")
```
 <!-- Cool - looks like most things are close to the threshold -->
 
Category
```{r}
ggplot(test, aes(x=Difference, fill=Category, alpha=0.5)) + geom_histogram() + ggtitle("Distribution of Differences to Next Threshold Across Category")
```

Realm
```{r}
ggplot(realms_long, aes(x=Difference, fill=Realms)) + geom_histogram() + ggtitle("Distribution of Differences Across Realm")

ggplot(realms_long, aes(x=Difference)) + geom_histogram() + facet_wrap(~Realms)

ggplot(realms_long, aes(x=Difference)) + geom_density() + facet_wrap(~Realms)  + ggtitle("Distribution of Differences Across Realm")

ggplot(realms_long, aes(x=Difference)) + geom_density() + facet_wrap(~Realms)  + ggtitle("Density of Differences Across Realm")
```


# Top 25% Differences

```{r, include=F}
# Top 25% of each category
crit_top = test %>% filter(Category == 'Critically Endangered') %>% filter(Difference >= 8.754866)

endan_top = test %>% filter(Category == 'Endangered') %>% filter(Difference >= 459.977142)

vul_top = test %>% filter(Category == 'Vulnerable') %>% filter(Difference >= 1278.32419)

# near_top = test %>% filter(Category == 'Near Threatened')

# also all combined
top_diff = full_join(crit_top, endan_top)
top_diff = full_join(top_diff, vul_top)
# bottom_PA = full_join(bottom_PA, near_PA)
```

### Class
```{r}
top_diff %>% group_by(className) %>% summarize(Count = n()) %>% arrange(-Count)

ggplot(top_diff, aes(x=className)) + geom_bar() + ggtitle("Distribution of Top 25% (of each category) AOH Differences")

ggplot(top_diff, aes(x=className, fill=className)) + geom_bar() + facet_wrap(~Category) + ggtitle("Distribution of Top 25% AOH Differences") + theme(axis.text.x = element_text(angle = 90))
```

### Category
```{r, include=F}
ggplot(top_diff, aes(x=Category, fill=Category)) + geom_bar() + ggtitle("Distribution of Top 25% (of each category) AOH Differences") + theme(axis.text.x = element_text(angle = 90))
```

### Realm
```{r, include=F}
crit_realms = realms_long %>% filter(Category == 'Critically Endangered') %>% filter(Difference >= 8.754419)

endan_realms = realms_long %>% filter(Category == 'Endangered') %>% filter(Difference >= 457.821961)

vul_realms = realms_long %>% filter(Category == 'Vulnerable') %>% filter(Difference >= 1265.166894)

# near_realms = realms %>% filter(Category == 'Near Threatened') 

# also all combined
top_realm = full_join(crit_realms, endan_realms)
top_realm = full_join(top_realm, vul_realms)
#top_realm = full_join(bottom_realm, near_realms)
```

```{r}
ggplot(top_realm, aes(x=Realms)) + geom_bar() + ggtitle("Distribution of Top 25% (of each category) AOH Differences Across Realm") + theme(axis.text.x = element_text(angle = 90))

ggplot(top_realm, aes(x=Realms, fill=Realms)) + geom_bar() + facet_wrap(~Category) + ggtitle("Distribution of Top 25% AOH Differences Across Realm") + theme(axis.text.x = element_text(angle = 90))
```



# GLM
```{r}
library(lme4)
```

### NES
= column for species name + class + whether NES (0 or 1)
glm(NES ~ class, family = "binomial")
```{r}
nes = data %>% mutate(NES = AOHkm < 1999)

ggplot(nes, aes(x = as.factor(NES))) +
    geom_bar() +
    facet_grid(Category ~ className) 
```
Near Threatened species are not NES, all other categories have majority NES. Amphibians have highest counts of NES across threatened categories but lower counts than birds and mammals for near threatened. 


Realm NES
```{r}
sort(unique(unlist(strsplit(unique(nes$realm), "[|]"))))

# Use 'grepl' to see if string is contained in another string
realms_v4 = 
    mutate(nes,
           Afrotropical = grepl("Afrotropical", realm),
           Australasian = grepl("Australasian", realm),
           Indomalayan = grepl("Indomalayan", realm),
           Nearctic = grepl("Nearctic", realm),
           Neotropical = grepl("Neotropical", realm),
           Oceanian = grepl("Oceanian", realm),
           Palearctic = grepl("Palearctic", realm), 
           Antarctic = grepl("Antarctic", realm))

# long
realms_long_nes <- 
  realms_v4 %>%  
  pivot_longer(cols=c(Afrotropical, Australasian, Indomalayan, Nearctic, 
                      Neotropical, Oceanian, Palearctic, Antarctic), 
               names_to='Realms', values_to='value') %>% 
  filter(value == TRUE) %>% 
  dplyr::select(species,className,NES,Realms,Category,AOHkm,Difference)
# I simplified the above so the dataframe is easier to inspect
```


```{r interaction-test}
glm_full_nes <- 
    glm(NES ~ className * Realms + className * Category + Realms * Category, 
        data = realms_long_nes , 
        family = "binomial", maxit=100)

# Drop & test each interaction in turn
glm_dropInter1_nes <- update(glm_full_nes, ~.- className:Realms)
anova(glm_dropInter1_nes, glm_full_nes, test = "LRT") # this interaction is not significant

glm_dropInter2_nes <- update(glm_full_nes, ~.- className:Category)
anova(glm_dropInter2_nes, glm_full_nes, test = "LRT") # this interaction is not significant

glm_dropInter3_nes <- update(glm_full_nes, ~.- Realms:Category)
anova(glm_dropInter3_nes, glm_full_nes, test = "LRT") # this interaction is not significant

#-------------------------------------------------------------------------------
# All had equally small effect (p-value of 1, small deviance). dropping realm:Category has highest DF
# so make this the new reference model and repeat with remaining interactions
# ------------------------------------------------------------------------------

glm_dropInter4_nes <- update(glm_dropInter3_nes, ~.- className:Realms)
anova(glm_dropInter4_nes, glm_dropInter3_nes, test = "LRT") # this interaction is not significant

glm_dropInter5_nes <- update(glm_dropInter3_nes, ~.- className:Category)
anova(glm_dropInter5_nes, glm_dropInter3_nes, test = "LRT") # this interaction is not significant

#-------------------------------------------------------------------------------
# Dropping either interaction term had same deviance and p-value effect. Dropping className:Category had higher DF
# so make this the new reference model and repeat with remaining interaction
# ------------------------------------------------------------------------------

glm_dropInter6_nes <- update(glm_dropInter5_nes, ~.- className:Realms)
anova(glm_dropInter6_nes, glm_dropInter5_nes, test = "LRT") # this interaction is not significant

#-------------------------------------------------------------------------------
# Drop ALL interactions - HOORAY
# now test significance of individual variables
# ------------------------------------------------------------------------------

# Test significance of class
glm_dropClass_nes <- update(glm_dropInter6_nes, ~.- className)
anova(glm_dropClass_nes, glm_dropInter6_nes, test = "LRT") # class is not significant

# Test significance of realm
glm_dropRealm_nes <- update(glm_dropInter6_nes, ~.- Realms)
anova(glm_dropRealm_nes, glm_dropInter6_nes, test = "LRT") # realm is not significant

# Test significance of category
glm_dropCat_nes <- update(glm_dropInter6_nes, ~.- Category)
anova(glm_dropCat_nes, glm_dropInter6_nes, test = "LRT") # category is significant

# Compare AIC as well, for good measure
AIC(glm_dropInter6_nes,glm_dropClass_nes,glm_dropRealm_nes,glm_dropCat_nes)
# all quite similar, except for dropping Category which produces a *much* worse model
```

--> I am not sure if we want to add the plot of model-predicted values? I would guess it is very similar (in that it is not helpful) to the last data set. 

### LM on Difference 
- try difference / max difference for these models 
```{r}
nes_difference <- nes %>% mutate(difference_prop = ifelse(Category == "Critically Endangered", Difference / 10, ifelse(Category == "Endangered", Difference / 490, ifelse(Category == "Vulnerable", Difference / 1500, NA))))

# Make a standard lm
lm_prop <- lm(difference_prop ~ className, 
          data = nes_difference)

# Model validation plots
par(mfrow = c(2,2))
plot(lm_prop)
summary(lm_prop)

# Make a standard lm
lm_prop_cat <- lm(difference_prop ~ Category, 
          data = nes_difference)

# Model validation plots
par(mfrow = c(2,2))
plot(lm_prop_cat)
summary(lm_prop_cat)

# Make a standard lm
lm_prop_realm <- lm(difference_prop ~ realm, 
          data = nes_difference)

# Model validation plots
par(mfrow = c(2,2))
plot(lm_prop_realm)
summary(lm_prop_realm)
    
```
- much smaller p-value for category (close to 0) than classname (0.06501)
- very high p-value for realm

```{r nes-proximity}
nes_difference <- nes %>% 
  mutate(difference_prop = ifelse(Category == "Critically Endangered", Difference / 10, 
                                  ifelse(Category == "Endangered", Difference / 490, 
                                         ifelse(Category == "Vulnerable", Difference / 1500, NA))),
         # Flip the scale so bigger numbers = closer to threshold for downlisting
         proximity = 1 - difference_prop) %>% 
  dplyr::select(species, AOHkm, Category, 
                Difference, difference_prop, proximity,
                className, realm) %>% 
  filter(
    # Drop NAs
    !(is.na(Difference)),
    # Restrict to species occurring in one Realm only
    !(grepl("[|]", realm))
    )

# Round negative values to zero (these are the ones that are < 1 km from the threshold)
nes_difference$proximity[nes_difference$proximity < 0] <- 0

# Plot the distribution of standardised differences
ggplot(nes_difference, 
       aes(x = proximity, fill = className)) +
  geom_histogram() +
  facet_wrap(~ Category)

# GLM with Poisson because relative diff is actually a proportion
glm_diff <- glm(proximity ~ className + realm + AOHkm, 
                data = nes_difference,
                family = "poisson") # ignore the non-integer warnings

# Model inference
summary(glm_diff)
anova(glm_diff, test = "LRT")

# Drop class
anova(update(glm_diff, ~.- className), glm_diff, test = "LRT")
# Drop realm
anova(update(glm_diff, ~.- realm), glm_diff, test = "LRT")
# Drop AOH
anova(update(glm_diff, ~.- AOHkm), glm_diff, test = "LRT")

### Only absolute AOH is a strong predictor or proximity to threshold

#############################
# Plot model-predicted values
#############################

# Make up some new data
newdat <- expand.grid(
  # All levels of taxonomic class
  className = unique(glm_diff$data$className),
  # All levels of realm
  realm = unique(glm_diff$data$realm),
  # Sequence of 100 made up values for AOH, bounded by observed data
  AOHkm = seq(min(glm_diff$data$AOHkm),
              max(glm_diff$data$AOHkm),
              length.out = 100))

# Use model to predict distance values
predvals <- predict(glm_diff, newdata = newdat, type = "link", se.fit = TRUE)

# Function to calculate inverse logit
inv.logit <- function(x) exp(x)/(1 + exp(x))

# Function to calculate critical significance value
ci.critical <- 
  function(siglevel) qnorm((100 - siglevel) / 100 / 2, lower.tail = FALSE)
ci_const = ci.critical(95)

# Add vals to newdat dataframe
newdat <- 
  newdat %>% 
  mutate(proximity = predvals$fit,
         se = predvals$se.fit,
         CI_lo = proximity - (ci_const * se),
         CI_hi = proximity + (ci_const * se)) 
# Back transform
newdat[,c("proximity", "CI_lo", "CI_hi")] <- 
  inv.logit(newdat[,c("proximity", "CI_lo", "CI_hi")])

# Plot model-predicted values
ggplot(newdat, 
       aes(x = AOHkm, y = proximity)) +
  # # Add confidence intervals
  # geom_ribbon(aes(ymin = CI_lo,
  #                 ymax = CI_hi,
  #                 fill = realm),
  #             alpha = 0.3) +
  # Add fitted line
  geom_line(aes(colour = realm)) +
  # Facet by taxonomic class
  facet_grid(~ className) +
  theme_classic()

# Since class and realm aren't relevant, we can also just pick a reference level
# or average across everything, and plot one line only
newdat_avg <- 
  group_by(newdat, AOHkm) %>% 
  summarise(proximity = median(proximity),
            CI_lo = median(CI_lo),
            CI_hi = median(CI_hi))
ggplot(newdat_avg, 
       aes(x = AOHkm, y = proximity)) +
  # Add confidence intervals
  geom_ribbon(aes(ymin = CI_lo,
                  ymax = CI_hi),
              alpha = 0.1) +
  # Add fitted line
  geom_line() +
  theme_classic()

# Proximity to threshold increases as absolute AOH increases
```


### Correlation between className and Category
```{r}
ggplot(nes, aes(className, Category)) + geom_count()

class_cat = table(nes$className, nes$Category) 

chisq <- chisq.test(class_cat)
chisq
```
p-value is very small (< 2.2e-16), therefore there is a strong correlation between Category and class


# Spatial Data
Save data frame for spatial analysis

```{r}

nes_top25 <- data %>% mutate(NES = AOHkm < 1999) %>% mutate(Top25 = ifelse(Category == 'Critically Endangered' & Difference >= 8.754866, 1, ifelse(Category == 'Endangered' & Difference >= 459.977142, 1, ifelse(Category == 'Vulnerable' & Difference >= 1278.32419, 1, 0))))

nes_top25 <- nes_top25 %>% rename(scientificName = species) %>% rename(top25 = Top25)

write.csv(nes_top25,"nes_top25.csv")

```

