---
title: "No Exit Species on the IUCN Red List"
author: "Sophia Richter"
date: "7/26/2021"
output:
    html_document: default
    pdf_document: default
---

My changes below might look major, but the code is going the exact same thing as
before. The purpose of dplyr and the pipe (%>%) system is to avoid the need to
overwrite objects or keep making new objects that are almost the same but with
a new column - you just add dataframe modifications on top of each other
in the same chunk of code. I was getting confused by the test vs. test 2 vs. 
data vs. nes

I encourage you to 'wrap' your code (press enter - as I do here) after 
'%>%', after '+' (in ggplot code) and at ~80 characters, to improve readability. 
You can actually get RStudio to add a vertical line for you at a specified 
number of characters, so you know where to wrap: 
Tools -> Global Options -> Code -> Display -> Show margin

I have added names to your code chunks to make them easier to navigate through.
You can see them in the markdown outline (Ctrl + Shift + O) if you turn them on:
Tools -> Global Options -> R Markdown -> Show in document outline -> Sections and Named Chunks

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F)
knitr::opts_chunk$set(warning=F)
knitr::opts_chunk$set(message=F)
```

```{r load-packages}
# Data wrangling
library(dplyr)
library(tidyr)
library(readr)
# Figures
library(GGally)
library(ggplot2)
# Stats
library(lme4)
```

```{r load-raw-data, eval = FALSE}
# Get a vector of all the filenames
files <- list.files(path = "jung2015_RL", 
                    # Keep the full file path
                    full.names = TRUE,
                    # Limit to CSV files
                    pattern = "\\.csv$")

# Read each one in with lapply & bind all
areadat <- 
    lapply(files, function(file_i){
        suppressMessages(read_csv(file_i))
    }) %>% 
    # Collapse list into dataframe
    bind_rows()
```

Merge area data and IUCN data
```{r merge-data, eval = FALSE}
# Function to reformat new country & realm info
gsubGeo <- function(x){gsub(", ",";",gsub("[]]","",gsub("[[]","",x)))}

# read iucn data
iucndat <- read_csv('iucn_dat.csv') %>% 
  # Rename the AOH variables to avoid confusion
  # (Jung refers to the land cover map by Jung et al. 2015)
  rename(areaAOH_jung = areaAOH,
         areaAOHinPA_jung = areaAOHinPA) %>% 
  # Reformat realm & country info
  mutate(Realms = gsubGeo(realms),
         countries = gsubGeo(countries))
merge_dat <- 
  inner_join(areadat, iucndat, by=c('species'='scientificName')) %>% 
  # Limit to the columns we need
  dplyr::select(species,className,areaAOH,Realms,countries,redlistCategory) %>% 
  # Record if species are realm/country endemic
  mutate(realm_endemic = !(grepl(";", Realms)),
         country_endemic = !(grepl(";", countries)))

```

```{r sort-realms, eval = FALSE}
# Look at the realm categories
realm_cats <- sort(unique(unlist(strsplit(unique(merge_dat$Realms), "[;]"))))

# Use 'grepl' to see if string is contained in another string
realm_info <- 
  lapply(realm_cats, function(x) grepl(x, merge_dat$Realms)) %>% 
  bind_cols()
colnames(realm_info) <- realm_cats

# Add to merge_dat
merge_dat <- cbind(merge_dat, realm_info)

# The above steps take ages, so better to save as .Rds and read back in
saveRDS(merge_dat, "HMEI-GDrive/data/sorted/merge_dat.Rds")
```

# Data Frame Modifications

```{r load-sorted-data}
merge_dat <- readRDS("HMEI-GDrive/data/sorted/merge_dat.Rds")
realm_cats <- sort(unique(unlist(strsplit(unique(merge_dat$Realms), "[;]"))))
```

```{r aoh-thresholds}
# Modify data frame
all_df <- merge_dat %>% 
  # remove species with 0 or NA AOH
  filter(!(is.na(areaAOH)), areaAOH > 0) %>%
  mutate(
    # convert AOH to km^2
    AOHkm = areaAOH/1000000,
    # Identify NES
    NES = AOHkm < 1999,
    # add column for whether AOH falls below 9 km (more than 1 km away from threshold)
    ten = AOHkm < 9, 
    # add column for whether AOH falls below more than 1 km below 500 km threshold
    five_hundred = AOHkm < 499,
    # add column for whether AOH falls below more than 1 km below 2000 km threshold
    two_thousand = AOHkm < 1999,
    # column if ten is lowest threshold
    ten_lowest = ten,
    # column if 500 is lowest threshold
    five_lowest = xor(ten, five_hundred),
    # column if 2000 is lowest threshold
    two_lowest = xor(five_hundred, two_thousand),
    # record what category should be
    Category = ifelse(ten_lowest == TRUE, "Critically Endangered", 
                      ifelse(five_lowest == TRUE, "Endangered", 
                             ifelse(two_lowest == TRUE, "Vulnerable", 
                                    "Near Threatened"))),
    # Calculate difference to next threshold
    Difference = ifelse(ten_lowest == TRUE, 10 - AOHkm, 
                        ifelse(five_lowest == TRUE, 500 - AOHkm, 
                               ifelse(two_lowest == TRUE, 2000 - AOHkm, 
                                      NA)))) %>% 
  # Calculate Difference between Category and redlistCategory
  mutate(num_Category = ifelse(Category == "Critically Endangered", 1, 
                               ifelse(Category == "Endangered", 2, 
                                      ifelse(Category == "Vulnerable", 3, 
                                             ifelse(Category == "Near Threatened", 4, 
                                                    NA)))),
         num_redlist = ifelse(redlistCategory == "Critically Endangered", 1, 
                              ifelse(redlistCategory == "Endangered", 2, 
                                     ifelse(redlistCategory == "Vulnerable", 3, 
                                            ifelse(redlistCategory == "Near Threatened", 4, 
                                                   NA)))),
         cat_red_change = num_Category - num_redlist,
         # Standardised difference
         difference_prop = ifelse(Category == "Critically Endangered", Difference / 10, 
                                  ifelse(Category == "Endangered", Difference / 490, 
                                         ifelse(Category == "Vulnerable", Difference / 1500, 
                                                NA))),
         # Flip the scale so bigger numbers = closer to threshold for downlisting
         proximity = 1 - difference_prop)

nes_df <-
  all_df %>% 
  # remove Differences that are NA
  filter(!(is.na(Difference)))

# Sanity check
unique(nes_df$NES)
```

```{r realm-long-df}
# Long dataframe
realms_long_all <- 
  all_df %>%  
  dplyr::select(-Realms) %>% 
  pivot_longer(cols= all_of(realm_cats), 
               names_to='Realms', values_to='value') %>% 
  filter(value == TRUE) %>% 
  dplyr::select(-value)
realms_long_nes <-
  realms_long_all %>% 
  filter(NES)
```

There are 2488 species with a habitat that qualifies a species for a threatened category. 

Negative value means Habitat Category is uplisted compared to redlistCategory, 0 means they match, positive means Habitat Category is downlisted compared to redlistCategory.

```{r summarize-category-change}
all_df %>% 
  group_by(cat_red_change) %>% 
  summarise(Sum = n())

change_dat <- 
  all_df %>% 
  filter(cat_red_change == -2) %>% 
  select(species, className, redlistCategory)
```
1313 species (52.8%) do not change category. 549 species (22.1%) have less habitat than redlistCategory (- value). 626 species (25.2%) have more habitat than redlistCategory (+ value).

# No Exit Species (NES)

Classes of NES
```{r nes-classes}
# Table
nes_df %>% 
  group_by(className) %>% 
  summarise(Count = n()) %>% 
  arrange(-Count)
```

Category of NES
```{r nes-category}
#table
nes_df %>% 
  group_by(Category) %>% 
  summarize(Count = n()) %>% 
  arrange(-Count)

# graphs
ggplot(nes_df, aes(x=className, fill=className)) + 
  geom_bar() + 
  facet_wrap(~Category) + 
  ggtitle("Class and Threshold of NES") + 
  theme(axis.text.x = element_blank()) + 
  labs(fill="Class") + xlab(" ") + ylab("Count")
```

Realm of Counts of NES

Realms of each Conservation Reliant Species
```{r nes-realm}
nes_df %>% 
  group_by(Realms) %>% 
  summarise(Count = n()) %>% 
  arrange(-Count)
```

Number of CRS in each realm
```{r realm-nes-count}
realms_long_nes %>% 
  group_by(Realms) %>% 
  summarize(Count = n()) %>% 
  arrange(-Count)

ggplot(realms_long_nes, aes(x=Realms)) + 
  geom_bar() + 
  theme(axis.text.x = element_text(angle = 90)) + 
  ggtitle("Number of CRS in each Realm")

ggplot(realms_long_nes, aes(x=Realms, fill=Realms)) + 
  geom_bar() + 
  facet_wrap(~Category) + 
  theme(axis.text.x = element_blank()) + 
  ggtitle("Category and Realm of CRS")

```

# Difference Distribution

Class
```{r diff-density-class}
ggplot(nes_df, aes(x=Difference, fill=className, alpha=0.01)) + 
  geom_histogram() + 
  ggtitle("Distribution of Differences to Next Threshold Across Class")

ggplot(nes_df, aes(x=Difference, fill=className, alpha=0.01)) + 
  geom_density() + 
  ggtitle("Density of Differences to Next Threshold Across Class")

ggplot(nes_df, aes(x=Difference)) + 
  geom_histogram() + 
  facet_wrap(~className) + 
  ggtitle("Distribution of Differences to Next Threshold Across Class")

ggplot(nes_df, aes(x=Difference)) + 
  geom_density() + 
  facet_wrap(~className) + 
  ggtitle("Density of Differences to Next Threshold Across Class")
```

Histogram of Differences by Class
```{r diff-hist-class}
ggplot(nes_df, aes(x=Difference, fill=className)) + 
  geom_histogram(alpha=0.7) + 
  ggtitle("Distribution of Differences to Next Threshold \n Across Class and Category") +
  facet_wrap(~ Category, scales = "free", 
             labeller = label_wrap_gen(width = 2, multi_line = TRUE)) + 
  labs(fill="Class") + 
  ylab("Count") + 
  xlab("Difference (km^2)")
```

Category
```{r diff-hist-category}
ggplot(nes_df, aes(x=Difference, fill=Category, alpha=0.5)) + 
  geom_histogram() + 
  ggtitle("Distribution of Differences to Next Threshold Across Category")
```

Realm
```{r diff-hist-realm}
ggplot(realms_long_nes, aes(x=Difference, fill=Realms)) + 
  geom_histogram() + 
  ggtitle("Distribution of Differences Across Realm")

ggplot(realms_long_nes, aes(x=Difference)) + 
  geom_histogram() + 
  facet_wrap(~Realms)

ggplot(realms_long_nes, aes(x=Difference)) + 
  geom_density() + 
  facet_wrap(~Realms) + 
  ggtitle("Distribution of Differences Across Realm")

ggplot(realms_long_nes, aes(x=Difference)) + 
  geom_density() + 
  facet_wrap(~Realms) + 
  ggtitle("Density of Differences Across Realm")
```


# Top 25% Differences

```{r top-25-diff, include=F}
# Top 25% of each category
crit_top = nes_df %>% filter(Category == 'Critically Endangered') %>% filter(Difference >= 8.754866)

endan_top = nes_df %>% filter(Category == 'Endangered') %>% filter(Difference >= 459.977142)

vul_top = nes_df %>% filter(Category == 'Vulnerable') %>% filter(Difference >= 1278.32419)

# near_top = test %>% filter(Category == 'Near Threatened')

# also all combined
top_diff = full_join(crit_top, endan_top)
top_diff = full_join(top_diff, vul_top)
# bottom_PA = full_join(bottom_PA, near_PA)
```

### Class
```{r top-diff-class}
top_diff %>% 
  group_by(className) %>% 
  summarise(Count = n()) %>% 
  arrange(-Count)

ggplot(top_diff, aes(x=className)) + 
  geom_bar() + 
  ggtitle("Distribution of Top 25% (of each category) AOH Differences")

ggplot(top_diff, aes(x=className, fill=className)) + 
  geom_bar() + 
  facet_wrap(~Category) + 
  ggtitle("Distribution of Top 25% AOH Differences") + 
  theme(axis.text.x = element_text(angle = 90))
```

### Category
```{r top-diff-category, include=F}
ggplot(top_diff, aes(x=Category, fill=Category)) + 
  geom_bar() + 
  ggtitle("Distribution of Top 25% (of each category) AOH Differences") + 
  theme(axis.text.x = element_text(angle = 90))
```

### Realm
```{r top-diff-realm, include=F}
crit_realms = realms_long_nes %>% filter(Category == 'Critically Endangered') %>% filter(Difference >= 8.754419)

endan_realms = realms_long_nes %>% filter(Category == 'Endangered') %>% filter(Difference >= 457.821961)

vul_realms = realms_long_nes %>% filter(Category == 'Vulnerable') %>% filter(Difference >= 1265.166894)

# near_realms = realms %>% filter(Category == 'Near Threatened') 

# also all combined
top_realm = full_join(crit_realms, endan_realms)
top_realm = full_join(top_realm, vul_realms)
#top_realm = full_join(bottom_realm, near_realms)

ggplot(top_realm, aes(x=Realms)) + 
  geom_bar() + 
  ggtitle("Distribution of Top 25% (of each category) AOH Differences Across Realm") + 
  theme(axis.text.x = element_text(angle = 90))

ggplot(top_realm, aes(x=Realms, fill=Realms)) + 
  geom_bar() + 
  facet_wrap(~Category) + 
  ggtitle("Distribution of Top 25% AOH Differences Across Realm") + 
  theme(axis.text.x = element_text(angle = 90))
```

# GLM

### NES
= column for species name + class + whether NES (0 or 1)
glm(NES ~ class, family = "binomial")

```{r nes-count-plot}
ggplot(all_df, aes(x = as.factor(NES))) +
  geom_bar() +
  facet_grid(Category ~ className) 
```
Near Threatened species are not NES, all other categories have majority NES. Amphibians have highest counts of NES across threatened categories but lower counts than birds and mammals for near threatened. 

Realm NES

```{r interaction-test}
# I think we should focus on endemic species here, to avoid double-counting
glm_full_nes <- 
  glm(NES ~ className * Realms + className * Category + Realms * Category, 
      data = filter(realms_long_all, realm_endemic),
      family = "binomial", maxit=100)

# Drop & test each interaction in turn
glm_dropInter1_nes <- update(glm_full_nes, ~.- className:Realms)
anova(glm_dropInter1_nes, glm_full_nes, test = "LRT") # this interaction is not significant

glm_dropInter2_nes <- update(glm_full_nes, ~.- className:Category)
anova(glm_dropInter2_nes, glm_full_nes, test = "LRT") # this interaction is not significant

glm_dropInter3_nes <- update(glm_full_nes, ~.- Realms:Category)
anova(glm_dropInter3_nes, glm_full_nes, test = "LRT") # this interaction is not significant

#-------------------------------------------------------------------------------
# All had equally small effect (p-value of 1, small deviance). dropping realm:Category has highest DF
# so make this the new reference model and repeat with remaining interactions
# ------------------------------------------------------------------------------

glm_dropInter4_nes <- update(glm_dropInter3_nes, ~.- className:Realms)
anova(glm_dropInter4_nes, glm_dropInter3_nes, test = "LRT") # this interaction is not significant

glm_dropInter5_nes <- update(glm_dropInter3_nes, ~.- className:Category)
anova(glm_dropInter5_nes, glm_dropInter3_nes, test = "LRT") # this interaction is not significant

#-------------------------------------------------------------------------------
# Dropping either interaction term had same deviance and p-value effect. Dropping className:Category had higher DF
# so make this the new reference model and repeat with remaining interaction
# ------------------------------------------------------------------------------

glm_dropInter6_nes <- update(glm_dropInter5_nes, ~.- className:Realms)
anova(glm_dropInter6_nes, glm_dropInter5_nes, test = "LRT") # this interaction is not significant

#-------------------------------------------------------------------------------
# Drop ALL interactions - HOORAY
# now test significance of individual variables
# ------------------------------------------------------------------------------

# Test significance of class
glm_dropClass_nes <- update(glm_dropInter6_nes, ~.- className)
anova(glm_dropClass_nes, glm_dropInter6_nes, test = "LRT") # class is not significant

# Test significance of realm
glm_dropRealm_nes <- update(glm_dropInter6_nes, ~.- Realms)
anova(glm_dropRealm_nes, glm_dropInter6_nes, test = "LRT") # realm is not significant

# Test significance of category
glm_dropCat_nes <- update(glm_dropInter6_nes, ~.- Category)
anova(glm_dropCat_nes, glm_dropInter6_nes, test = "LRT") # category is significant

# Compare AIC as well, for good measure
AIC(glm_dropInter6_nes,glm_dropClass_nes,glm_dropRealm_nes,glm_dropCat_nes)
# all quite similar, except for dropping Category which produces a *much* worse model
```

--> I am not sure if we want to add the plot of model-predicted values? I would guess it is very similar (in that it is not helpful) to the last data set. 

### LM on Difference 
- try difference / max difference for these models 
```{r}
nes_difference <- 
  nes_df %>% 
  filter(
    # Drop NAs
    !(is.na(Difference)),
    # Restrict to species occurring in one Realm only
    realm_endemic
  )
# Make a standard lm
lm_prop <- lm(difference_prop ~ className, 
              data = nes_df)

# Model validation plots
par(mfrow = c(2,2))
plot(lm_prop)
summary(lm_prop)

# Make a standard lm
lm_prop_cat <- lm(difference_prop ~ Category, 
                  data = nes_difference)

# Model validation plots
par(mfrow = c(2,2))
plot(lm_prop_cat)
summary(lm_prop_cat)

# Make a standard lm
lm_prop_realm <- lm(difference_prop ~ Realms, 
                    data = nes_df)

# Model validation plots
par(mfrow = c(2,2))
plot(lm_prop_realm)
summary(lm_prop_realm)

```
- much smaller p-value for category (close to 0) than classname (0.06501)
- very high p-value for realm

```{r nes-proximity}
# Round negative values to zero (these are the ones that are < 1 km from the threshold)
nes_difference$proximity[nes_difference$proximity < 0] <- 0

# Plot the distribution of standardised differences
ggplot(nes_difference, 
       aes(x = proximity, fill = className)) +
  geom_histogram() +
  facet_wrap(~ Category)

# GLM with Poisson because relative diff is actually a proportion
glm_diff <- glm(proximity ~ className + Realms + AOHkm, 
                data = nes_difference,
                family = "binomial") # ignore the non-integer warnings

# Model inference
summary(glm_diff)
anova(glm_diff, test = "LRT")

# Drop class
class_results <- anova(update(glm_diff, ~.- className), glm_diff, test = "LRT")
# Drop realm
realm_results <- anova(update(glm_diff, ~.- Realms), glm_diff, test = "LRT")
# Drop AOH
aoh_results <- anova(update(glm_diff, ~.- AOHkm), glm_diff, test = "LRT")

### Only absolute AOH is a strong predictor or proximity to threshold

#############################
# Plot model-predicted values
#############################

# Make up some new data
newdat <- expand.grid(
  # All levels of taxonomic class
  className = unique(glm_diff$data$className),
  # All levels of realm
  Realms = unique(glm_diff$data$Realms),
  # Sequence of 100 made up values for AOH, bounded by observed data
  AOHkm = seq(min(glm_diff$data$AOHkm),
              max(glm_diff$data$AOHkm),
              length.out = 100))

# Use model to predict distance values
predvals <- predict(glm_diff, newdata = newdat, type = "link", se.fit = TRUE)

# Function to calculate inverse logit
inv.logit <- function(x) exp(x)/(1 + exp(x))

# Function to calculate critical significance value
ci.critical <- 
  function(siglevel) qnorm((100 - siglevel) / 100 / 2, lower.tail = FALSE)
ci_const = ci.critical(95)

# Add vals to newdat dataframe
newdat <- 
  newdat %>% 
  mutate(proximity = predvals$fit,
         se = predvals$se.fit,
         CI_lo = proximity - (ci_const * se),
         CI_hi = proximity + (ci_const * se)) 
# Back transform
newdat[,c("proximity", "CI_lo", "CI_hi")] <- 
  inv.logit(newdat[,c("proximity", "CI_lo", "CI_hi")])

# Plot model-predicted values
ggplot(newdat, 
       aes(x = AOHkm, y = proximity)) +
  # # Add confidence intervals
  # geom_ribbon(aes(ymin = CI_lo,
  #                 ymax = CI_hi,
  #                 fill = realm),
  #             alpha = 0.3) +
  # Add fitted line
  geom_line(aes(colour = Realms)) +
  # Facet by taxonomic class
  facet_grid(~ className) +
  theme_classic()

# Since class and realm aren't relevant, we can also just pick a reference level
# or average across everything, and plot one line only
newdat_avg <- 
  group_by(newdat, AOHkm) %>% 
  summarise(proximity = median(proximity),
            CI_lo = median(CI_lo),
            CI_hi = median(CI_hi))
ggplot(newdat_avg, 
       aes(x = AOHkm, y = proximity)) +
  # Add confidence intervals
  geom_ribbon(aes(ymin = CI_lo,
                  ymax = CI_hi),
              alpha = 0.1) +
  # Add fitted line
  geom_line() +
  theme_classic()

# Proximity to threshold increases as absolute AOH increases
```

### Correlation between className and Category
```{r class-category-corr}
ggplot(nes_df, aes(className, Category)) + 
  geom_count()

class_cat = table(nes_df$className, nes_df$Category) 

chisq <- chisq.test(class_cat)
chisq
```
p-value is very small (< 2.2e-16), therefore there is a strong correlation between Category and class


# Spatial Data
Save data frame for spatial analysis

```{r nes-top25-spatial}
nes_top25 <- 
  all_df %>% 
  mutate(top25 = ifelse(Category == 'Critically Endangered' & Difference >= 8.754866, 1, 
                        ifelse(Category == 'Endangered' & Difference >= 459.977142, 1, 
                               ifelse(Category == 'Vulnerable' & Difference >= 1278.32419, 1, 
                                      0)))) %>% 
  rename(scientificName = species) %>%
  mutate(
    # Match species name to how it appears in occurrence dataframe
    scientificName = gsub(" ", "_", scientificName)) 
```

```{r nes-top25-prop}
# Summarize NES by realm
realm_nes <-
  nes_top25 %>%
  filter(!(is.na(Realms))) %>% 
  # Count species in each realm
  group_by(Realms, className) %>% 
  summarise(n = n(),
            n_NES = length(which(NES)),
            n_nonNES = length(which(!(NES))),
            prop_NES = n_NES / n)
```

```{r save-data}
# Inspect the loaded objects - what might we need in the manuscript?
ls()
save(all_df,nes_df,realms_long_all,realms_long_nes,
     glm_diff,class_results,realm_results,aoh_results, 
     file = "HMEI-GDrive/data/sorted/msdata.Rdata")
```
