---
title: "No exit: the conservation reliance of species with little remaining habitat"
output:
    bookdown::word_document2:
        reference_docx: "../../../HMEI-GDrive/resources/word-templates/ms_general.docx"
        number_sections: false
    bookdown::pdf_document2:
        toc: no
        latex_engine: xelatex
        fig_caption: yes
        keep_tex: yes
link-citations: yes
csl: conservation-biology.csl
biblio-style: apalike
bibliography: [refs.bib, funky-refs.bib, pckgs.bib]
fontsize: 12pt
geometry: margin=2.5cm
mainfont: Calibri
papersize: a4
documentclass: article
header-includes:
- \usepackage{hyperref}
- \usepackage{fancyhdr}
- \pagestyle{fancy}
- \fancyhf{}
- \fancyfoot[CO,CE]{\thepage}
- \usepackage{setspace}
- \doublespacing
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = "../../../")
knitr::opts_chunk$set(echo = FALSE,message = FALSE,error = FALSE, warning = FALSE,
                      dpi = 300, fig.path = "figs/main/", fig.align = "center", 
                      fig.width = 16.6/2.54, fig.height = 15/2.54)
```

```{r packages}
library(readr)
library(kableExtra)
library(ggplot2)
library(stringr)
library(dplyr)
library(ggtext)
library(png)
library(grid)
library(RColorBrewer)
library(Hmisc)

# Define function to capitalise first letter
capitalise <- function(string) {
  substr(string, 1, 1) <- toupper(substr(string, 1, 1))
  string
}
```

```{r data}
load("HMEI-GDrive/data/sorted/msdata.Rdata")
```

```{r functions}
# Function to make significance levels pretty
siglevel <- function(pval){
  ifelse(pval < 0.001, "P < 0.001",
         ifelse(pval < 0.05, 
                paste("P = ", signif(pval, 2), sep = ""),
                "P > 0.05 ")
         )
}
```

<!-- _6000 words maximum: papers on applications of conservation science to specific goals for management, policy, or education that report findings important to decision making, planning, and implementation of conservation and provide opportunities for learning. They can include discussions of setbacks, failures, and unintended consequences._ -->
________________________________________________________________________________

# Abstract

Species that are "conservation reliant" require specific and continued management to maintain self-sustaining populations. The global standard for assessing species’ risk of extinction is the International Union for Conservation of Nature Red List of Threatened Species (Red List). Species can be listed as threatened based on habitat area and, conversely, improvements in threat status are constrained by habitat area. This reflects the real risks to species whose populations are confined to habitat of insufficient area. Recent advances in data availability have allowed us to create habitat maps and quantify Area of Habitat (AOH) for `r nrow(all_df)` threatened terrestrial species. Of these, we find that `r nrow(nes_df)` species have AOH < 2,000 km^2^ - the threshold for classifying species as threatened, under criterion B2. For these "No Exit Species", no amount of effort will permit their exit from threatened status unless accompanied by gains in habitat area. The average amount of additional habitat required is sobering; most No Exit Species are closer to the threshold for uplisting (increased extinction risk) than the threshold for downlisting (decreased extinction risk), and species at greatest risk of extinction tend to be furthest from the downlisting threshold. The majority of No Exit Species are concentrated in a few key countries, including Colombia, Mexico, Madagascar, and Indonesia. As we establish the framework for post-2020 conservation policy, it is vital that we allocate limited conservation resources to where they can have the biggest impact. Expansion of Protected Areas under "30 x 30" will help to safeguard species on the brink, while firm commitments to habitat restoration - through initiatives like the UN Decade on Ecosystem Restoration - will be pivotal for the recovery of the majority of threatened species. Without such concerted effort, No Exit Species will be conservation reliant for the long haul.

__295/300 words__

##  Article Impact Statement

_In ≤140 characters (including spaces and punctuation) reveal the paper’s primary importance to conservation. Emphasize practical or policy importance. See Instructions to Authors for more information._

Recovery of threatened species is fundamentally constrained by habitat area; protection and restoration remain priorities for conservation.

__139/140 characters__

## Reviewers

_Suggest at least 4 reviewers_

* David Williams, University of Leeds (UK), D.R.Williams@leeds.ac.uk
* Stuart Pimm, Duke University (USA), stuart.pimm@duke.edu
* Thomas Brooks, IUCN (Switzerland), t.brooks@iucn.org
* Ana Rodrigues, Centre National de la Recherche Scientifique (France), ana.rodrigues@cefe.cnrs.fr

# Introduction

Many threatened species are able to maintain self-sustaining populations in the wild only with ongoing species-specific management interventions. They are referred to as "conservation reliant species" [@scott_conservation-reliant_2010; @goble_conservation-reliant_2012; @scott_recovery_2005]. With continual management, such as control of invasive species or prescribed fires where early successional habitat is required [@wilcove_management_1998], conservation reliant species can be sustained in the wild and may even rebound [@bocettiUsingConservationManagement2012]. With cessation of management, however, such species quickly decline [@akcakaya_assessing_2019]. This conception of conservation reliance considers only the downward trajectory of threatened species, and how sustained interventions act as a barrier to further decline. Here, we argue that an additional form of conservation reliance is inherent in cases where an upward trajectory towards recovery of threatened species is fundamentally blocked by insufficient remaining habitat. Populations of such species must be diligently monitored and, at least occasionally, actively managed in order to reduce serious extinction risks associated with their limited ranges or small population sizes, including inbreeding, stochastic extinction events @lacyConsideringThreatsViability2000; @pflugerHabitatLossCauses2019; @brookContributionInbreedingExtinction2002], and habitat degradation (due to edge effects invasive species, or other factors) [@ewers_fragmentation_2013;@pfeifer_creation_2017]. We calculate how many of the world's threatened species fall into this category, and therefore require habitat restoration if they are to attain viable, self-sustaining populations. If decision makers wish to avert the extinction of these species, they must account for the cost of habitat restoration or, in its absence, active management, when developing conservation plans.

Area of available habitat is one of five criteria for categorizing species' risk of extinction by the International Union for Conservation of Nature (IUCN) Red List of Threatened Species. The Red List is the global standard for assessing the state of biodiversity, with categories of extinction risk ranging from Least Concern and Near Threatened to the gradated 'threatened' categories of: 'Vulnerable' 'Endangered,' and 'Critically Endangered' [@mace_quantification_2008; @iucn_iucn_2020]. As nations develop their post-2020 biodiversity goals, it is imperative that they identify those species whose recovery is intrinsically limited by available habitat, in order to set realistic expectations regarding conservation success. Moreover, to pragmatically allocate limited conservation resources, decision makers must be able to identify situations where upfront investment in costly habitat protection and restoration could save money in the long-term [@scott_conservation-reliant_2010], and then be able to identify where such restoration would have the biggest pay-offs for biodiversity [@kimball_cost-effective_2015]. 

The Red List sets quantitative criteria for each of its extinction risk categories [@iucn_guidelines_2019]. Under its 'geographic range' criterion (criterion B), species are classed as Vulnerable, Endangered, or Critically Endangered if their habitat area falls below specific thresholds. A simple but pertinent observation thus emerges: a threatened species cannot be moved ("downlisted") to a lower threat category on the Red List if the total area of its available habitat places it in a higher threat category [@iucn_iucn_2020], regardless of how well the species may be faring in terms of other criteria, such as population trends. This reflects the real threats faced by populations with insufficient habitat, and the associated barriers imposed on species recovery. 

Previously, estimates of habitat area were largely confined to 'extent of occurrence' (criterion B1), defined as the area of the minimum polygon that can be drawn to encompass all known, projected or inferred occurrences of a species [@iucn_guidelines_2019]. These range polygons are available for many species, but frequently suffer from 'errors of commission' when areas included within the range are not ecologically suitable for the species, often because the elevation or habitat type is not appropriate. Recent advances in data availability and resolution [@jung_global_2020], coupled with extensive data on species' habitat preferences [@iucn_iucn_2020; @birdlife_2020], now permit the refinement of species' extent of occurrence to the areas within a species' range that indeed have the appropriate habitat type and elevation [@brooks_measuring_2019], now termed the 'Area of Habitat' (AOH). With AOH maps we have unprecedented capacity to quantify both how much habitat each threatened species has, and the extent to which habitat area limits their recovery potential.

We explore how many species are habitat-limited and therefore conservation reliant, by comparing their AOH area to the thresholds used by the Red List (criterion B2). The AOH of every threatened species falls somewhere between the threshold for moving the species to a higher threat category ("uplisting"), and the threshold for moving it to a lower threat category ("downlisting"). These thresholds, in turn, can be used to estimate roughly the prospects for near and mid-term conservation success, given available habitat. Proximity of a species' AOH to the threshold for uplisting identifies that species as being on the brink, one whose remaining habitat must be protected to safeguard it from further decline. Conversely, the proximity of a species' habitat area to the threshold for downlisting identifies that species as potentially being an 'achievable win', whereby a relatively small increase in habitat could permit that species to be downlisted, and arguably be made less reliant on sustained conservation management. More ominously, because these thresholds identify species with habitat area meeting the threshold for categorization as threatened, they implicitly highlight the species that, based on the amount of habitat that currently remains, will never exit threatened status (i.e., move from Vulnerable to either Near Threatened or Least Concern) without concerted efforts to create new habitat. As such, these species are destined to remain conservation dependent until restoration efforts succeed in increasing their AOH and, presumably, their population sizes. We use the term 'No Exit Species' to describe species in this precarious state.

# Methods

## Area of Habitat

Following @brooks_measuring_2019, we created AOH maps for all threatened (Critically Endangered, Endangered, and Vulnerable) terrestrial species in those taxonomic groups that have range data and have been comprehensively assessed by the IUCN Red List to at least family level (5,008 species as of 14^th^ July 2020: all amphibians, birds, and mammals, plus select families in the classes Magnoliopsida and Reptilia; Table \@ref(tab:tab-1)). Range polygons were downloaded from the websites of the Red List [@iucn_iucn_2020] and BirdLife International [@birdlife_2020], and refined to AOH by retaining only grid cells that fall within the range polygons, and meet the species' habitat and elevation preferences as defined by the Red List [@iucn_iucn_2020]. All spatial analyses were conducted in Google Earth Engine [@gorelick_google_2017].

```{r tab-1}
# Function to count species in each assessed group
comprehensive_spp <- function(i){
  clade_level <- paste(i["clade"], "Name", sep = "")
  clade_name <- i["scientific"]
  # Get species in this clade
  spp <- all_df[all_df[,clade_level] == clade_name, "species"]
  # Return
  return(length(spp))
}

tab_1 <- 
  read.csv("HMEI-GDrive/data/assessed_final_2022-01-13.csv", 
           stringsAsFactors = FALSE) %>% 
  filter(class %in% unique(all_df$className)) %>% 
  dplyr::select(-included)
tab_1$count <- unlist(apply(tab_1, MARGIN = 1, comprehensive_spp))
class <- tab_1[tab_1$clade == "class",]
family <- tab_1[tab_1$clade == "family" &
                  tab_1$scientific %in% unique(all_df$familyName),]
order <- tab_1[tab_1$clade == "order" & tab_1$scientific == "crocodylia" ,]
tab_1 <-
  rbind(family, class, order) %>% 
    filter(count > 0) %>% 
  mutate(clade = factor(clade,
                        levels = c("class", "order", "family"))) %>% 
  arrange(clade, class) %>% 
  rename("English name" = common,
         "Scientific name" = scientific,
         "Taxonomic level" = clade,
         "Taxonomic class" = class,
         "Total threatened species" = count)
tab_1 <- as.data.frame(apply(tab_1, MARGIN = 2, capitalise))

# Format table
kable(tab_1, "simple", booktabs = TRUE, align = "l",
      caption = "Taxa comprehensively assessed on the IUCN Red List and included in our analyses.")
```

Following Red List assessment protocol [@iucn_guidelines_2019], we excluded habitats that are defined by the Red List as either 'marginal' in their suitability for a given species, or as 'passage', which reflects seasonality of use (e.g. stopover habitat for migratory birds). For migratory species we considered only the smaller of either the breeding or non-breeding range, in keeping with the Red List protocol [@iucn_guidelines_2019]. We did, however, include habitat types classified has having 'unknown' suitability or seasonality, because this situation likely affects many threatened species that are too rare or poorly known for more refined classifications to be possible.

To identify cells with suitable habitat, we used the 1 km-resolution IUCN habitat types map for the year 2015 [@jung_global_2020]. The Red List categorizes habitats at increasingly fine levels of resolution. Thus, level 1 habitat categories refer to the broad categories of Forest, Savanna, Shrubland, Grassland, Wetlands (inland), Rocky Areas, Deserts and Artificial habitats, each of which contains finer level 2 categories, e.g., 'Forest - Boreal' or 'Grassland - Tundra'. Where a species' habitat preference is given only to level 1 by IUCN, we conservatively assumed that all nested level 2 categories are suitable for that species. To identify cells with suitable elevation, we used the EarthEnv digital elevation model [@robinson_earthenv-dem90:_2014]. All maps were resampled to 2 km, following @iucn_guidelines_2019, and projected to a Mollweide projection to ensure accuracy of area calculations.

To quantify the extent to which AOH is a limiting factor in decreasing a species' risk of extinction, we compared AOH to the fixed thresholds under criterion B2, used by the Red List to assign species to the Red List categories. The Red List uses five criteria for this classification, and species are assigned to the highest risk category for which they qualify under _any_ of the criteria [@iucn_guidelines_2019]. For the three 'threatened' Red List categories of Vulnerable, Endangered and Critically Endangered, the AOH thresholds are, respectively, < 2000 km^2^; < 500 km^2^; and < 10 km^2^. We categorized species as likely to be conservation reliant if their AOH falls below any of the three thresholds, as all three imply a significant risk of extinction based solely on habitat amount; the higher the Red List category, the greater that risk.

Red List categorization under criterion B also requires that two of three additional subcriteria are met. These are: (1) the habitat is severely fragmented _or_ restricted to &le; 10 locations (Vulnerable), &le; 5 locations (Endangered), or 1 location (Critically Endangered); (2) there is a continuing decline in habitat area or quality, or the number of locations or number of mature individuals; and (3) there are extreme fluctuations in habitat area or quality, or in the number of locations or number of mature individuals. To accurately incorporate these subcriteria into our analyses would have required us to quantify habitat fragmentation and temporal trends in population size and habitat quality and quantity for all of the species [@santini_applying_2019]. Unfortunately, quantification of these characteristics cannot yet be reliably automated for all threatened species in our dataset. We therefore chose to follow a precautionary approach, based on habitat area alone. In other words, a species is assumed to be at or approaching the level of risk associated with each Red List category if its habitat area falls below the area thresholds defined under criterion B2. This is acceptable only because we are explicitly not attempting to validate or automate Red List classifications. Rather, we are exploring the extent to which habitat area limits species' capacities to recover. 

We classify species as 'No Exit Species' when their extant habitat area falls below the threshold for categorization as Vulnerable (< 2000 km^2^) under Red List criterion B2. We know _a priori_ that all No Exit Species are at high risk of extinction because we assessed only species that are classified as threatened (Critically Endangered, Endangered, or Vulnerable) on the Red List. No Exit Species are constrained to remain threatened indefinitely because of their limited habitat area. 

To give a hypothetical example, if a species is classified as Endangered and has 300 km^2^ of extant habitat, it cannot be downlisted to Vulnerable until its habitat area is increased above the threshold for Endangered (i.e. > 500 km^2^). In this case, at least 200 km^2^ (500 - 300 km^2^) of additional habitat would be required before the species could qualify for downlisting. To quantify the minimum additional habitat area that is required by each No Exit Species, we calculated the difference between each species' current AOH and the nearest threshold value that must be exceeded to permit the species to be downlisted to the next lowest threat category (henceforth: downlisting threshold). In other words, after identifying the lowest threshold below which each species' AOH falls, we calculated the difference between its current AOH value and that threshold. Where the AOH falls just beneath the threshold, relatively little additional habitat must be restored to permit the species to be downlisted (assuming, of course, that the restored area matched the habitat requirements of the species in question and that the species can colonize the new habitat). However, where AOH falls well below the downlisting threshold, relatively large gains in habitat area are needed to permit downlisting, and, indeed, the AOH of such species may instead be closer to the threshold beneath which, under criterion B2, it could qualify for uplisting to the next highest threat category (henceforth: uplisting threshold). 

## Statistical analyses

Red List thresholds for classification as under criterion B2 do not increase linearly as extinction risk decreases. For a species to move from Critically Endangered to Endangered, its AOH must increase above 10 km^2^, which at most represents an increase in habitat area of 10 km^2^. Moving from Endangered to Vulnerable can require up to 490 km^2^ of additional habitat (500 - 10 = 490 km^2^), and from Vulnerable to Near Threatened up to 1500 km^2^ of additional habitat (2000 - 500 km^2^). We refer to this quantity as the 'absolute difference from downlisting threshold', with a maximum value of either 10 km ^2^, 490 km ^2^ or 1500 km ^2^, depending on which downlisting threshold is the nearest. The minimum value, in contrast, is 0 km^2^, which represents species with an AOH precisely equal to the value of the nearest downlisting threshold (i.e. a species with AOH of exactly 10 km^2^, 500 km^2^, or 2000 km^2^). Thus, all No Exit Species fall into one of three groups depending on the downlisting threshold that they are nearest to; all have AOH < 2000 km^2^, and an absolute difference from downlisting threshold that ranges between 0-10 km^2^, 0-490 km ^2^ or 0-1500 km^2^. 

The absolute difference from downlisting threshold is informative because it represents the actual amount of additional habitat that is required for each threatened species to be able to be downlisted. However, it is also useful to consider a relative measure of difference from downlisting threshold, because this enables comparison across the different threshold groups. Thus, within each group, we identified species whose AOH falls in the upper quantile of absolute difference from the nearest downlisting threshold. These we termed 'Extreme No Exit Species', to describe those species whose AOH is furthest from the nearest downlisting threshold. We then calculated the 'relative difference from downlisting threshold' as the absolute difference divided by the maximum possible difference for each threshold (i.e. 10, 490 and 1500 km^2^). As an example, an Endangered species with 1250 km^2^ AOH has a larger absolute difference from the downlisting threshold for Vulnerable (2000 - 1250 = 750 km^2^) than does a Critically Endangered species with 5 km^2^ AOH, which requires only 5 km^2^ more habitat to be capable of being downlisted to Endangered (10 - 5 = 5 km^2^); however, both species have a relative difference from downlisting threshold of 0.5 (absolute differe / maximum difference: 750 / 1500 = 0.5 and 5 / 10 = 0.5). Thus, relative difference from downlisting threshold is a standardized measure that accounts for the fact that the threshold values do not increase linearly. To explore geographic patterns in the difference from downlisting threshold, we determined the countries and zoogeographic realms in which each species occurs, based on the overlap between species' range polygons and maps of global administrative areas [@gadm_2018] and zoogeographic realms [@holt_update_2013].

To compare the relative difference from downlisting threshold across all No Exit Species, we first plotted the frequency distribution of values using a Gaussian kernel density estimate [@wickham_ggplot2_2016]. A density curve is similar to a histogram, and used to display the distribution of continuous data based on the probability of observing data at each location along the x axis. The frequency distribution for relative difference from downlisting threshold is indicative of the extent to which species' AOH values are clustered around the downlisting threshold vs. the uplisting threshold. If the former, most species require relatively little additional habitat to permit them to be downlisted. If the latter, most species are as far from the downlisting threshold as possible. Indeed, in the latter case, species may instead be closer to the uplisting threshold, where relatively little loss of habitat would potentially cause species to be uplisted to a higher threat level. 

Lastly, we used a generalized linear model (GLM) with a binomial distribution, using the function `glm` in the `lme4` R  package [@bates_fitting_2015], to test whether relative difference from downlisting threshold varied by taxonomic class, zoogeographic realm and absolute AOH. To avoid double-counting species in this analysis, we limited our focus to species that occur in a single realm. Model inference was based on Likelihood Ratio Tests, dropping each variable in turn and comparing the reduced model to the full model [@zuur_mixed_2009]. All statistical analyses were conducted in R version 4.1.2 [@r_core_team_2021].

# Results

Of the `r nrow(all_df)` threatened species with AOH data, `r nrow(nes_df)` (`r signif(100*nrow(nes_df)/nrow(all_df),2)`%) are identifiable as No Exit Species: species that have a total AOH of < 2000 km^2^ (Fig. \@ref(fig:fig-1)). The remaining `r nrow(all_df) - nrow(nes_df)` threatened species with AOH > 2000 km^2^ are, therefore, not considered to be limited by available habitat quantity, and instead they are likely to be listed as threatened because of their small or declining population sizes.

Most No Exit Species are amphibians (`r length(which(nes_df$className == "amphibia"))`; `r signif(length(which(nes_df$className == "amphibia")) / nrow(nes_df) * 100,2)`%), followed by birds (`r length(which(nes_df$className == "aves"))`; `r signif(length(which(nes_df$className == "aves")) / nrow(nes_df) * 100,2)`%), mammals (`r length(which(nes_df$className == "mammalia"))`; `r signif(length(which(nes_df$className == "mammalia")) / nrow(nes_df) * 100,2)`%), plants in assessed families of Magnoliopsida (`r length(which(nes_df$className == "magnoliopsida"))`; `r signif(length(which(nes_df$className == "magnoliopsida")) / nrow(nes_df) * 100,2)`%), and, finally, species in assessed families of reptiles (`r length(which(nes_df$className == "reptilia"))`; `r signif(length(which(nes_df$className == "reptilia")) / nrow(nes_df) * 100,2)`%). Based on the Red List thresholds for listing under criterion B2, `r length(which(nes_df$ten_lowest))` species in our dataset have a total AOH that falls below the 10 km^2^ threshold for Critically Endangered, `r length(which(nes_df$five_lowest))` species have more than 10 km^2^ of habitat but less than 500 km^2^ (Endangered), and `r length(which(nes_df$two_lowest))` species have between 500-2000 km^2^ of habitat (Vulnerable). 

Extreme No Exit Species are those species whose existing AOH are furthest (top 25%) from the nearest downlisting threshold. We identified `r sum(nes_df$diff_top25)` such species, comprising `r sum(nes_df$diff_top25[nes_df$className=="amphibia"])` amphibians, `r sum(nes_df$diff_top25[nes_df$className=="aves"])` birds, `r sum(nes_df$diff_top25[nes_df$className=="mammalia"])` mammals, `r sum(nes_df$diff_top25[nes_df$className=="reptilia"])` reptiles, and `r sum(nes_df$diff_top25[nes_df$className=="magnoliopsida"])` plants.

```{r fig-1, fig.cap = "The total number of threatened species classified as 'No Exit Species' (NES), within each taxonomic class and AOH range. AOH ranges correspond to the threshold values used by the Red List to classify species under, criterion B2, as Critically Endangered (a), Endangered (b), or Vulnerable (c).",fig.width = 16/2.54, fig.height = 10/2.54}
nes <- nes_df %>% 
  mutate(Category = str_replace(Category, "Critically Endangered", 
                                paste("(a) AOH < 10 km","\U00B2",sep= ""))) %>% 
  mutate(Category = str_replace(Category, "Endangered", 
                                paste("(b) 10 < AOH < 500 km","\U00B2",sep= ""))) %>% 
  mutate(Category = str_replace(Category, "Vulnerable", 
                                paste("(c) 500 < AOH < 2000 km","\U00B2",sep= "")))

# relevel Categories
nes$Category <- 
  factor(nes$Category, 
         levels = c(paste("(a) AOH < 10 km","\U00B2",sep= ""), 
                    paste("(b) 10 < AOH < 500 km","\U00B2",sep= ""), 
                    paste("(c) 500 < AOH < 2000 km","\U00B2",sep= "")))
nes$className <- 
  factor(nes$className, 
         levels = unique(nes$className),
         labels = capitalize(unique(nes$className)))

# Code to add pics to labels
classes <- levels(nes$className)
labels <- paste(
  "<img src=", 
  "'",
  file.path("HMEI-GDrive/resources/phylopic", paste(tolower(classes), 'png', sep = ".")),
  "' ",
  "width='10' />", 
  classes,
  sep = ""
)
names(labels) <- classes

max_val <- 834
ggplot(nes, aes(x=className, fill=className)) + 
    geom_bar(alpha = 0.7) + 
    facet_wrap(~Category, labeller = label_wrap_gen(width = 25, multi_line = TRUE)) +
    theme_bw() + theme(panel.grid = element_blank()) +
    labs(fill="Class") +
    ylab("Count") + 
    scale_fill_brewer(palette="Set1") +
    scale_y_continuous(breaks = seq(0, max_val, by = 100), limits=c(0, max_val)) +
    theme(axis.text.x = element_markdown(color = "black",
                                         angle = 45, 
                                         hjust = 1, vjust = 1,
                                         margin = unit(c(-5, 0, 0, 0), "pt")),
          axis.title.x = element_blank(),
          legend.position = "none") +
    scale_x_discrete(labels = labels)
```

Considering the difference between species' AOH and the nearest downlisting threshold highlights those threatened species that are particularly near or far from being able to be downlisted to the next lowest threat category - in other words, the extent to which a species' potential for future downlisting is contingent on habitat restoration or creation. For No Exit Species, there are three possible ranges of AOH values, bounded by the different Red List thresholds of criterion B2: AOH < 10 km^2^; 10 < AOH &le; 500 km^2^; and 500 < AOH &le; 2000 km^2^. In each range and across different taxonomic classes, the AOH of most species falls closer to the threshold for uplisting (maximum difference) than the threshold for downlisting (minimum difference, i.e. 0 km^2^). This can be seen in the left skewed distribution of differences between AOH and the nearest downlisting threshold (Fig. \@ref(fig:fig-2)). 

We found no indication that species' difference from the nearest downlisting threshold differs between taxonomic classes (Deviance = `r signif(class_results[2,4],3)`, DF = `r class_results[2,3]`, `r siglevel(class_results[2,5])`) or biogeographic realms (Deviance = `r signif(realm_results[2,4],3)`, DF = `r realm_results[2,3]`, `r siglevel(realm_results[2,5])`). Species' with greater absolute AOH are relatively closer to the threshold for downlisting, and thus require relatively less additional habitat area to permit future downlisting (Deviance = `r signif(aoh_results[2,4],3)`, DF = `r aoh_results[2,3]`, `r siglevel(aoh_results[2,5])`), although this value may still be large in absolute terms.

```{r fig-2, fig.cap = "Distribution of the difference between species' absolute AOH and the nearest threshold for downlisting. Curves are shaded by taxonomic class and are derived from a Gaussian kernel density estimate. This alternative to a histogram is used to display the distribution of continuous data, based on the probability of observing data at each location along the x axis.",fig.width = 16/2.54, fig.height = 10/2.54}
ggplot(nes_df, aes(x=Difference, fill=className)) +
  geom_density(alpha = 0.3, colour = "transparent") +
  facet_wrap(~ Category, scales = "free", 
             labeller = label_wrap_gen(width = 15, multi_line = TRUE)) + 
  theme_bw() + theme(panel.grid = element_blank()) +
  labs(fill="Class") + 
  ylab("Density") +
  xlab(paste("Difference between AOH and downlisting threshold (km","\U00B2",")",sep= "")) + 
  theme(axis.text.x = element_text(angle = 45)) + 
  scale_fill_brewer(palette="Set1") 
```

By mapping total No Exit Species by country (Fig. 3), we find that a few countries harbor a disproportionate share of all species whose recovery on the Red List is limited by habitat area. Across all taxa, the three countries with the most No Exit Species are Colombia, Mexico, and Madagascar. With the addition of Indonesia, these countries also contain the highest number of Extreme No Exit Species - the species with AOH values furthest, in relative terms, from the nearest threshold for downlisting. Comparing among zoogeographic realms, the three realms with the most No Exit Species are Neotropical (489 species), Oriental (439 species), and Panamanian (431 species). These are also the three realms with the most Extreme No Exit Species.

Breaking down our geographic results by taxonomic class, No Exit amphibians are largely found in Mexico, Ecuador, Colombia and Madagascar, while No Exit birds are in tropical South America (Peru, Colombia, Brazil) and Indonesia, and mammals in Madagascar and Indonesia. Not all threatened magnoliopsid and reptile groups were able to be considered in these analyses, because they had not been comprehensively assessed on the Red List at the time of downloading these data (14^th^ July 2020). For assessed groups, China contains the majority of No Exit plants, while No Exit reptiles included in our analyses are largely found in Madagascar.

(ref:cap-3) The total number of threatened species within each country classified as (a) 'No Exit Species' (NES) and (b) 'Extreme No Exit Species' (ENES), which are those species in the upper quartile of differences between AOH and the nearest threshold for downlisting. Countries are shaded from low count (yellow) to high (red).

```{r fig-3, fig.cap = "(ref:cap-3)"}
map <- 
  readPNG("HMEI-GDrive/figs/map_country_combi.png") %>% 
  rasterGrob(interpolate = TRUE)
ggplot() +
  theme_void() +
  scale_x_continuous(expand = c(0,0)) +
  scale_y_continuous(expand = c(0,0)) +
  annotation_custom(map,
                    xmin = -Inf, xmax = Inf,
                    ymin = -Inf, ymax = Inf)
```

# Discussion

Recent advances in quantifying species' habitat areas, facilitated by the availability of global habitat maps and data on species' habitat preferences, allowed us to quantify the number of threatened, terrestrial species whose recovery from endangerment is fundamentally limited by the amount of remaining habitat for them. We identified `r nrow(nes_df)` species whose remaining habitat is below the threshold amount used by the Red List to categorize species as threatened (i.e. Vulnerable, Endangered, Critically Endangered) [@iucn_guidelines_2019]. We refer to these as "No Exit Species", because they cannot exit 'threatened' status (i.e., maintain viable, self-sustaining populations in the wild) without an increase in the amount of habitat available to them. Such species require active management to maintain existing populations, or habitat restoration to remove this restriction on their recovery. Intrinsically, therefore, No Exit Species should be included within the definition of "conservation reliant" [@scott_conservation-reliant_2010; @goble_conservation-reliant_2012; @scott_recovery_2005].

Alarmingly, the `r nrow(nes_df)` No Exit Species identified here account for over 60% of the `r nrow(all_df)` threatened species for which we had AOH data. This is comparable to the 60% of species on the US Endangered Species Act identified by @wilcove_management_1998 as conservation reliant, who based their definition of reliance on the need for invasive species control or prescribed burns. Similarly, @scott_conservation-reliant_2010 described 84% of species on the Endangered Species Act as conservation reliant, according to species' recovery plans. We emphasize that previous use of the term "conservation reliance" focused on species' downward trajectories and the role of continued conservation management in halting (but not necessarily reversing) further decline. Typically, ongoing management is needed to tackle ongoing threats [@scott_conservation-reliant_2010], but there are also situations in which species' populations remain too small or fragmented to be self-sustaining, even if the original driver of decline has abated. This could be because new threats have arisen, to which populations are now more susceptible, or because populations are held in stasis at a much reduced level and cannot rebound because they have insufficient habitat into which to expand. No Exit Species encapsulate this predicament: insufficient habitat prevents recovery, and thus confines their populations to a depressed state that typically requires conservation management in perpetuity just to maintain the status quo.

Most No Exit Species are amphibians, followed by birds and mammals and species in select magnoliopsid and reptile families. Many No Exit Species are recognized as Evolutionarily Distinct and Globally Endangered (EDGE) species [@zoologicalsocietyoflondonEDGE], with 40 amphibian, 27 bird, 30 mammal, and 4 reptile No Exit Species included in the top 100 EDGE species for their taxa. Examples include Archey's frog (_Leiopelma archeyi_), Barrio's frog (_Insuetophrynus acarpicus_), and the kagu (_Rhynochetos jubatus_). These species are both highly threatened and evolutionarily distinctive, such that their loss would have a disproportionate impact on phylogenetic diversity. Other No Exit Species are charismatic flagships and serve as useful umbrella species, including the Iberian Lynx (_Lynx pardinus_), the Tapanuli Orangutan (_Pongo tapanuliensis_), and the Indian Rhinoceros (_Rhinoceros unicornis_).

The prevalence of conservation reliance among threatened species is concerning because conservation is expensive, and funding is fickle [@deutz_financing_2020]. In recognising the sustained commitment necessary to maintain threatened species when habitat is limited, the often high upfront cost of habitat restoration could emerge as a compelling alternative [@kimball_cost-effective_2015]. Restoration through natural regeneration in particular [@brancalion_what_2019] could be cheaper and more enduring than relying on continued monetary and political commitment to conservation management _ad infinitum _. Unfortunately, financial accounting of conservation projects remains elusive [@white_what_2022], and this undermines efforts to assess costs and benefits to optimally direct limited resources.

For most No Exit Species, the absolute amount of their remaining habitat puts them at high risk of crossing the uplisting threshold into a higher threat category on the Red List (Fig. \@ref(fig:fig-2)). Indeed, most No Exit Species are in a precarious state: they are closer to the threshold for uplisting than for downlisting. Given the extent of conversion of natural ecosystems predicted for the remainder of this century [@jenkinsProspectsBiodiversity2003], the future for No Exit Species seems increasingly tenuous. The "30 x 30" goal to protect 30% of the world's land and coastal seas by 2030 [@30by30_2021], adopted by countries such as the UK and USA  [@bidenExecutiveOrder140082021; @50CountriesAnnounce], could prevent some No Exit Species from being uplisted to a higher threat category on the basis of habitat loss, but habitat protection per se cannot rescue No Exit Species from their threatened status.  For that, habitat restoration is required.

To permit downlisting to the next lowest risk category, the total amount of habitat needed across all of the No Exit Species is `r prettyNum(round(sum(nes_df$Difference[nes_df$Category == "Critically Endangered"]),0),big.mark = ",")` km^2^ for species with < 10 km^2^ habitat, `r prettyNum(round(sum(nes_df$Difference[nes_df$Category == "Endangered"]),0),big.mark = ",")` km^2^ for species with between 10 and 500 km^2^ habitat, and `r prettyNum(round(sum(nes_df$Difference[nes_df$Category == "Vulnerable"]),0),big.mark = ",")` km^2^ for species with between 500 and 2,000 km^2^ habitat. Undoubtedly there will be overlap between habitat requirements of different species, such that restoring habitat in one area could benefit multiple No Exit (and other rare) species. However, given the global distribution of No Exit Species (Fig. \@ref(fig:fig-3)), there remains a lot of ground to cover (restore, actually) to ensure that habitat area is no longer a barrier to species' ultimate recovery. Moreover, even this level of habitat restoration would only remove 'No Exit' status for species whose AOH is expanded to exceed the Vulnerable threshold of 2,000 km^2^. Species moving across the Critically Endangered (> 10 km^2^) or Endangered (> 500 km ^2^) threshold will still be conservation reliant, despite what would be relatively large gains in habitat area. 

To avoid making it even harder in the future to recover No Exist Species, land conservation efforts, such as 30x30 [@ConventionBiodiversityProposes], could target currently unprotected habitats for these species, thereby reducing the risk of future uplisting. Many priority sites for biodiversity protection have already been identified in the form of Key Biodiversity Areas (KBAs), according to the number of threatened species the site supports or because the site represents the majority of the geographic range or population of particular threatened species [@iucn_global_2016]. Given these criteria, it is likely that KBAs encompass a high number of No Exit Species. An arguably bigger challenge is to identify and restore currently unsuitable areas such that No Exit Species are able to be downlisted to lower threat categories, with the ultimate goal of restoring truly self-sustaining wild populations.

We note several limitations to our approach, most of which stem from the fact that we use AOH to identify No Exit Species. The first, as noted earlier, is that we considered only habitat area and not other habitat characteristics that could further restrict the true amount of usable habitat, such as habitat quality and fragmentation [@santini_applying_2019]. Similarly, while population sizes and trends are influenced by habitat availability, many species' populations are also depressed by additional threats that we did not consider, including invasive species and overexploitation, for example [@sodhiCausesConsequencesSpecies2009a]. Thus, the list of No Exit Species that we present here represents those for which habitat availability is likely to be a limiting factor in their recovery, but not necessarily the only limiting factor. Restoration of habitat can remove this habitat limitation barrier, and thus permit - but not guarantee - recovery of the many species that are threatened by habitat loss. Also, we identified No Exit Species by comparing species' AOH to the thresholds used by the Red List to classify species into categories of extinction risk, but some have argued these thresholds are not equally meaningful across taxa [@cardosoInvertebrates2012; @martinAreIUCNStandard2009]. Nonetheless, although imperfect, the Red List represents a heavily reviewed, widely accepted, and standardized protocol for measuring progress towards biodiversity targets [@keithScientificFoundationsIUCN2013].

Habitat loss and degradation will continue into the foreseeable future [@jenkinsProspectsBiodiversity2003], presenting ever greater challenges to the species that already have too little habitat. Here, in the year of the 15th Conference of Parties of the Convention on Biodiversity, and in the wake of growing concern about an ongoing extinction crisis [@ipbes_global_2019], we identify the species and regions most in need of habitat protection and restoration. No Exit Species are threatened species whose recovery is prevented by insufficient habitat area, making them reliant on ongoing monitoring and management. By assessing the amount of additional habitat that No Exit species would require to permit downlisting, we have determined that the majority of species are unfortunately closer to the threshold for uplisting than for downlisting, and therefore require substantial gains in habitat area to facilitate their ultimate recovery. As the major driver of species extinctions and with no obvious end in sight, insatiable habitat conversion must be met with an equal and opposite commitment to habitat protection and restoration. 

# References
