---
title: "No exit: the conservation reliance of species with too little habitat"
subtitle: 'Supplementary Information'
output:
    bookdown::pdf_document2:
        toc: no
        latex_engine: xelatex
        fig_caption: yes
        keep_tex: yes
        includes:
            in_header: header.tex 
fontsize: 12pt
geometry: margin=2.5cm
papersize: a4
documentclass: article
link-citations: yes
csl: elsevier-harvard.csl
biblio-style: apalike
bibliography: [refs.bib, funky-refs.bib, pckgs.bib]
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = "../../../")
knitr::opts_chunk$set(echo = FALSE,message = FALSE,error = FALSE, warning = FALSE,
                      dpi = 300, fig.path = "figs/som/", fig.align = "center", 
                      fig.width = 16.6/2.54, fig.height = 15/2.54)
```

```{r packages}
library(readr)
library(ggplot2)
library(stringr)
library(dplyr)
library(ggtext)
library(png)
library(grid)
library(RColorBrewer)
library(Hmisc)

# Define function to capitalise first letter
capitalise <- function(string) {
  substr(string, 1, 1) <- toupper(substr(string, 1, 1))
  string
}
```

```{r data}
load("HMEI-GDrive/data/sorted/msdata.Rdata")
```

_______________________________________________________________________________________

\pagebreak
\raggedright
\beginsupplement

\pagebreak

(ref:cap-1) The total number of threatened species within each country classified as 'No Exit Species'. Results are summarised for each taxonomic group (rows), and shaded from low count (yellow) to high (red).

```{r fig-s-1, fig.cap = "(ref:cap-1)", fig.width = 7/2.54, fig.height = 20/2.54}
map <- 
  readPNG("HMEI-GDrive/figs/nes_map_country.png") %>% 
  rasterGrob(interpolate = TRUE)
ggplot() +
  theme_void() +
  scale_x_continuous(expand = c(0,0)) +
  scale_y_continuous(expand = c(0,0)) +
  annotation_custom(map,
                    xmin = -Inf, xmax = Inf,
                    ymin = -Inf, ymax = Inf)
```

\pagebreak

(ref:cap-2) The total number of threatened species within each country classified as 'Extreme No Exit Species'. Results are summarised for each taxonomic group (rows), and shaded from low count (yellow) to high (red).

```{r fig-s-2, fig.cap = "(ref:cap-2)", fig.width = 7/2.54, fig.height = 20/2.54}
map <- 
  readPNG("HMEI-GDrive/figs/top25_map_country.png") %>% 
  rasterGrob(interpolate = TRUE)
ggplot() +
  theme_void() +
  scale_x_continuous(expand = c(0,0)) +
  scale_y_continuous(expand = c(0,0)) +
  annotation_custom(map,
                    xmin = -Inf, xmax = Inf,
                    ymin = -Inf, ymax = Inf)
```

\pagebreak

(ref:cap-3) The total number of threatened species within each zoogeographic realm classified as 'No Exit Species'. Results are summarised for each taxonomic group (rows), and shaded from low count (yellow) to high (red).

```{r fig-s-3, fig.cap = "(ref:cap-3)", fig.width = 7/2.54, fig.height = 20/2.54}
map <- 
  readPNG("HMEI-GDrive/figs/top25_map_realm.png") %>% 
  rasterGrob(interpolate = TRUE)
ggplot() +
  theme_void() +
  scale_x_continuous(expand = c(0,0)) +
  scale_y_continuous(expand = c(0,0)) +
  annotation_custom(map,
                    xmin = -Inf, xmax = Inf,
                    ymin = -Inf, ymax = Inf)
```

\pagebreak

(ref:cap-4) The total number of threatened species within each zoogeographic realm classified as 'Extreme No Exit Species'. Results are summarised for each taxonomic group (rows), and shaded from low count (yellow) to high (red).

```{r fig-s-4, fig.cap = "(ref:cap-4)", fig.width = 7/2.54, fig.height = 20/2.54}
map <- 
  readPNG("HMEI-GDrive/figs/top25_map_realm.png") %>% 
  rasterGrob(interpolate = TRUE)
ggplot() +
  theme_void() +
  scale_x_continuous(expand = c(0,0)) +
  scale_y_continuous(expand = c(0,0)) +
  annotation_custom(map,
                    xmin = -Inf, xmax = Inf,
                    ymin = -Inf, ymax = Inf)
```

\pagebreak

(ref:cap-5) The model-predicted relationship between species' proximity to threshold and absolute Area of Habitat (AOH), averaged across taxonomic classes and zoogeographic realms. Proximity to threshold is scaled from 0 to 1, with 0 representing the minimum difference between species' AOH and the downlisting threshold (i.e. no difference), and 1 representing the maximum difference between species' AOH and the downlisting threshold (the value of the maximum difference is dependent on which downlisting threshold AOH is being compared to). Grey shading represents 95% Confidence Intervals.

```{r fig-s-5, fig.cap = "(ref:cap-5)", fig.width = 16.6/2.54, fig.height = 18/2.54}
ggplot(newdat_avg, 
       aes(x = AOHkm, y = proximity)) +
    # Add confidence intervals
    geom_ribbon(aes(ymin = CI_lo,
                    ymax = CI_hi),
                alpha = 0.1) +
    # Add fitted line
    geom_line() +
    xlab(paste("AOH (km","\U00B2",")",sep= "")) + 
    ylab("Proximity to downlisting threshold") + 
    theme_bw() +
    theme(panel.grid = element_blank()) +
    scale_y_continuous(breaks = seq(0,1,0.1))
```


